---
title: "GO Enrichment Analysis - Sclerotinia sclerotiorum Pathogen DEGs"
author: "RNA-seq Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```
# 1. Setup and Library Loading

```{r libraries}
# Load required packages
library(clusterProfiler)

library(ggplot2)
library(dplyr)
library(tidyr)
library(enrichplot)
library(GOSemSim)

cat("✓ Libraries loaded successfully\n")
```
---

# 2. Load and Prepare Data

## 2.1 Set Working Directory
```{r set_wd}
# Set to your analysis directory
# TODO: set the working directory to the folder containing your STAR DEG CSVs
# setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_R/STAR/R/Enrichment_result")

cat("Working directory:", getwd(), "\n")
```
## 2.2 Load DEG File

```{r load_annotation}
cat("\n=== LOADING SCLEROTINIA GO ANNOTATION ===\n\n")

# Load JGI MycoCosm GO annotation
go_file <- "./Sclsc1_GeneCatalog_proteins_20110903_GO.tab"

if(!file.exists(go_file)) {
  stop("❌ GO annotation file not found: ", go_file)
}

go_annot <- read.delim(go_file, header = TRUE, stringsAsFactors = FALSE)

cat("✓ Loaded GO annotation file\n")
cat("  File:", go_file, "\n")
cat("  Rows:", nrow(go_annot), "\n")
cat("  Columns:", ncol(go_annot), "\n\n")

cat("Column names:\n")
print(colnames(go_annot))
cat("\n")

# Show structure
cat("Sample annotation data:\n")
print(head(go_annot, 10))

cat("\nAnnotation statistics:\n")
cat("  Unique protein IDs:", length(unique(go_annot$X.proteinId)), "\n")
cat("  Unique GO terms:", length(unique(go_annot$goAcc)), "\n")
cat("  Total gene-GO pairs:", nrow(go_annot), "\n")

cat("\n  Protein ID range in annotation:\n")
cat("    Min:", min(go_annot$X.proteinId), "\n")
cat("    Max:", max(go_annot$X.proteinId), "\n")
```
```{r create_go_mapping}
cat("\n=== CREATING GO-TO-GENE MAPPING ===\n\n")

# The GO annotation file has:
#   X.proteinId (numeric) → goAcc (GO:XXXXXXX)
# Our fixed DEG IDs are now numeric protein IDs

# Create term2gene dataframe (GO term → gene mapping)
term2gene <- data.frame(
  GO = go_annot$goAcc,
  GeneID = as.character(go_annot$X.proteinId),  # Convert to character for clusterProfiler
  stringsAsFactors = FALSE
)

cat("Created term2gene mapping:\n")
cat("  Total GO-gene pairs:", nrow(term2gene), "\n")
cat("  Unique GO terms:", length(unique(term2gene$GO)), "\n")
cat("  Unique genes:", length(unique(term2gene$GeneID)), "\n\n")

cat("Sample mapping:\n")
print(head(term2gene, 10))

# Also create term2name (GO term → GO name)
term2name <- data.frame(
  GO = go_annot$goAcc,
  Name = go_annot$goName,
  stringsAsFactors = FALSE
) %>%
  distinct(GO, .keep_all = TRUE)

cat("\nCreated term2name mapping:\n")
cat("  Unique GO terms with names:", nrow(term2name), "\n\n")

cat("Sample GO term names:\n")
print(head(term2name, 10))
```

---

# Per-Timepoint GO Enrichment Analysis (24h, 48h, 96h)

The analysis below is run independently for each of the three timepoints.
Each iteration loads its own DEG CSV, runs the same enrichment pipeline,
produces the same plots, and saves results to per-timepoint output folders.
A cross-timepoint summary is generated at the end.

```{r enrichment_loop, results='asis', fig.width=12, fig.height=8}
# === Per-timepoint loop ===
timepoints <- c("24h", "48h", "96h")
master_results <- list()

# Stash the static term2gene/term2name from earlier setup chunks so the loop
# can rely on them by stable name in every iteration (they don't change).
.term2gene_static <- term2gene
.term2name_static <- term2name

for (tp in timepoints) {
  tp_suffix <- paste0("_", tp)

  cat("\n\n---\n\n")
  cat(sprintf("# === TIMEPOINT: %s ===\n\n", tp))

  # Re-bind annotation objects in case they were shadowed
  term2gene <- .term2gene_static
  term2name <- .term2name_static

  # ╔══════════════════════════════════════════╗
  # ║ Section: load_degs
  # ╚══════════════════════════════════════════╝
  cat("\n=== LOADING PATHOGEN DEG FILE ===\n\n")

  # Load the DEG file WITH row.names = 1 (gene IDs in rownames)
  deg_file <- sprintf("./Ha_pathogen_aggressive_vs_low_DEGs_%s.csv", tp)

  # Check if file exists
  if(!file.exists(deg_file)) {
    stop("❌ DEG file not found: ", deg_file)
  }

  cat("✓ File found:", basename(deg_file), "\n\n")

  # Load with gene IDs as rownames
  pathogen_degs <- read.csv(deg_file, stringsAsFactors = FALSE, row.names = 1)

  cat("✓ Loaded DEG file\n")
  cat("  Total DEGs:", nrow(pathogen_degs), "\n")
  cat("  Columns:", ncol(pathogen_degs), "\n\n")

  # Show structure
  cat("File structure:\n")
  str(pathogen_degs)

  cat("\nFirst few rows:\n")
  print(head(pathogen_degs))


  # ╔══════════════════════════════════════════╗
  # ║ Section: explore_degs
  # ╚══════════════════════════════════════════╝
  cat("\n=== DEG FILE OVERVIEW ===\n\n")

  # Show first few rows (gene IDs are in rownames)
  cat("First few genes (original JGI format):\n")
  print(head(rownames(pathogen_degs)))

  cat("\n\nFirst few data rows:\n")
  print(head(pathogen_degs))

  # Check for required columns
  cat("\nChecking for log2FoldChange column...\n")
  lfc_cols <- grep("log.*fold|lfc", colnames(pathogen_degs), ignore.case = TRUE, value = TRUE)
  if(length(lfc_cols) > 0) {
    cat("✓ Found log2FoldChange column:", lfc_cols[1], "\n")
  } else {
    cat("⚠️ No log2FoldChange column found\n")
  }


  # ╔══════════════════════════════════════════╗
  # ║ Section: fix_gene_ids
  # ╚══════════════════════════════════════════╝
  cat("\n=== EXTRACTING PROTEIN IDs ===\n\n")

  cat("Gene ID format: SS1G_#####T0\n")
  cat("                └─────┬─────\n")
  cat("                      └─ Numeric protein ID (extract this)\n\n")

  # Gene names are already in rownames - copy them
  pathogen_degs$gene_name <- rownames(pathogen_degs)

  # Extract numeric protein ID from gene name
  # SS1G_12396T0 → 12396
  pathogen_degs$protein_id <- as.numeric(gsub("SS1G_(\\d+)T\\d+", "\\1", rownames(pathogen_degs)))

  # Keep original ID for reference
  pathogen_degs$original_gene_id <- rownames(pathogen_degs)

  # Reorder columns to put IDs first
  pathogen_degs <- pathogen_degs %>%
    select(gene_name, protein_id, original_gene_id, everything())

  cat("Extraction examples:\n")
  extraction_df <- data.frame(
    Gene_Name = head(pathogen_degs$gene_name, 10),
    Protein_ID = head(pathogen_degs$protein_id, 10),
    stringsAsFactors = FALSE
  )
  print(extraction_df)

  cat("\n\nExtraction statistics:\n")
  cat("  Total genes:", nrow(pathogen_degs), "\n")
  cat("  Successfully extracted protein IDs:", sum(!is.na(pathogen_degs$protein_id)), "\n")
  cat("  Failed extractions:", sum(is.na(pathogen_degs$protein_id)), "\n")
  cat("  Unique protein IDs:", length(unique(pathogen_degs$protein_id)), "\n")

  if(sum(is.na(pathogen_degs$protein_id)) > 0) {
    cat("\n⚠️ Warning: Some IDs failed to extract!\n")
    cat("Failed IDs:\n")
    print(pathogen_degs$gene_name[is.na(pathogen_degs$protein_id)])
  } else {
    cat("\n✓ All protein IDs extracted successfully!\n")
  }

  cat("\nUpdated dataframe structure:\n")
  str(pathogen_degs)

  cat("\nSample data:\n")
  print(head(pathogen_degs[, 1:6], 10))


  # ╔══════════════════════════════════════════╗
  # ║ Section: separate_regulation
  # ╚══════════════════════════════════════════╝
  cat("\n=== SEPARATING UP AND DOWN-REGULATED GENES ===\n\n")

  # Find log2FoldChange column
  lfc_col <- grep("log.*fold|lfc", colnames(pathogen_degs), ignore.case = TRUE, value = TRUE)[1]

  if(!is.na(lfc_col)) {
    # Separate by regulation direction
    upregulated <- pathogen_degs[pathogen_degs[[lfc_col]] > 0, ]
    downregulated <- pathogen_degs[pathogen_degs[[lfc_col]] < 0, ]
    
    cat("Regulation breakdown:\n")
    cat("  Up-regulated:", nrow(upregulated), "genes\n")
    cat("  Down-regulated:", nrow(downregulated), "genes\n")
    cat("  Total:", nrow(pathogen_degs), "genes\n")
  } else {
    cat("⚠️ Cannot separate by regulation - treating all as one group\n")
    upregulated <- pathogen_degs
    downregulated <- data.frame()
  }




  # ╔══════════════════════════════════════════╗
  # ║ Section: check_coverage
  # ╚══════════════════════════════════════════╝
  cat("\n=== CHECKING ANNOTATION COVERAGE ===\n\n")

  # Get unique DEG protein IDs and annotation IDs (both as character for comparison)
  deg_protein_ids <- unique(as.character(pathogen_degs$protein_id[!is.na(pathogen_degs$protein_id)]))
  annot_protein_ids <- unique(term2gene$GeneID)

  # Find overlap
  overlap_ids <- intersect(deg_protein_ids, annot_protein_ids)

  cat("Coverage Statistics:\n")
  cat("  Unique DEG protein IDs:", length(deg_protein_ids), "\n")
  cat("  Unique annotated protein IDs:", length(annot_protein_ids), "\n")
  cat("  Overlap (genes with annotation):", length(overlap_ids), "\n")
  cat("  Coverage:", round(100 * length(overlap_ids) / length(deg_protein_ids), 1), "%\n\n")

  if(length(overlap_ids) == 0) {
    cat("❌ ERROR: No overlap between DEG IDs and annotation IDs!\n")
    cat("\nSample DEG protein IDs:\n")
    print(head(deg_protein_ids, 10))
    cat("\nSample annotation protein IDs:\n")
    print(head(annot_protein_ids, 10))
    stop("Gene ID mismatch - check ID extraction")
  }

  cat("✓ Good overlap found!\n\n")

  cat("Sample matching protein IDs:\n")
  print(head(overlap_ids, 10))

  # Show which DEG genes they correspond to
  cat("\nCorresponding gene names:\n")
  matching_genes <- pathogen_degs %>%
    filter(as.character(protein_id) %in% head(overlap_ids, 10)) %>%
    select(protein_id, gene_name, original_gene_id)
  print(matching_genes)

  # Check coverage by ontology
  cat("\n\nCoverage by GO ontology:\n")
  go_types <- go_annot %>%
    filter(as.character(X.proteinId) %in% overlap_ids) %>%
    group_by(gotermType) %>%
    summarise(
      genes = n_distinct(X.proteinId),
      terms = n_distinct(goAcc)
    )
  print(go_types)

  cat("\n** Note: 20-40% coverage is NORMAL and EXPECTED for fungal pathogens **\n")
  cat("** Many genes are hypothetical proteins or lack experimental characterization **\n")
  cat("** This coverage is sufficient for meaningful enrichment analysis **\n")


  # ╔══════════════════════════════════════════╗
  # ║ Section: prepare_gene_lists
  # ╚══════════════════════════════════════════╝
  cat("\n=== PREPARING GENE LISTS FOR ENRICHMENT ===\n\n")

  # Convert protein IDs to character for clusterProfiler
  # Remove NAs first
  all_genes_list <- as.character(pathogen_degs$protein_id[!is.na(pathogen_degs$protein_id)])
  up_genes_list <- if(nrow(upregulated) > 0) {
    as.character(upregulated$protein_id[!is.na(upregulated$protein_id)])
  } else {
    character(0)
  }
  down_genes_list <- if(nrow(downregulated) > 0) {
    as.character(downregulated$protein_id[!is.na(downregulated$protein_id)])
  } else {
    character(0)
  }

  cat("Gene lists prepared:\n")
  cat("  All DEGs:", length(all_genes_list), "genes\n")
  cat("  Up-regulated:", length(up_genes_list), "genes\n")
  cat("  Down-regulated:", length(down_genes_list), "genes\n\n")

  cat("Sample gene IDs (protein IDs):\n")
  cat("  All:", paste(head(all_genes_list, 5), collapse = ", "), "\n")
  if(length(up_genes_list) > 0) {
    cat("  Up:", paste(head(up_genes_list, 5), collapse = ", "), "\n")
  }
  if(length(down_genes_list) > 0) {
    cat("  Down:", paste(head(down_genes_list, 5), collapse = ", "), "\n")
  }


  # ╔══════════════════════════════════════════╗
  # ║ Section: enrich_all
  # ╚══════════════════════════════════════════╝
  cat("\n=== GO ENRICHMENT: ALL DEGs ===\n\n")

  # Run enrichment using custom term2gene
  ego_all <- enricher(
    gene = all_genes_list,
    TERM2GENE = term2gene,
    TERM2NAME = term2name,
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    minGSSize = 10,
    maxGSSize = 500
  )

  if(!is.null(ego_all) && nrow(ego_all@result) > 0) {
    cat("✓ Found", nrow(ego_all@result), "enriched GO terms\n\n")
    cat("Top 10 enriched terms:\n")
    print(head(ego_all@result[, c("ID", "Description", "GeneRatio", "pvalue", "qvalue")], 10))
  } else {
    cat("⚠️ No significantly enriched terms found for all DEGs\n")
    cat("   (This can happen with small gene sets or low coverage)\n")
  }


  # ╔══════════════════════════════════════════╗
  # ║ Section: enrich_up
  # ╚══════════════════════════════════════════╝
  cat("\n=== GO ENRICHMENT: UP-REGULATED DEGs ===\n\n")

  if(length(up_genes_list) >= 10) {
    ego_up <- enricher(
      gene = up_genes_list,
      TERM2GENE = term2gene,
      TERM2NAME = term2name,
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2,
      minGSSize = 10,
      maxGSSize = 500
    )
    
    if(!is.null(ego_up) && nrow(ego_up@result) > 0) {
      cat("✓ Found", nrow(ego_up@result), "enriched GO terms\n\n")
      cat("Top 10 enriched terms:\n")
      print(head(ego_up@result[, c("ID", "Description", "GeneRatio", "pvalue", "qvalue")], 10))
    } else {
      cat("⚠️ No significantly enriched terms found for up-regulated genes\n")
      ego_up <- NULL
    }
  } else {
    cat("⚠️ Not enough up-regulated genes (need at least 10)\n")
    ego_up <- NULL
  }


  # ╔══════════════════════════════════════════╗
  # ║ Section: enrich_down
  # ╚══════════════════════════════════════════╝
  cat("\n=== GO ENRICHMENT: DOWN-REGULATED DEGs ===\n\n")

  if(length(down_genes_list) >= 10) {
    ego_down <- enricher(
      gene = down_genes_list,
      TERM2GENE = term2gene,
      TERM2NAME = term2name,
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2,
      minGSSize = 10,
      maxGSSize = 500
    )
    
    if(!is.null(ego_down) && nrow(ego_down@result) > 0) {
      cat("✓ Found", nrow(ego_down@result), "enriched GO terms\n\n")
      cat("Top 10 enriched terms:\n")
      print(head(ego_down@result[, c("ID", "Description", "GeneRatio", "pvalue", "qvalue")], 10))
    } else {
      cat("⚠️ No significantly enriched terms found for down-regulated genes\n")
      ego_down <- NULL
    }
  } else {
    cat("⚠️ Not enough down-regulated genes (need at least 10)\n")
    ego_down <- NULL
  }


  # ╔══════════════════════════════════════════╗
  # ║ Section: dotplot_all
  # ╚══════════════════════════════════════════╝
  # METHOD 2: Custom ggplot2 dotplot (RECOMMENDED - Always works!)
  # ----------------------------------------------------------------------------

  create_custom_dotplot <- function(ego, title = "GO Enrichment", top_n = 20) {
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No results to plot")
      return(NULL)
    }
    
    # Get data
    df <- ego@result
    
    # Keep top N by p-value
    df <- df[order(df$pvalue), ][1:min(top_n, nrow(df)), ]
    
    # Parse GeneRatio
    df$GeneRatio_numeric <- sapply(strsplit(df$GeneRatio, "/"), 
                                    function(x) as.numeric(x[1])/as.numeric(x[2]))
    
    # Get gene count
    df$Count <- sapply(strsplit(df$GeneRatio, "/"), 
                       function(x) as.numeric(x[1]))
    
    # Order by GeneRatio for plotting
    df$Description <- factor(df$Description, levels = df$Description[order(df$GeneRatio_numeric)])
    
    # Create plot
    p <- ggplot(df, aes(x = GeneRatio_numeric, y = Description)) +
      geom_point(aes(size = Count, color = pvalue)) +
      scale_color_gradient(low = "red", high = "blue", 
                          name = "p-value",
                          guide = guide_colorbar(reverse = TRUE)) +
      scale_size_continuous(name = "Gene Count", range = c(3, 8)) +
      labs(title = title,
           x = "Gene Ratio",
           y = NULL) +
      theme_bw(base_size = 12) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        legend.position = "right"
      )
    
    return(p)
  }

  # Create custom dotplot
  p_all <- create_custom_dotplot(ego_all, 
                                    title = "GO Enrichment - All Pathogen DEGs", 
                                    top_n = 20)
  print(p_all)

  p_up <- create_custom_dotplot(ego_up, 
                                    title = "GO Enrichment - Upregulated Pathogen DEGs", 
                                    top_n = 20)
  print(p_up)

  p_down <- create_custom_dotplot(ego_down, 
                                    title = "GO Enrichment - Downregulated Pathogen DEGs", 
                                    top_n = 20)
  print(p_down)





  # ╔══════════════════════════════════════════╗
  # ║ Section: barplot_all_2
  # ╚══════════════════════════════════════════╝

  # ALTERNATIVE: Simple barplot (always reliable!)
  # ----------------------------------------------------------------------------

  create_simple_barplot <- function(ego, title = "GO Enrichment", top_n = 20) {
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No results to plot")
      return(NULL)
    }
    
    df <- ego@result
    df <- df[order(df$pvalue), ][1:min(top_n, nrow(df)), ]
    
    # Parse count
    df$Count <- sapply(strsplit(df$GeneRatio, "/"), 
                       function(x) as.numeric(x[1]))
    
    # Wrap descriptions
    df$Description_wrapped <- stringr::str_wrap(df$Description, width = 50)
    df$Description_wrapped <- factor(df$Description_wrapped,
                                     levels = rev(df$Description_wrapped))
    
    # Barplot
    p <- ggplot(df, aes(x = Count, y = Description_wrapped, fill = pvalue)) +
      geom_col() +
      scale_fill_gradient(low = "red", high = "blue",
                         name = "p-value",
                         guide = guide_colorbar(reverse = TRUE)) +
      labs(title = title,
           x = "Gene Count",
           y = NULL) +
      theme_minimal(base_size = 11) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.y = element_text(size = 9),
        legend.position = "right"
      )
    
    return(p)
  }

  # Create custom barplot
  p_all <- create_simple_barplot(ego_all,
                                title = "GO Enrichment - All Pathogen DEGs",
                                top_n = 20)
  print(p_all)

  p_up <- create_simple_barplot(ego_up,
                                title = "GO Enrichment - All Pathogen DEGs",
                                top_n = 20)
  print(p_up)

  p_down <- create_simple_barplot(ego_down,
                                title = "GO Enrichment - All Pathogen DEGs",
                                top_n = 20)
  print(p_down)




  # ╔══════════════════════════════════════════╗
  # ║ Section: networks
  # ╚══════════════════════════════════════════╝
  # All DEGs - Gene-Concept Network
  if(!is.null(ego_all) && nrow(ego_all@result[ego_all@result$p.adjust < 0.05, ]) >= 5) {
    cat("Creating gene-concept network...\n")
    p7 <- cnetplot(ego_all, 
                   categorySize = "pvalue",
                   showCategory = 5,
                   colorEdge = TRUE,
                   node_label = "category")
    print(p7)
  }

  # Enrichment Map
  if(!is.null(ego_all) && nrow(ego_all@result[ego_all@result$p.adjust < 0.05, ]) >= 5) {
    cat("Creating enrichment map...\n")
    ego_all_pair <- pairwise_termsim(ego_all)
    p8 <- emapplot(ego_all_pair, 
                   showCategory = 20,
                   cex_label_category = 0.7)
    print(p8)
  }


  # ╔══════════════════════════════════════════╗
  # ║ Section: cnetplot
  # ╚══════════════════════════════════════════╝
  # ============================================================================
  # FIX FOR CNETPLOT ERROR WITH CUSTOM ENRICHER() RESULTS
  # ============================================================================
  # 
  # Error: graph_from_data_frame(): `d` should contain at least two columns
  # 
  # This happens because cnetplot() expects specific data structure that may not
  # be present in custom enricher() results.
  #
  # Solutions:
  # 1. Fix the enricher object for cnetplot compatibility
  # 2. Create custom network plot with ggplot2/igraph
  # 3. Use alternative visualizations (heatmap, upset plot)
  # ============================================================================

  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(stringr)

  # ============================================================================
  # METHOD 1: Fix enricher result for cnetplot
  # ============================================================================

  fix_enricher_for_cnetplot <- function(ego) {
    if(is.null(ego) || nrow(ego@result) == 0) {
      return(NULL)
    }
    
    # Ensure geneID column contains proper gene IDs (separated by "/")
    # This should already be there from enricher(), but we'll verify
    if(!"geneID" %in% colnames(ego@result)) {
      warning("No geneID column found - cannot create network plot")
      return(NULL)
    }
    
    # Make sure gene IDs are properly formatted
    # They should be in format: "gene1/gene2/gene3"
    ego@result$geneID <- as.character(ego@result$geneID)
    
    # Add Count column if missing
    if(!"Count" %in% colnames(ego@result)) {
      ego@result$Count <- sapply(strsplit(ego@result$GeneRatio, "/"), 
                                  function(x) as.numeric(x[1]))
    }
    
    return(ego)
  }

  # Try to fix and plot
  ego_all_fixed <- fix_enricher_for_cnetplot(ego_all)

  if(!is.null(ego_all_fixed)) {
    tryCatch({
      p3 <- cnetplot(ego_all_fixed, showCategory=5, colorEdge=TRUE, circular=FALSE) +
        ggtitle("Gene-Concept Network - Top 5 Terms") +
        theme(plot.title = element_text(hjust = 0.5, face = "bold"))
      print(p3)
    }, error = function(e) {
      cat("cnetplot still failed. Using alternative visualization below.\n")
      cat("Error:", e$message, "\n\n")
    })
  }


  # ============================================================================
  # METHOD 2: Custom Gene-Term Heatmap (RECOMMENDED ALTERNATIVE)
  # ============================================================================

  create_gene_term_heatmap <- function(ego, term2gene, top_terms = 10, max_genes = 50) {
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No results to plot")
      return(NULL)
    }
    
    # Get top terms
    top_results <- ego@result[order(ego@result$pvalue), ][1:min(top_terms, nrow(ego@result)), ]
    
    # Extract gene-term relationships
    gene_term_list <- list()
    
    for(i in 1:nrow(top_results)) {
      term_id <- top_results$ID[i]
      term_name <- top_results$Description[i]
      genes <- unlist(strsplit(top_results$geneID[i], "/"))
      
      # Limit genes if too many
      if(length(genes) > max_genes) {
        genes <- genes[1:max_genes]
      }
      
      gene_term_list[[i]] <- data.frame(
        Gene = genes,
        Term = str_wrap(term_name, width = 30),
        GO_ID = term_id,
        pvalue = top_results$pvalue[i],
        stringsAsFactors = FALSE
      )
    }
    
    # Combine all
    gene_term_df <- do.call(rbind, gene_term_list)
    
    # Create presence/absence matrix
    gene_term_matrix <- gene_term_df %>%
      mutate(present = 1) %>%
      select(Gene, Term, present) %>%
      pivot_wider(names_from = Term, values_from = present, values_fill = 0)
    
    # Convert to long format for ggplot
    gene_term_long <- gene_term_matrix %>%
      pivot_longer(-Gene, names_to = "Term", values_to = "Present") %>%
      filter(Present == 1)
    
    # Add term order by p-value
    term_order <- top_results %>%
      mutate(Term = str_wrap(Description, width = 30)) %>%
      arrange(pvalue) %>%
      pull(Term)
    
    gene_term_long$Term <- factor(gene_term_long$Term, levels = rev(term_order))
    
    # Count how many terms each gene appears in
    gene_counts <- gene_term_long %>%
      group_by(Gene) %>%
      summarise(n_terms = n()) %>%
      arrange(desc(n_terms))
    
    # Order genes by frequency
    gene_term_long$Gene <- factor(gene_term_long$Gene, 
                                   levels = gene_counts$Gene)
    
    # Create heatmap
    p <- ggplot(gene_term_long, aes(x = Gene, y = Term)) +
      geom_tile(fill = "#2166AC", color = "white", linewidth = 0.5) +
      labs(
        title = "Gene-Term Relationship Matrix",
        subtitle = paste("Top", top_terms, "GO terms"),
        x = "Genes",
        y = "GO Terms"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
        axis.text.y = element_text(size = 9),
        panel.grid = element_blank(),
        axis.title = element_text(face = "bold")
      )
    
    return(p)
  }

  # Create heatmap
  cat("\n=== Creating Gene-Term Heatmap (Alternative to cnetplot) ===\n\n")
  p_heatmap <- create_gene_term_heatmap(ego_all, term2gene, top_terms = 10, max_genes = 40)
  print(p_heatmap)


  # ============================================================================
  # METHOD 3: UpSet Plot - Show Gene Overlap Between Terms
  # ============================================================================

  create_upset_style_plot <- function(ego, top_terms = 8) {
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No results to plot")
      return(NULL)
    }
    
    # Get top terms
    top_results <- ego@result[order(ego@result$pvalue), ][1:min(top_terms, nrow(ego@result)), ]
    
    # Extract gene lists for each term
    term_genes <- list()
    for(i in 1:nrow(top_results)) {
      term_name <- str_wrap(top_results$Description[i], width = 40)
      genes <- unlist(strsplit(top_results$geneID[i], "/"))
      term_genes[[term_name]] <- genes
    }
    
    # Calculate intersections
    all_genes <- unique(unlist(term_genes))
    
    # Create binary matrix
    gene_matrix <- matrix(0, nrow = length(all_genes), ncol = length(term_genes))
    rownames(gene_matrix) <- all_genes
    colnames(gene_matrix) <- names(term_genes)
    
    for(i in 1:length(term_genes)) {
      gene_matrix[term_genes[[i]], i] <- 1
    }
    
    # Count genes per term and overlaps
    term_sizes <- colSums(gene_matrix)
    
    # Create a simple bar plot showing term sizes
    term_size_df <- data.frame(
      Term = factor(names(term_sizes), levels = names(term_sizes)),
      Size = as.numeric(term_sizes),
      stringsAsFactors = FALSE
    )
    
    p <- ggplot(term_size_df, aes(x = reorder(Term, Size), y = Size)) +
      geom_col(fill = "#2166AC", alpha = 0.8) +
      geom_text(aes(label = Size), hjust = -0.2, size = 3.5) +
      coord_flip() +
      labs(
        title = "Number of Genes per GO Term",
        subtitle = paste("Top", top_terms, "enriched terms"),
        x = "GO Term",
        y = "Number of Genes"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        axis.text.y = element_text(size = 9),
        axis.title = element_text(face = "bold"),
        panel.grid.minor = element_blank()
      ) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
    
    return(p)
  }

  # Create upset-style plot
  cat("\n=== Creating Gene Count Plot ===\n\n")
  p_upset <- create_upset_style_plot(ego_all, top_terms = 10)
  print(p_upset)


  # ============================================================================
  # METHOD 4: Network Plot using igraph (Manual Construction)
  # ============================================================================

  create_manual_network_plot <- function(ego, top_terms = 5, max_genes_per_term = 10) {
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No results to plot")
      return(NULL)
    }
    
    library(igraph)
    
    # Get top terms
    top_results <- ego@result[order(ego@result$pvalue), ][1:min(top_terms, nrow(ego@result)), ]
    
    # Create edge list (Term -> Gene connections)
    edges <- data.frame()
    
    for(i in 1:nrow(top_results)) {
      term_name <- paste0("T", i, ": ", str_wrap(top_results$Description[i], width = 25))
      genes <- unlist(strsplit(top_results$geneID[i], "/"))
      
      # Limit genes
      if(length(genes) > max_genes_per_term) {
        genes <- genes[1:max_genes_per_term]
      }
      
      # Create edges
      term_edges <- data.frame(
        from = term_name,
        to = genes,
        stringsAsFactors = FALSE
      )
      
      edges <- rbind(edges, term_edges)
    }
    
    # Create graph
    g <- graph_from_data_frame(edges, directed = FALSE)
    
    # Set node attributes
    V(g)$type <- ifelse(V(g)$name %in% edges$from, "Term", "Gene")
    V(g)$color <- ifelse(V(g)$type == "Term", "#E31A1C", "#1F78B4")
    V(g)$size <- ifelse(V(g)$type == "Term", 12, 6)
    V(g)$label.cex <- ifelse(V(g)$type == "Term", 0.8, 0.6)
    
    # Plot
    par(mar = c(1, 1, 3, 1))
    plot(g,
         layout = layout_with_fr(g),
         vertex.label.color = "black",
         vertex.label.family = "sans",
         vertex.frame.color = NA,
         edge.color = "gray70",
         edge.width = 0.5,
         main = "Gene-Concept Network\n(Top 5 Terms)")
    
    legend("topright",
           legend = c("GO Terms", "Genes"),
           col = c("#E31A1C", "#1F78B4"),
           pch = 19,
           pt.cex = 2,
           bty = "n")
  }

  # Create manual network plot
  cat("\n=== Creating Manual Network Plot ===\n\n")
  create_manual_network_plot(ego_all, top_terms = 5, max_genes_per_term = 8)




  # ╔══════════════════════════════════════════╗
  # ║ Section: test2
  # ╚══════════════════════════════════════════╝
  library(enrichplot)
  library(ggplot2)
  library(igraph)
  library(ggraph)
  library(dplyr)
  library(tidyr)

  # ============================================================================
  # METHOD 1: Fix enrichResult object for network plots
  # ============================================================================

  fix_enricher_for_networks <- function(ego, term2gene) {
    if(is.null(ego) || nrow(ego@result) == 0) {
      return(NULL)
    }
    
    # Make sure the geneID column is properly formatted
    # It should be a character string with genes separated by "/"
    if(!"geneID" %in% colnames(ego@result)) {
      stop("No geneID column found in enrichment results")
    }
    
    # Parse gene counts if not already done
    if(!"Count" %in% colnames(ego@result)) {
      ego@result$Count <- sapply(strsplit(ego@result$GeneRatio, "/"), 
                                  function(x) as.numeric(x[1]))
    }
    
    # Ensure the gene list is attached to the object
    # This is critical for cnetplot to work
    if(is.null(ego@gene)) {
      # Extract unique genes from all terms
      all_genes <- unique(unlist(strsplit(ego@result$geneID, "/")))
      ego@gene <- all_genes
    }
    
    return(ego)
  }

  # Apply the fix
  ego_all_fixed <- fix_enricher_for_networks(ego_all, term2gene)

  # Now try cnetplot
  if(!is.null(ego_all_fixed)) {
    tryCatch({
      p_cnet <- cnetplot(ego_all_fixed, 
                         categorySize = "pvalue",
                         showCategory = 5,
                         colorEdge = TRUE,
                         node_label = "category")
      print(p_cnet)
    }, error = function(e) {
      cat("Standard cnetplot failed, using custom method...\n")
      cat("Error was:", e$message, "\n")
    })
  }

  # ============================================================================
  # METHOD 2: Custom Gene-Concept Network (ALWAYS WORKS!)
  # ============================================================================

  create_custom_cnetplot <- function(ego, 
                                     term2gene,
                                     top_n = 5,
                                     title = "Gene-Concept Network") {
    
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No enrichment results to plot")
      return(NULL)
    }
    
    # Get top N terms
    top_terms <- ego@result[order(ego@result$pvalue), ][1:min(top_n, nrow(ego@result)), ]
    
    # Create gene-term edges
    edges_list <- list()
    
    for(i in 1:nrow(top_terms)) {
      term_id <- top_terms$ID[i]
      term_name <- top_terms$Description[i]
      genes <- unlist(strsplit(top_terms$geneID[i], "/"))
      
      if(length(genes) > 0) {
        edges_list[[i]] <- data.frame(
          from = term_name,
          to = genes,
          term_pvalue = top_terms$pvalue[i],
          stringsAsFactors = FALSE
        )
      }
    }
    
    # Combine all edges
    edges <- do.call(rbind, edges_list)
    
    if(nrow(edges) == 0) {
      message("No gene-term relationships found")
      return(NULL)
    }
    
    # Create node attributes
    nodes <- data.frame(
      name = unique(c(edges$from, edges$to)),
      stringsAsFactors = FALSE
    )
    
    # Identify node types
    nodes$type <- ifelse(nodes$name %in% top_terms$Description, "Term", "Gene")
    
    # Add term p-values
    nodes$pvalue <- NA
    for(term in top_terms$Description) {
      nodes$pvalue[nodes$name == term] <- top_terms$pvalue[top_terms$Description == term]
    }
    
    # Create igraph object
    g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
    
    # Create layout
    set.seed(123)
    layout <- create_layout(g, layout = 'fr')
    
    # Calculate node sizes
    layout$size <- ifelse(layout$type == "Term", 8, 3)
    
    # Create plot
    p <- ggraph(layout) +
      geom_edge_link(aes(edge_alpha = 0.3), color = "grey70", width = 0.8) +
      geom_node_point(aes(color = type, size = size, alpha = ifelse(type == "Term", -log10(pvalue), 0.7))) +
      geom_node_text(aes(label = ifelse(type == "Term", name, "")), 
                     size = 3, repel = TRUE, max.overlaps = 20) +
      scale_color_manual(values = c("Term" = "#E64B35", "Gene" = "#4DBBD5"),
                        name = "Node Type") +
      scale_size_continuous(range = c(3, 10), guide = "none") +
      scale_alpha_continuous(range = c(0.5, 1), guide = "none") +
      labs(title = title) +
      theme_graph(base_family = 'sans') +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "right"
      )
    
    return(p)
  }

  # Create custom gene-concept network
  p_custom_cnet <- create_custom_cnetplot(
    ego_all, 
    term2gene,
    top_n = 5,
    title = "Gene-Concept Network - Top 5 Terms"
  )

  if(!is.null(p_custom_cnet)) {
    tryCatch(
      print(p_custom_cnet),
      error = function(e) {
        message("\u26a0 Custom gene-concept network skipped (likely ggraph/igraph version mismatch): ",
                conditionMessage(e))
        message("  Fix permanently with: install.packages(c(\"ggraph\", \"igraph\", \"tidygraph\"))")
      }
    )
  }

  # ============================================================================
  # METHOD 3: Simplified Gene-Concept Network (Alternative style)
  # ============================================================================

  create_simple_cnetplot <- function(ego, top_n = 5, title = "Gene-Concept Network") {
    
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No enrichment results to plot")
      return(NULL)
    }
    
    # Get top terms
    top_terms <- ego@result[order(ego@result$pvalue), ][1:min(top_n, nrow(ego@result)), ]
    
    # Wrap long term names
    top_terms$Description_short <- stringr::str_wrap(top_terms$Description, width = 30)
    
    # Create edges
    edges <- data.frame()
    for(i in 1:nrow(top_terms)) {
      genes <- unlist(strsplit(top_terms$geneID[i], "/"))
      if(length(genes) > 0) {
        temp <- data.frame(
          Term = top_terms$Description_short[i],
          Gene = genes,
          pvalue = top_terms$pvalue[i],
          stringsAsFactors = FALSE
        )
        edges <- rbind(edges, temp)
      }
    }
    
    # Create graph
    g <- graph_from_data_frame(edges[, c("Term", "Gene")], directed = FALSE)
    
    # Set node attributes
    V(g)$type <- ifelse(V(g)$name %in% edges$Term, "Term", "Gene")
    V(g)$pvalue <- NA
    for(term in unique(edges$Term)) {
      V(g)$pvalue[V(g)$name == term] <- unique(edges$pvalue[edges$Term == term])
    }
    
    # Create layout
    set.seed(42)
    coords <- layout_with_fr(g)
    
    # Convert to data frame for ggplot
    node_df <- data.frame(
      name = V(g)$name,
      type = V(g)$type,
      pvalue = V(g)$pvalue,
      x = coords[, 1],
      y = coords[, 2],
      stringsAsFactors = FALSE
    )
    
    edge_df <- data.frame(
      from = ends(g, E(g))[, 1],
      to = ends(g, E(g))[, 2],
      stringsAsFactors = FALSE
    )
    
    # Add coordinates for edges
    edge_df$x <- node_df$x[match(edge_df$from, node_df$name)]
    edge_df$y <- node_df$y[match(edge_df$from, node_df$name)]
    edge_df$xend <- node_df$x[match(edge_df$to, node_df$name)]
    edge_df$yend <- node_df$y[match(edge_df$to, node_df$name)]
    
    # Plot
    p <- ggplot() +
      geom_segment(data = edge_df, 
                   aes(x = x, y = y, xend = xend, yend = yend),
                   color = "grey70", alpha = 0.4, linewidth = 0.5) +
      geom_point(data = node_df,
                 aes(x = x, y = y, color = type, 
                     size = ifelse(type == "Term", 10, 3),
                     alpha = ifelse(type == "Term", -log10(pvalue), 0.6))) +
      geom_text(data = node_df[node_df$type == "Term", ],
                aes(x = x, y = y, label = name),
                size = 3, fontface = "bold", color = "black",
                hjust = 0.5, vjust = -1.2, check_overlap = TRUE) +
      scale_color_manual(values = c("Term" = "#E64B35", "Gene" = "#4DBBD5"),
                        name = "Node Type") +
      scale_size_identity() +
      scale_alpha_continuous(range = c(0.5, 1), guide = "none") +
      labs(title = title) +
      theme_void() +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "right"
      )
    
    return(p)
  }

  # Create simple network
  p_simple_cnet <- create_simple_cnetplot(
    ego_all,
    top_n = 5,
    title = "Gene-Concept Network - Top 5 Terms"
  )

  if(!is.null(p_simple_cnet)) {
    tryCatch(
      print(p_simple_cnet),
      error = function(e) {
        message("\u26a0 Simple gene-concept network skipped (likely ggraph/igraph version mismatch): ",
                conditionMessage(e))
        message("  Fix permanently with: install.packages(c(\"ggraph\", \"igraph\", \"tidygraph\"))")
      }
    )
  }

  # ============================================================================
  # METHOD 4: Fix and Create Enrichment Map (emapplot)
  # ============================================================================

  create_custom_emapplot <- function(ego, top_n = 20, title = "Enrichment Map") {
    
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No enrichment results to plot")
      return(NULL)
    }
    
    # Get top terms
    top_terms <- ego@result[order(ego@result$pvalue), ][1:min(top_n, nrow(ego@result)), ]
    
    if(nrow(top_terms) < 2) {
      message("Need at least 2 terms for enrichment map")
      return(NULL)
    }
    
    # Calculate term similarity based on shared genes
    n_terms <- nrow(top_terms)
    sim_matrix <- matrix(0, nrow = n_terms, ncol = n_terms)
    rownames(sim_matrix) <- top_terms$Description
    colnames(sim_matrix) <- top_terms$Description
    
    for(i in 1:n_terms) {
      genes_i <- unlist(strsplit(top_terms$geneID[i], "/"))
      for(j in i:n_terms) {
        if(i == j) {
          sim_matrix[i, j] <- 1
        } else {
          genes_j <- unlist(strsplit(top_terms$geneID[j], "/"))
          # Jaccard similarity
          intersection <- length(intersect(genes_i, genes_j))
          union <- length(union(genes_i, genes_j))
          sim_matrix[i, j] <- sim_matrix[j, i] <- intersection / union
        }
      }
    }
    
    # Create edges for similar terms (similarity > 0.2)
    edges <- data.frame()
    for(i in 1:(n_terms-1)) {
      for(j in (i+1):n_terms) {
        if(sim_matrix[i, j] > 0.2) {
          edges <- rbind(edges, data.frame(
            from = rownames(sim_matrix)[i],
            to = colnames(sim_matrix)[j],
            similarity = sim_matrix[i, j],
            stringsAsFactors = FALSE
          ))
        }
      }
    }
    
    if(nrow(edges) == 0) {
      message("No significant similarities between terms (all < 0.2)")
      return(NULL)
    }
    
    # Create nodes dataframe
    nodes <- data.frame(
      name = top_terms$Description,
      pvalue = top_terms$pvalue,
      count = sapply(strsplit(top_terms$GeneRatio, "/"), function(x) as.numeric(x[1])),
      stringsAsFactors = FALSE
    )
    
    # Wrap long names
    nodes$name_short <- stringr::str_wrap(nodes$name, width = 25)
    
    # Create graph
    g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
    
    # Layout
    set.seed(123)
    layout <- create_layout(g, layout = 'fr')
    
    # Plot
    p <- ggraph(layout) +
      geom_edge_link(aes(edge_width = similarity, edge_alpha = similarity), 
                     color = "grey60") +
      geom_node_point(aes(size = count, color = -log10(pvalue))) +
      geom_node_text(aes(label = name_short), 
                     size = 2.5, repel = TRUE, max.overlaps = 30,
                     segment.size = 0.2) +
      scale_edge_width_continuous(range = c(0.5, 2), name = "Gene\nSimilarity") +
      scale_edge_alpha_continuous(range = c(0.3, 0.8), guide = "none") +
      scale_size_continuous(range = c(3, 10), name = "Gene\nCount") +
      scale_color_gradient(low = "blue", high = "red", name = "-log10\n(p-value)") +
      labs(title = title) +
      theme_graph(base_family = 'sans') +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "right"
      )
    
    return(p)
  }

  # Create custom enrichment map
  p_custom_emap <- create_custom_emapplot(
    ego_all,
    top_n = 15,
    title = "Enrichment Map - Term Similarities"
  )

  if(!is.null(p_custom_emap)) {
    tryCatch(
      print(p_custom_emap),
      error = function(e) {
        message("\u26a0 Custom enrichment map skipped (likely ggraph/igraph version mismatch): ",
                conditionMessage(e))
        message("  Fix permanently with: install.packages(c(\"ggraph\", \"igraph\", \"tidygraph\"))")
      }
    )
  }

  # ============================================================================
  # METHOD 5: Try standard emapplot with proper preparation
  # ============================================================================

  # For standard emapplot to work, we need to ensure term similarity is calculated
  tryCatch({
    # Calculate pairwise term similarity
    ego_all_sim <- pairwise_termsim(ego_all)
    
    # Now try emapplot
    p_emap_standard <- emapplot(ego_all_sim, 
                                showCategory = 15,
                                cex_label_category = 0.7)
    print(p_emap_standard)
    
  }, error = function(e) {
    cat("Standard emapplot failed:\n")
    cat("Error:", e$message, "\n")
    cat("Using custom method instead (see above)\n")
  })

  # ============================================================================
  # COMPLETE REPLACEMENT CODE FOR YOUR NETWORK SECTION
  # ============================================================================

  cat("\n\n")
  cat("=================================================================\n")
  cat("COPY THE CODE BELOW TO REPLACE YOUR NETWORK VISUALIZATION SECTION\n")
  cat("=================================================================\n\n")


  # ╔══════════════════════════════════════════╗
  # ║ Section: networks_2
  # ╚══════════════════════════════════════════╝
  library(igraph)
  library(ggraph)

  # Function to create custom gene-concept network
  create_custom_cnetplot <- function(ego, top_n = 5, title = "Gene-Concept Network") {
    
    if(is.null(ego) || nrow(ego@result) == 0) {
      message("No enrichment results to plot")
      return(NULL)
    }
    
    top_terms <- ego@result[order(ego@result$pvalue), ][1:min(top_n, nrow(ego@result)), ]
    
    # Create edges
    edges <- data.frame()
    for(i in 1:nrow(top_terms)) {
      genes <- unlist(strsplit(top_terms$geneID[i], "/"))
      if(length(genes) > 0) {
        temp <- data.frame(
          Term = stringr::str_wrap(top_terms$Description[i], width = 30),
          Gene = genes,
          pvalue = top_terms$pvalue[i],
          stringsAsFactors = FALSE
        )
        edges <- rbind(edges, temp)
      }
    }
    
    if(nrow(edges) == 0) return(NULL)
    
    # Create graph
    g <- graph_from_data_frame(edges[, c("Term", "Gene")], directed = FALSE)
    V(g)$type <- ifelse(V(g)$name %in% edges$Term, "Term", "Gene")
    V(g)$pvalue <- NA
    for(term in unique(edges$Term)) {
      V(g)$pvalue[V(g)$name == term] <- unique(edges$pvalue[edges$Term == term])
    }
    
    # Layout
    set.seed(42)
    layout <- create_layout(g, layout = "fr")
    
    # Plot
    p <- ggraph(layout) +
      geom_edge_link(aes(alpha = 0.3), color = "grey70", width = 0.8) +
      geom_node_point(aes(color = type, 
                         size = ifelse(type == "Term", 8, 3),
                         alpha = ifelse(type == "Term", -log10(pvalue), 0.7))) +
      geom_node_text(aes(label = ifelse(type == "Term", name, "")), 
                     size = 3, repel = TRUE, max.overlaps = 20) +
      scale_color_manual(values = c("Term" = "#E64B35", "Gene" = "#4DBBD5"),
                        name = "Node Type") +
      scale_size_identity() +
      scale_alpha_continuous(range = c(0.5, 1), guide = "none") +
      labs(title = title) +
      theme_graph() +
      theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
            legend.position = "right")
    
    return(p)
  }

  # Function to create enrichment map
  create_custom_emapplot <- function(ego, top_n = 15, title = "Enrichment Map") {
    
    if(is.null(ego) || nrow(ego@result) == 0) return(NULL)
    
    top_terms <- ego@result[order(ego@result$pvalue), ][1:min(top_n, nrow(ego@result)), ]
    if(nrow(top_terms) < 2) return(NULL)
    
    # Calculate similarity
    n_terms <- nrow(top_terms)
    sim_matrix <- matrix(0, nrow = n_terms, ncol = n_terms)
    
    for(i in 1:n_terms) {
      genes_i <- unlist(strsplit(top_terms$geneID[i], "/"))
      for(j in i:n_terms) {
        if(i == j) {
          sim_matrix[i, j] <- 1
        } else {
          genes_j <- unlist(strsplit(top_terms$geneID[j], "/"))
          intersection <- length(intersect(genes_i, genes_j))
          union <- length(union(genes_i, genes_j))
          sim_matrix[i, j] <- sim_matrix[j, i] <- intersection / union
        }
      }
    }
    
    # Create edges
    edges <- data.frame()
    for(i in 1:(n_terms-1)) {
      for(j in (i+1):n_terms) {
        if(sim_matrix[i, j] > 0.2) {
          edges <- rbind(edges, data.frame(
            from = i, to = j,
            similarity = sim_matrix[i, j]
          ))
        }
      }
    }
    
    if(nrow(edges) == 0) {
      message("No significant term similarities found")
      return(NULL)
    }
    
    # Create nodes
    nodes <- data.frame(
      id = 1:nrow(top_terms),
      name = stringr::str_wrap(top_terms$Description, width = 25),
      pvalue = top_terms$pvalue,
      count = sapply(strsplit(top_terms$GeneRatio, "/"), function(x) as.numeric(x[1]))
    )
    
    # Create graph
    g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
    
    # Layout
    set.seed(123)
    layout <- create_layout(g, layout = "fr")
    
    # Plot
    p <- ggraph(layout) +
      geom_edge_link(aes(edge_width = similarity, edge_alpha = similarity), 
                     color = "grey60") +
      geom_node_point(aes(size = count, color = -log10(pvalue))) +
      geom_node_text(aes(label = name), size = 2.5, repel = TRUE, 
                     max.overlaps = 30) +
      scale_edge_width_continuous(range = c(0.5, 2), name = "Similarity") +
      scale_edge_alpha_continuous(range = c(0.3, 0.8), guide = "none") +
      scale_size_continuous(range = c(3, 10), name = "Gene Count") +
      scale_color_gradient(low = "blue", high = "red", name = "-log10(p)") +
      labs(title = title) +
      theme_graph() +
      theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
    
    return(p)
  }

  # Create Gene-Concept Network
  if(!is.null(ego_all) && nrow(ego_all@result) >= 5) {
    cat("Creating gene-concept network...\\n")
    p_cnet <- create_custom_cnetplot(ego_all, top_n = 5, 
                                     title = "Gene-Concept Network - Top 5 Terms")
    if(!is.null(p_cnet)) {
      tryCatch(
        print(p_cnet),
        error = function(e) {
          message("\u26a0 Gene-concept network skipped (likely ggraph/igraph version mismatch): ",
                  conditionMessage(e))
          message("  Fix permanently with: install.packages(c(\"ggraph\", \"igraph\", \"tidygraph\"))")
        }
      )
    }
  }

  # Create Enrichment Map
  if(!is.null(ego_all) && nrow(ego_all@result) >= 5) {
    cat("Creating enrichment map...\\n")
    p_emap <- create_custom_emapplot(ego_all, top_n = 15,
                                     title = "Enrichment Map - Term Similarities")
    if(!is.null(p_emap)) {
      tryCatch(
        print(p_emap),
        error = function(e) {
          message("\u26a0 Enrichment map skipped (likely ggraph/igraph version mismatch): ",
                  conditionMessage(e))
          message("  Fix permanently with: install.packages(c(\"ggraph\", \"igraph\", \"tidygraph\"))")
        }
      )
    }
  }





  # ╔══════════════════════════════════════════╗
  # ║ Section: networks_3
  # ╚══════════════════════════════════════════╝
  library(igraph)
  library(ggraph)


  # Function to create enrichment map
  create_custom_emapplot <- function(ego, top_n = 15, title = "Enrichment Map") {
    
    if(is.null(ego) || nrow(ego@result) == 0) return(NULL)
    
    top_terms <- ego@result[order(ego@result$pvalue), ][1:min(top_n, nrow(ego@result)), ]
    if(nrow(top_terms) < 2) return(NULL)
    
    # Calculate similarity
    n_terms <- nrow(top_terms)
    sim_matrix <- matrix(0, nrow = n_terms, ncol = n_terms)
    
    for(i in 1:n_terms) {
      genes_i <- unlist(strsplit(top_terms$geneID[i], "/"))
      for(j in i:n_terms) {
        if(i == j) {
          sim_matrix[i, j] <- 1
        } else {
          genes_j <- unlist(strsplit(top_terms$geneID[j], "/"))
          intersection <- length(intersect(genes_i, genes_j))
          union <- length(union(genes_i, genes_j))
          sim_matrix[i, j] <- sim_matrix[j, i] <- intersection / union
        }
      }
    }
    
    # Create edges
    edges <- data.frame()
    for(i in 1:(n_terms-1)) {
      for(j in (i+1):n_terms) {
        if(sim_matrix[i, j] > 0.2) {
          edges <- rbind(edges, data.frame(
            from = i, to = j,
            similarity = sim_matrix[i, j]
          ))
        }
      }
    }
    
    if(nrow(edges) == 0) {
      message("No significant term similarities found")
      return(NULL)
    }
    
    # Create nodes
    nodes <- data.frame(
      id = 1:nrow(top_terms),
      name = stringr::str_wrap(top_terms$Description, width = 25),
      pvalue = top_terms$pvalue,
      count = sapply(strsplit(top_terms$GeneRatio, "/"), function(x) as.numeric(x[1]))
    )
    
    # Create graph
    g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
    
    # Layout
    set.seed(123)
    layout <- create_layout(g, layout = "fr")
    
    # Plot
    p <- ggraph(layout) +
      geom_edge_link(aes(edge_width = similarity, edge_alpha = similarity), 
                     color = "grey60") +
      geom_node_point(aes(size = count, color = -log10(pvalue))) +
      geom_node_text(aes(label = name), size = 2.5, repel = TRUE, 
                     max.overlaps = 30) +
      scale_edge_width_continuous(range = c(0.5, 2), name = "Similarity") +
      scale_edge_alpha_continuous(range = c(0.3, 0.8), guide = "none") +
      scale_size_continuous(range = c(3, 10), name = "Gene Count") +
      scale_color_gradient(low = "blue", high = "red", name = "-log10(p)") +
      labs(title = title) +
      theme_graph() +
      theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
    
    return(p)
  }


  # Create Enrichment Map
  if(!is.null(ego_all) && nrow(ego_all@result) >= 5) {
    cat("Creating enrichment map...\\n")
    p_emap <- create_custom_emapplot(ego_all, top_n = 15,
                                     title = "Enrichment Map - Term Similarities")
    if(!is.null(p_emap)) {
      tryCatch(
        print(p_emap),
        error = function(e) {
          message("\u26a0 Enrichment map skipped (likely ggraph/igraph version mismatch): ",
                  conditionMessage(e))
          message("  Fix permanently with: install.packages(c(\"ggraph\", \"igraph\", \"tidygraph\"))")
        }
      )
    }
  }






  # ╔══════════════════════════════════════════╗
  # ║ Section: save_results
  # ╚══════════════════════════════════════════╝
  cat("\n=== SAVING ENRICHMENT RESULTS ===\n\n")

  # Create output directory
  output_dir <- paste0("./sclerotinia_pathogen_enrichment_results", tp_suffix)
  if(!dir.exists(output_dir)) {
    dir.create(output_dir)
    cat("✓ Created output directory:", output_dir, "\n")
  }

  # Save all DEGs enrichment
  if(!is.null(ego_all) && nrow(ego_all@result) > 0) {
    write.csv(ego_all@result, 
              file.path(output_dir, "pathogen_all_DEGs_GO_enrichment.csv"),
              row.names = FALSE)
    cat("✓ Saved: pathogen_all_DEGs_GO_enrichment.csv\n")
  }

  # Save up-regulated enrichment
  if(!is.null(ego_up) && nrow(ego_up@result) > 0) {
    write.csv(ego_up@result,
              file.path(output_dir, "pathogen_upregulated_GO_enrichment.csv"),
              row.names = FALSE)
    cat("✓ Saved: pathogen_upregulated_GO_enrichment.csv\n")
  }

  # Save down-regulated enrichment
  if(!is.null(ego_down) && nrow(ego_down@result) > 0) {
    write.csv(ego_down@result,
              file.path(output_dir, "pathogen_downregulated_GO_enrichment.csv"),
              row.names = FALSE)
    cat("✓ Saved: pathogen_downregulated_GO_enrichment.csv\n")
  }

  # Save gene ID mapping (all three ID types)
  gene_mapping <- pathogen_degs %>%
    select(protein_id, gene_name, original_gene_id) %>%
    filter(!is.na(protein_id))

  write.csv(gene_mapping,
            file.path(output_dir, "gene_id_mapping_complete.csv"),
            row.names = FALSE)
  cat("✓ Saved: gene_id_mapping_complete.csv\n")

  # Save DEG file with protein IDs for future use
  write.csv(pathogen_degs,
            file.path(output_dir, "pathogen_DEGs_with_protein_ids.csv"),
            row.names = FALSE)
  cat("✓ Saved: pathogen_DEGs_with_protein_ids.csv\n")

  cat("\nAll results saved to:", output_dir, "\n")


  # ╔══════════════════════════════════════════╗
  # ║ Section: summary
  # ╚══════════════════════════════════════════╝
  cat("\n=== ANALYSIS SUMMARY ===\n\n")

  cat("Input Data:\n")
  cat("  Total DEGs:", nrow(pathogen_degs), "\n")
  cat("  Up-regulated:", nrow(upregulated), "\n")
  cat("  Down-regulated:", nrow(downregulated), "\n\n")

  cat("Gene ID Transformation:\n")
  cat("  Original format: Ss_jgi|Sclsc1|PROTEIN_ID|GENE_NAME\n")
  cat("  Extracted: Numeric protein IDs\n")
  cat("  Successfully parsed:", sum(!is.na(pathogen_degs$protein_id)), "genes\n\n")

  cat("Annotation Coverage:\n")
  cat("  Genes with GO annotation:", length(overlap_ids), "\n")
  cat("  Coverage:", round(100 * length(overlap_ids) / length(deg_protein_ids), 1), "%\n")
  cat("  (Note: 20-40% is NORMAL and EXPECTED for fungal pathogens)\n\n")

  cat("Enrichment Results:\n")
  if(!is.null(ego_all) && nrow(ego_all@result) > 0) {
    cat("  All DEGs:", nrow(ego_all@result), "enriched terms (p < 0.05)\n")
  }
  if(!is.null(ego_up) && nrow(ego_up@result) > 0) {
    cat("  Up-regulated:", nrow(ego_up@result), "enriched terms (p < 0.05)\n")
  }
  if(!is.null(ego_down) && nrow(ego_down@result) > 0) {
    cat("  Down-regulated:", nrow(ego_down@result), "enriched terms (p < 0.05)\n")
  }

  cat("\n✓ Analysis complete!\n")
  cat("  Results saved in:", output_dir, "\n\n")

  cat("Next Steps:\n")
  cat("  1. Review enriched GO terms for biological insights\n")
  cat("  2. Focus on pathogenicity-related processes (secretion, degradation, etc.)\n")
  cat("  3. Consider eggNOG-mapper for CAZyme and effector annotation\n")
  cat("  4. Validate top candidate genes with literature or experiments\n")


  # ─── Stash results for cross-timepoint summary ───
  master_results[[tp]] <- list(
    n_degs        = if(exists("pathogen_degs"))     nrow(pathogen_degs)    else NA_integer_,
    n_up          = if(exists("upregulated"))       nrow(upregulated)      else NA_integer_,
    n_down        = if(exists("downregulated"))     nrow(downregulated)    else NA_integer_,
    coverage_pct  = if(exists("coverage_pct"))      coverage_pct           else NA_real_,
    ego_all       = if(exists("ego_all"))           ego_all                else NULL,
    ego_up        = if(exists("ego_up"))            ego_up                 else NULL,
    ego_down      = if(exists("ego_down"))          ego_down               else NULL
  )

  cat(sprintf("\n✓ Timepoint %s complete\n\n", tp))
}  # end for-loop

cat("\n\nAll timepoints processed.\n")

```

---

# Cross-Timepoint Summary

## DEG and Enrichment Counts per Timepoint

```{r cross_tp_summary}
cat("\n=== CROSS-TIMEPOINT SUMMARY ===\n\n")

summary_df <- data.frame(
  Timepoint    = character(),
  Total_DEGs   = integer(),
  Up           = integer(),
  Down         = integer(),
  Coverage_pct = numeric(),
  GO_all       = integer(),
  GO_up        = integer(),
  GO_down      = integer(),
  stringsAsFactors = FALSE
)

for (tp in names(master_results)) {
  r <- master_results[[tp]]
  summary_df <- rbind(summary_df, data.frame(
    Timepoint    = tp,
    Total_DEGs   = ifelse(is.na(r$n_degs),       0L, r$n_degs),
    Up           = ifelse(is.na(r$n_up),         0L, r$n_up),
    Down         = ifelse(is.na(r$n_down),       0L, r$n_down),
    Coverage_pct = ifelse(is.na(r$coverage_pct), 0,  r$coverage_pct),
    GO_all       = ifelse(!is.null(r$ego_all),  nrow(r$ego_all@result),  0L),
    GO_up        = ifelse(!is.null(r$ego_up),   nrow(r$ego_up@result),   0L),
    GO_down      = ifelse(!is.null(r$ego_down), nrow(r$ego_down@result), 0L),
    stringsAsFactors = FALSE
  ))
}

print(summary_df)

write.csv(summary_df,
          "./sclerotinia_pathogen_cross_timepoint_summary.csv",
          row.names = FALSE)
cat("\n✓ Saved: sclerotinia_pathogen_cross_timepoint_summary.csv\n")
```

## GO Term Overlap Across Timepoints

```{r cross_tp_overlap, fig.width=10, fig.height=6}
go_terms_by_tp <- list()
for (tp in names(master_results)) {
  r <- master_results[[tp]]
  if(!is.null(r$ego_all) && nrow(r$ego_all@result) > 0) {
    sig <- r$ego_all@result[r$ego_all@result$pvalue < 0.05, ]
    go_terms_by_tp[[tp]] <- sig$ID
  } else {
    go_terms_by_tp[[tp]] <- character(0)
  }
}

if(length(go_terms_by_tp) >= 2) {
  cat("\n=== GO TERM OVERLAP ACROSS TIMEPOINTS ===\n\n")
  for (i in 1:(length(go_terms_by_tp) - 1)) {
    for (j in (i+1):length(go_terms_by_tp)) {
      tp_i <- names(go_terms_by_tp)[i]
      tp_j <- names(go_terms_by_tp)[j]
      shared <- length(intersect(go_terms_by_tp[[tp_i]], go_terms_by_tp[[tp_j]]))
      cat(sprintf("  %s ∩ %s: %d shared GO terms\n", tp_i, tp_j, shared))
    }
  }
  if(length(go_terms_by_tp) == 3) {
    shared_all <- Reduce(intersect, go_terms_by_tp)
    cat(sprintf("  All three timepoints: %d shared GO terms\n\n", length(shared_all)))
    if(length(shared_all) > 0) {
      cat("GO terms enriched at ALL three timepoints:\n\n")
      shared_names <- term2name[term2name$GO %in% shared_all, ]
      print(shared_names)
      write.csv(shared_names,
                "./sclerotinia_pathogen_GO_shared_all_timepoints.csv",
                row.names = FALSE)
      cat("\n✓ Saved: sclerotinia_pathogen_GO_shared_all_timepoints.csv\n")
    }
  }
}
```

---

```{r session_info}
sessionInfo()
```
