---
title: "Isolate-Level ANOVA & Aggressiveness Group Confirmation"
subtitle: "Per-gene ANOVA on S. sclerotiorum isolate effect with emmeans/CLD"
author: "Analysis by wan00965"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
```

# Overview

**Goal:** Determine whether *S. sclerotiorum* isolates cluster by
aggressiveness phenotype at the transcriptional level.

**Hypothesis — Expected grouping:**

| Group | Isolates | Color |
|-------|----------|-------|
| Low aggressive | WISS47, JS659 | Blue |
| High aggressive | MNSS6, Xtra7 | Red |
| Soybean-specific | SSPotter | Brown |
| Sunflower-specific | BN172 | Purple |

**Approach:**

1. Per-gene ANOVA (isolate + timepoint + block) → genes with significant isolate effect
2. emmeans + Tukey pairwise + CLD on each significant gene
3. CLD concordance → do WISS47 & JS659 share a group? MNSS6 & Xtra7?
4. Planned contrasts → High vs Low, SSPotter vs BN172, Specialists vs Generalists
5. Heatmap of significant genes → visual confirmation of clustering
6. Block (biological replicate) variance check

**Four datasets:**

- A. Pathogen genes on soybean
- B. Pathogen genes on sunflower
- C. Soybean host genes
- D. Sunflower host genes

# Setup

```{r load-libraries}
library(DESeq2)
library(car)
library(emmeans)
library(multcomp)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(RColorBrewer)
library(gridExtra)
library(vegan)

cat("All libraries loaded.\n")
```

```{r set-directory}
# TODO: set the working directory to the folder containing your STAR count matrices and DEG CSVs
# setwd("/path/to/your/STAR/output")
dir.create("isolate_anova_results", showWarnings = FALSE)
```

# Load Data

```{r load-counts}
counts_soy_host     <- read.table("soybean_host_only_STAR.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)
counts_soy_pathogen <- read.table("pathogen_soybean_STAR.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)
counts_sun_host     <- read.table("sunflower_host_only_STAR.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)
counts_sun_pathogen <- read.table("pathogen_sunflower_STAR.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)

cat(sprintf("Soybean:   %d host genes, %d pathogen genes, %d samples\n",
            nrow(counts_soy_host), nrow(counts_soy_pathogen), ncol(counts_soy_host)))
cat(sprintf("Sunflower: %d host genes, %d pathogen genes, %d samples\n",
            nrow(counts_sun_host), nrow(counts_sun_pathogen), ncol(counts_sun_host)))
```

## Build Metadata

```{r build-metadata}
build_metadata <- function(sample_names) {
  meta <- data.frame(
    sample    = sample_names,
    timepoint = sapply(sample_names, function(x) {
      parts <- strsplit(x, "_")[[1]]
      if (parts[length(parts)] == "NC") "NC" else paste0(parts[1], "h")
    }),
    block   = sapply(strsplit(sample_names, "_"), `[`, 2),
    isolate = sapply(strsplit(sample_names, "_"), `[`, 4),
    row.names = sample_names,
    stringsAsFactors = FALSE
  )
  # Remap blocks 4,5,6 → 1,2,3 so that biological replicates are

  # consistently numbered across timepoint batches in both crops.
  block_remap <- c("4" = "1", "5" = "2", "6" = "3")
  idx <- meta$block %in% names(block_remap)
  if (any(idx)) {
    cat(sprintf("  Remapping %d samples: block 4->1, 5->2, 6->3\n", sum(idx)))
    meta$block[idx] <- block_remap[meta$block[idx]]
  }
  meta$aggr_group <- dplyr::case_when(
    meta$isolate %in% c("WISS47", "JS659")  ~ "Low",
    meta$isolate %in% c("MNSS6", "Xtra7")   ~ "High",
    meta$isolate == "SSPotter"               ~ "Soy-specific",
    meta$isolate == "BN172"                  ~ "Sun-specific",
    meta$isolate == "NC"                     ~ "NC",
    TRUE                                     ~ NA_character_
  )
  meta$timepoint  <- factor(meta$timepoint,  levels = c("NC","24h","48h","96h"))
  meta$block      <- factor(meta$block, levels = c("1", "2", "3"))
  meta$isolate    <- factor(meta$isolate,
                            levels = c("NC","WISS47","JS659","MNSS6","Xtra7","BN172","SSPotter"))
  meta$aggr_group <- factor(meta$aggr_group,
                            levels = c("Low","High","Soy-specific","Sun-specific","NC"))
  return(meta)
}

meta_soy <- build_metadata(colnames(counts_soy_host))
meta_sun <- build_metadata(colnames(counts_sun_host))

cat("── Soybean samples ──\n")
print(table(meta_soy$isolate, meta_soy$timepoint))
cat("\n── Soybean: timepoint x block (after remapping) ──\n")
print(table(meta_soy$timepoint, meta_soy$block))
cat("\n── Sunflower samples ──\n")
print(table(meta_sun$isolate, meta_sun$timepoint))
cat("\n── Sunflower: timepoint x block (after remapping) ──\n")
print(table(meta_sun$timepoint, meta_sun$block))
```

## Prepare VST Objects (exclude NC)

```{r prepare-vst}
make_vst <- function(counts, metadata, design_formula,
                     min_count = 10, min_samples = 3) {
  counts <- round(counts)
  keep   <- rowSums(counts >= min_count) >= min_samples
  counts <- counts[keep, ]
  dds    <- DESeqDataSetFromMatrix(countData = counts,
                                   colData   = metadata,
                                   design    = design_formula)
  dds    <- DESeq(dds)
  # varianceStabilizingTransformation() instead of vst() so this works on
  # the small pathogen matrices (vst() needs >=1000 genes with mean count > 5
  # to subsample for parametric fit; pathogen matrices are typically smaller).
  vst_obj <- varianceStabilizingTransformation(dds, blind = FALSE)
  idx     <- vst_obj$timepoint != "NC"
  vst_obj <- vst_obj[, idx]
  colData(vst_obj) <- droplevels(colData(vst_obj))
  return(vst_obj)
}

cat("Building VST objects (this takes a few minutes)...\n\n")

vst_soy_pathogen <- make_vst(counts_soy_pathogen, meta_soy, ~ timepoint)
cat(sprintf("Soybean pathogen: %d genes x %d samples\n",
            nrow(vst_soy_pathogen), ncol(vst_soy_pathogen)))

vst_sun_pathogen <- make_vst(counts_sun_pathogen, meta_sun, ~ timepoint)
cat(sprintf("Sunflower pathogen: %d genes x %d samples\n",
            nrow(vst_sun_pathogen), ncol(vst_sun_pathogen)))

vst_soy_host <- make_vst(counts_soy_host, meta_soy, ~ timepoint)
cat(sprintf("Soybean host: %d genes x %d samples\n",
            nrow(vst_soy_host), ncol(vst_soy_host)))

vst_sun_host <- make_vst(counts_sun_host, meta_sun, ~ timepoint)
cat(sprintf("Sunflower host: %d genes x %d samples\n",
            nrow(vst_sun_host), ncol(vst_sun_host)))
```

## Diagnostic: verify metadata & test one gene

```{r diagnostic-check}
cat("── colData columns in vst_soy_pathogen ──\n")
print(colnames(colData(vst_soy_pathogen)))

cat("\n── Factor levels after dropping NC ──\n")
cat("  isolate:  ", levels(vst_soy_pathogen$isolate), "\n")
cat("  timepoint:", levels(vst_soy_pathogen$timepoint), "\n")
cat("  block:    ", levels(vst_soy_pathogen$block), "\n")

cat("\n── Cross-tabulation: isolate x timepoint x block ──\n")
print(ftable(xtabs(~ isolate + timepoint + block,
                    data = as.data.frame(colData(vst_soy_pathogen)))))

cat("\n── timepoint x block crossing (after remapping) ──\n")
diag_meta <- as.data.frame(colData(vst_soy_pathogen))
print(table(diag_meta$timepoint, diag_meta$block))

# Df-based confounding test
diag_fit     <- lm(assay(vst_soy_pathogen)[1, ] ~ isolate + timepoint + block, data = diag_meta)
diag_av      <- anova(diag_fit)
tp_df_actual <- diag_av["timepoint", "Df"]
tp_df_expect <- nlevels(diag_meta$timepoint) - 1
cat(sprintf("\n  Timepoint Df with block in model: %d  (expected: %d)  -->  %s\n",
            tp_df_actual, tp_df_expect,
            ifelse(tp_df_actual == tp_df_expect,
                   "GOOD - block is estimable",
                   "PROBLEM - block confounded")))

# Test ANOVA on first gene
cat("\n── Test ANOVA on first gene ──\n")
test_expr <- assay(vst_soy_pathogen)[1, ]
test_meta <- as.data.frame(colData(vst_soy_pathogen))
test_df   <- data.frame(expr = test_expr, test_meta)

cat(sprintf("  Samples: %d, isolate levels: %d, timepoint levels: %d, block levels: %d\n",
            nrow(test_df), nlevels(test_df$isolate),
            nlevels(test_df$timepoint), nlevels(test_df$block)))

test_fit <- tryCatch(
  lm(expr ~ isolate + timepoint + block, data = test_df),
  error = function(e) { cat("  lm() ERROR:", conditionMessage(e), "\n"); NULL }
)

if (!is.null(test_fit)) {
  cat(sprintf("  lm() succeeded. Residual df: %d\n", test_fit$df.residual))
  al <- alias(test_fit)
  if (!is.null(al$Complete) && nrow(al$Complete) > 0) {
    cat("  WARNING: Aliased coefficients:\n")
    print(al$Complete)
  }

  # Try all Anova types
  for (atype in c("III", "II")) {
    av <- tryCatch({
      if (atype == "III") options(contrasts = c("contr.sum", "contr.poly"))
      Anova(test_fit, type = atype)
    }, error = function(e) NULL)
    if (!is.null(av)) {
      cat(sprintf("  Type %s ANOVA works.\n", atype))
      cat("  Row names:", paste(rownames(av), collapse = ", "), "\n")
      print(av)
      break
    } else {
      cat(sprintf("  Type %s ANOVA failed.\n", atype))
    }
  }
}
```


# Reusable ANOVA Engine

```{r define-engine}
isolate_colors <- c("WISS47" = "#3ACAFF", "JS659"   = "#00BFB2",
                     "MNSS6"  = "#FF3173", "Xtra7"   = "#FF5F3A",
                     "BN172"  = "#e682fa", "SSPotter" = "#8c564b")
aggr_colors <- c("Low" = "#1f77b4", "High" = "#d62728",
                  "Soy-specific" = "#8c564b", "Sun-specific" = "#e682fa")
tp_colors <- c("24h" = "#FFD700", "48h" = "#FF6347", "96h" = "#8B0000")

# ── Safe ANOVA: tries Type III, II, then I ──
safe_anova <- function(fit) {
  av <- tryCatch({
    options(contrasts = c("contr.sum", "contr.poly"))
    Anova(fit, type = "III")
  }, error = function(e) NULL)
  if (!is.null(av)) return(av)
  av <- tryCatch(Anova(fit, type = "II"), error = function(e) NULL)
  if (!is.null(av)) return(av)
  tryCatch(anova(fit), error = function(e) NULL)
}

# ── Extract F and p robustly from any anova-like table ──
extract_stats <- function(av, term) {
  av_df <- as.data.frame(av)
  f_col <- grep("^F", colnames(av_df), value = TRUE)[1]
  p_col <- grep("Pr\\(", colnames(av_df), value = TRUE)[1]
  if (term %in% rownames(av_df) && !is.na(f_col) && !is.na(p_col)) {
    return(c(F_val = av_df[term, f_col], p_val = av_df[term, p_col]))
  }
  c(F_val = NA_real_, p_val = NA_real_)
}

# ════════════════════════════════════════════════════════════════════
# CORE: per-gene ANOVA -> pairwise -> CLD -> planned contrasts
# ════════════════════════════════════════════════════════════════════
run_isolate_anova <- function(vst_obj, label, out_prefix) {

  expr_mat <- assay(vst_obj)
  meta     <- as.data.frame(colData(vst_obj))
  n_genes  <- nrow(expr_mat)

  cat(sprintf("\n====== %s: %d genes x %d samples ======\n",
              label, n_genes, ncol(expr_mat)))
  cat(sprintf("  Factors: isolate(%d) timepoint(%d) block(%d)\n",
              nlevels(meta$isolate), nlevels(meta$timepoint), nlevels(meta$block)))

  # Diagnostic: show timepoint x block crossing
  tp_block_tab <- table(meta$timepoint, meta$block)
  cat("\n-- timepoint x block cross-tab --\n")
  print(tp_block_tab)

  # Safety check: verify timepoint Df is not absorbed by block
  test_fit_full <- lm(expr_mat[1, ] ~ isolate + timepoint + block, data = meta)
  av_test       <- anova(test_fit_full)
  tp_df_actual  <- av_test["timepoint", "Df"]
  tp_df_expect  <- nlevels(meta$timepoint) - 1
  use_block     <- (tp_df_actual == tp_df_expect)

  if (!use_block) {
    cat(sprintf("\n  ** WARNING: block partially confounded (timepoint Df = %d, expected %d). Dropping block. **\n",
                tp_df_actual, tp_df_expect))
    model_formula_str <- "isolate + timepoint"
  } else {
    cat(sprintf("\n  Block OK (timepoint Df = %d as expected) -- including in models.\n",
                tp_df_actual))
    model_formula_str <- "isolate + timepoint + block"
  }

  # 1. PERMANOVA
  dist_mat <- vegdist(t(expr_mat), method = "euclidean")
  set.seed(42)
  perm_formula <- as.formula(paste("dist_mat ~", model_formula_str))
  perm_res <- adonis2(perm_formula,
                       data = meta, permutations = 999,
                       method = "euclidean", by = "margin")
  cat("\n-- PERMANOVA (marginal) --\n")
  print(perm_res)
  write.csv(as.data.frame(perm_res),
            file.path("isolate_anova_results", paste0(out_prefix, "_PERMANOVA.csv")))

  # 2. Per-gene ANOVA
  cat("Running per-gene ANOVA...\n")
  first_error <- NULL

  anova_list <- lapply(seq_len(n_genes), function(i) {
    df <- data.frame(expr = expr_mat[i, ], meta)
    gene_formula <- as.formula(paste("expr ~", model_formula_str))
    fit <- tryCatch(
      lm(gene_formula, data = df),
      error = function(e) {
        if (is.null(first_error)) first_error <<- conditionMessage(e)
        NULL
      })
    if (is.null(fit)) return(NULL)

    av <- safe_anova(fit)
    if (is.null(av)) {
      if (is.null(first_error)) first_error <<- "all Anova types failed"
      return(NULL)
    }

    iso <- extract_stats(av, "isolate")
    tp  <- extract_stats(av, "timepoint")
    blk <- if (use_block) extract_stats(av, "block") else c(F_val = NA_real_, p_val = NA_real_)

    data.frame(gene        = rownames(expr_mat)[i],
               F_isolate   = iso["F_val"],  p_isolate   = iso["p_val"],
               F_timepoint = tp["F_val"],   p_timepoint = tp["p_val"],
               F_block     = blk["F_val"],  p_block     = blk["p_val"],
               stringsAsFactors = FALSE, row.names = NULL)
  })

  valid <- Filter(Negate(is.null), anova_list)

  if (length(valid) == 0) {
    cat("  ALL per-gene ANOVAs failed!\n")
    if (!is.null(first_error)) cat("  First error:", first_error, "\n")
    return(list(pergene = data.frame(), perm = perm_res,
                pairwise = NULL, cld_all = NULL,
                concordance = NULL, planned = NULL))
  }

  pergene <- do.call(rbind, valid)
  n_ok    <- nrow(pergene)
  if (n_ok < n_genes) {
    cat(sprintf("  %d / %d genes failed (%.1f%%). First error: %s\n",
                n_genes - n_ok, n_genes, 100*(n_genes - n_ok)/n_genes,
                ifelse(is.null(first_error), "unknown", first_error)))
  }

  pergene$padj_isolate   <- p.adjust(pergene$p_isolate,   method = "BH")
  pergene$padj_timepoint <- p.adjust(pergene$p_timepoint, method = "BH")
  pergene$padj_block     <- p.adjust(pergene$p_block,     method = "BH")

  n_sig_iso <- sum(pergene$padj_isolate   < 0.05, na.rm = TRUE)
  n_sig_tp  <- sum(pergene$padj_timepoint < 0.05, na.rm = TRUE)
  n_sig_blk <- sum(pergene$padj_block     < 0.05, na.rm = TRUE)

  cat(sprintf("\n-- Per-gene ANOVA summary (padj < 0.05) --\n"))
  cat(sprintf("  Isolate:   %d / %d (%.1f%%)\n", n_sig_iso, n_ok, 100*n_sig_iso/n_ok))
  cat(sprintf("  Timepoint: %d / %d (%.1f%%)\n", n_sig_tp,  n_ok, 100*n_sig_tp/n_ok))
  cat(sprintf("  Block:     %d / %d (%.1f%%)\n", n_sig_blk, n_ok, 100*n_sig_blk/n_ok))

  write.csv(pergene, file.path("isolate_anova_results",
                                paste0(out_prefix, "_pergene_ANOVA.csv")),
            row.names = FALSE)

  # 3. Block variation diagnostic
  cat(sprintf("\n-- Block (biological replicate) variation --\n"))
  if (!use_block) {
    cat("  Block dropped (confounded with timepoint). No block test available.\n")
  } else {
    blk_pct <- 100 * n_sig_blk / n_ok
    cat(sprintf("  Significant block effect: %d (%.1f%%)\n", n_sig_blk, blk_pct))
    if      (blk_pct < 5)  cat("  -> Low. Replicates are consistent.\n")
    else if (blk_pct < 15) cat("  -> Moderate. Some replicate variation.\n")
    else                    cat("  -> High. Consider batch correction.\n")
  }

  # 4. emmeans + Tukey pairwise + CLD
  sig_genes <- pergene$gene[which(pergene$padj_isolate < 0.05)]

  cat(sprintf("\n-- emmeans pairwise on %d isolate-sig genes --\n",
              length(sig_genes)))

  if (length(sig_genes) == 0) {
    cat("  No significant isolate-effect genes.\n")
    return(list(pergene = pergene, perm = perm_res,
                pairwise = NULL, cld_all = NULL,
                concordance = NULL, planned = NULL))
  }

  genes_to_test <- head(sig_genes, 1000)
  pw_list  <- list()
  cld_list <- list()
  pw_errors <- 0

  for (g in genes_to_test) {
    ok <- tryCatch({
      df  <- data.frame(expr = expr_mat[g, ], meta)
      gene_formula <- as.formula(paste("expr ~", model_formula_str))
      fit <- lm(gene_formula, data = df)
      emm <- emmeans(fit, ~ isolate)
      pw  <- pairs(emm, adjust = "tukey")
      pw_df      <- as.data.frame(summary(pw));  pw_df$gene <- g
      cld_df     <- as.data.frame(cld(emm, Letters = letters, adjust = "tukey"))
      cld_df$gene <- g
      pw_list[[g]]  <<- pw_df
      cld_list[[g]] <<- cld_df
      TRUE
    }, error = function(e) { pw_errors <<- pw_errors + 1; FALSE })
  }
  if (pw_errors > 0)
    cat(sprintf("  %d emmeans errors (skipped)\n", pw_errors))

  if (length(pw_list) == 0) {
    cat("  All emmeans failed.\n")
    return(list(pergene = pergene, perm = perm_res,
                pairwise = NULL, cld_all = NULL,
                concordance = NULL, planned = NULL))
  }

  pairwise_all <- do.call(rbind, pw_list)
  cld_all      <- do.call(rbind, cld_list)

  write.csv(pairwise_all, file.path("isolate_anova_results",
                                     paste0(out_prefix, "_pairwise_isolate.csv")),
            row.names = FALSE)
  write.csv(cld_all, file.path("isolate_anova_results",
                                paste0(out_prefix, "_CLD_isolate.csv")),
            row.names = FALSE)
  cat(sprintf("  Saved %d pairwise comparisons for %d genes\n",
              nrow(pairwise_all), length(pw_list)))

  # 5. CLD concordance
  cat("\n-- CLD Concordance with Hypothesis --\n")

  check_same_group <- function(cld_g, iso1, iso2) {
    g1 <- trimws(cld_g$.group[cld_g$isolate == iso1])
    g2 <- trimws(cld_g$.group[cld_g$isolate == iso2])
    if (length(g1) == 0 || length(g2) == 0) return(NA)
    length(intersect(strsplit(g1,"")[[1]], strsplit(g2,"")[[1]])) > 0
  }

  tested_genes <- names(cld_list)
  concordance <- do.call(rbind, lapply(tested_genes, function(g) {
    cg <- cld_all[cld_all$gene == g, ]
    data.frame(gene = g,
               low_same         = check_same_group(cg, "WISS47", "JS659"),
               high_same        = check_same_group(cg, "MNSS6",  "Xtra7"),
               low_vs_high_diff = !check_same_group(cg, "WISS47", "MNSS6"),
               stringsAsFactors = FALSE)
  }))

  nt <- sum(!is.na(concordance$low_same))
  if (nt > 0) {
    cat(sprintf("  WISS47 ~ JS659 (Low):     %d / %d (%.1f%%)\n",
                sum(concordance$low_same, na.rm=T), nt, 100*mean(concordance$low_same, na.rm=T)))
    cat(sprintf("  MNSS6  ~ Xtra7 (High):    %d / %d (%.1f%%)\n",
                sum(concordance$high_same, na.rm=T), nt, 100*mean(concordance$high_same, na.rm=T)))
    cat(sprintf("  Low != High (separated):  %d / %d (%.1f%%)\n",
                sum(concordance$low_vs_high_diff, na.rm=T), nt,
                100*mean(concordance$low_vs_high_diff, na.rm=T)))
  }
  write.csv(concordance, file.path("isolate_anova_results",
                                    paste0(out_prefix, "_CLD_concordance.csv")),
            row.names = FALSE)

  # 6. Planned contrasts
  cat("\n-- Planned contrasts --\n")
  first_cld <- cld_list[[1]]
  emm_order <- as.character(first_cld$isolate)
  cat("  emmeans isolate order:", paste(emm_order, collapse=", "), "\n")

  mk <- function(weights) {
    vec <- rep(0, length(emm_order))
    for (nm in names(weights)) {
      idx <- which(emm_order == nm)
      if (length(idx)==1) vec[idx] <- weights[[nm]]
    }
    vec
  }
  con_defs <- list(
    "High vs Low" = mk(list(WISS47=-0.5, JS659=-0.5, MNSS6=0.5, Xtra7=0.5)),
    "SSPotter vs BN172" = mk(list(BN172=-1, SSPotter=1)),
    "Specialists vs Generalists" = mk(list(WISS47=-0.25, JS659=-0.25,
                                            MNSS6=-0.25, Xtra7=-0.25,
                                            BN172=0.5, SSPotter=0.5))
  )

  planned_list <- list()
  con_errors <- 0
  for (g in tested_genes) {
    ok <- tryCatch({
      df  <- data.frame(expr = expr_mat[g, ], meta)
      gene_formula <- as.formula(paste("expr ~", model_formula_str))
      fit <- lm(gene_formula, data = df)
      emm <- emmeans(fit, ~ isolate)
      con_df <- as.data.frame(summary(contrast(emm, con_defs)))
      con_df$gene <- g
      planned_list[[g]] <<- con_df
      TRUE
    }, error = function(e) { con_errors <<- con_errors + 1; FALSE })
  }
  if (con_errors > 0) cat(sprintf("  %d contrast errors\n", con_errors))

  if (length(planned_list) > 0) {
    planned_all <- do.call(rbind, planned_list)
    planned_all$padj <- p.adjust(planned_all$p.value, method = "BH")
    write.csv(planned_all, file.path("isolate_anova_results",
                                      paste0(out_prefix, "_planned_contrasts.csv")),
              row.names = FALSE)
    for (ct in unique(planned_all$contrast)) {
      sub <- planned_all[planned_all$contrast == ct, ]
      cat(sprintf("  %-30s %d / %d sig\n", ct, sum(sub$padj<0.05, na.rm=T), nrow(sub)))
    }
  } else {
    planned_all <- NULL
    cat("  All contrasts failed.\n")
  }

  list(pergene = pergene, perm = perm_res,
       pairwise = pairwise_all, cld_all = cld_all,
       concordance = concordance, planned = planned_all)
}
```


```{r define-viz-functions}
plot_isolate_heatmap <- function(vst_obj, pergene, title_str,
                                  n_top = 200, cluster_cols = TRUE) {
  if (is.null(pergene) || nrow(pergene) == 0) {
    cat("  No results for heatmap.\n"); return(invisible(NULL))
  }
  sig <- pergene %>% filter(padj_isolate < 0.05) %>% arrange(padj_isolate) %>% head(n_top)
  genes <- intersect(sig$gene, rownames(assay(vst_obj)))
  if (length(genes) < 5) {
    cat("  <5 significant genes -- skipping heatmap.\n"); return(invisible(NULL))
  }
  mat <- t(scale(t(assay(vst_obj)[genes, ])))
  ann <- data.frame(Timepoint  = vst_obj$timepoint,
                    Isolate    = vst_obj$isolate,
                    Aggr_group = vst_obj$aggr_group,
                    row.names  = colnames(mat))
  ann_colors <- list(
    Timepoint  = tp_colors[levels(droplevels(ann$Timepoint))],
    Isolate    = isolate_colors[levels(droplevels(ann$Isolate))],
    Aggr_group = aggr_colors[levels(droplevels(ann$Aggr_group))])
  iso_order <- c("WISS47","JS659","MNSS6","Xtra7","BN172","SSPotter")
  co <- order(match(ann$Isolate, iso_order), match(ann$Timepoint, c("24h","48h","96h")))
  pheatmap(mat[, co], annotation_col = ann[co,,drop=FALSE],
           annotation_colors = ann_colors, show_rownames = FALSE,
           show_colnames = TRUE, cluster_rows = TRUE, cluster_cols = cluster_cols,
           color = colorRampPalette(c("blue","white","red"))(100),
           breaks = seq(-2,2,length.out=101), main = title_str,
           fontsize_row = 5, fontsize_col = 7)
}

plot_sig_bar <- function(pergene, title_str) {
  if (is.null(pergene) || nrow(pergene) == 0) return(ggplot() + theme_void())
  df <- data.frame(
    Factor = factor(c("Isolate","Timepoint","Block"), levels=c("Isolate","Timepoint","Block")),
    n_sig  = c(sum(pergene$padj_isolate<0.05,na.rm=T),
               sum(pergene$padj_timepoint<0.05,na.rm=T),
               sum(pergene$padj_block<0.05,na.rm=T)))
  ggplot(df, aes(x=Factor, y=n_sig, fill=Factor)) +
    geom_col(alpha=0.85) + geom_text(aes(label=n_sig), vjust=-0.5, size=5) +
    scale_fill_manual(values=c("#1f77b4","#FFD700","#999999")) +
    labs(title=title_str, y="Sig genes (padj<0.05)", x="") +
    theme_bw(base_size=14) + theme(legend.position="none")
}

plot_pval_histograms <- function(pergene, title_str) {
  if (is.null(pergene) || nrow(pergene)==0) return(invisible(NULL))
  par(mfrow=c(1,3))
  hist(pergene$p_isolate,   breaks=50, col="#1f77b4", main=paste(title_str,"Isolate"),   xlab="p")
  hist(pergene$p_timepoint, breaks=50, col="#FFD700",  main=paste(title_str,"Timepoint"), xlab="p")
  hist(pergene$p_block,     breaks=50, col="#999999",  main=paste(title_str,"Block"),     xlab="p")
  par(mfrow=c(1,1))
}

plot_planned_volcano <- function(planned_df, contrast_name, title_str) {
  if (is.null(planned_df) || nrow(planned_df)==0)
    return(ggplot() + labs(title=paste(title_str,"no data")) + theme_void())
  sub <- planned_df[planned_df$contrast==contrast_name,]
  if (nrow(sub)==0) return(ggplot()+theme_void())
  sub$sig <- ifelse(sub$padj<0.05,"Sig","NS")
  ggplot(sub, aes(x=estimate, y=-log10(padj), color=sig)) +
    geom_point(alpha=0.6, size=2) +
    scale_color_manual(values=c(Sig="red",NS="grey60")) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed") +
    labs(title=paste(title_str,contrast_name), x="Estimate", y="-log10(padj)") +
    theme_bw(base_size=13) + theme(legend.position="none")
}
```


# A. Pathogen Genes on Soybean

```{r anova-path-soy}
res_path_soy <- run_isolate_anova(vst_soy_pathogen, "Pathogen - Soybean", "pathogen_soybean")
```

```{r viz-path-soy-bar, fig.height=4}
print(plot_sig_bar(res_path_soy$pergene, "Pathogen on Soybean"))
```

```{r viz-path-soy-pval, fig.height=4}
plot_pval_histograms(res_path_soy$pergene, "Path-Soy")
```

```{r viz-path-soy-heatmap, fig.height=10}
plot_isolate_heatmap(vst_soy_pathogen, res_path_soy$pergene,
                     "Pathogen on Soybean - Isolate DEGs (clustered)", n_top=200, cluster_cols=TRUE)
```

```{r viz-path-soy-heatmap2, fig.height=10}
plot_isolate_heatmap(vst_soy_pathogen, res_path_soy$pergene,
                     "Pathogen on Soybean - Ordered by Isolate", n_top=200, cluster_cols=FALSE)
```

```{r viz-path-soy-con, fig.height=5, fig.width=14}
if (!is.null(res_path_soy$planned)) {
  grid.arrange(
    plot_planned_volcano(res_path_soy$planned, "High vs Low", "Path-Soy"),
    plot_planned_volcano(res_path_soy$planned, "SSPotter vs BN172", "Path-Soy"),
    plot_planned_volcano(res_path_soy$planned, "Specialists vs Generalists", "Path-Soy"),
    nrow=1)
}
```


# B. Pathogen Genes on Sunflower

```{r anova-path-sun}
res_path_sun <- run_isolate_anova(vst_sun_pathogen, "Pathogen - Sunflower", "pathogen_sunflower")
```

```{r viz-path-sun-bar, fig.height=4}
print(plot_sig_bar(res_path_sun$pergene, "Pathogen on Sunflower"))
```

```{r viz-path-sun-pval, fig.height=4}
plot_pval_histograms(res_path_sun$pergene, "Path-Sun")
```

```{r viz-path-sun-heatmap, fig.height=10}
plot_isolate_heatmap(vst_sun_pathogen, res_path_sun$pergene,
                     "Pathogen on Sunflower - Isolate DEGs (clustered)", n_top=200, cluster_cols=TRUE)
```

```{r viz-path-sun-heatmap2, fig.height=10}
plot_isolate_heatmap(vst_sun_pathogen, res_path_sun$pergene,
                     "Pathogen on Sunflower - Ordered by Isolate", n_top=200, cluster_cols=FALSE)
```

```{r viz-path-sun-con, fig.height=5, fig.width=14}
if (!is.null(res_path_sun$planned)) {
  grid.arrange(
    plot_planned_volcano(res_path_sun$planned, "High vs Low", "Path-Sun"),
    plot_planned_volcano(res_path_sun$planned, "SSPotter vs BN172", "Path-Sun"),
    plot_planned_volcano(res_path_sun$planned, "Specialists vs Generalists", "Path-Sun"),
    nrow=1)
}
```


# C. Soybean Host Genes

```{r anova-host-soy}
res_host_soy <- run_isolate_anova(vst_soy_host, "Soybean host genes", "host_soybean")
```

```{r viz-host-soy-bar, fig.height=4}
print(plot_sig_bar(res_host_soy$pergene, "Soybean Host"))
```

```{r viz-host-soy-pval, fig.height=4}
plot_pval_histograms(res_host_soy$pergene, "Host-Soy")
```

```{r viz-host-soy-heatmap, fig.height=10}
plot_isolate_heatmap(vst_soy_host, res_host_soy$pergene,
                     "Soybean Host - Isolate DEGs (clustered)", n_top=200, cluster_cols=TRUE)
```

```{r viz-host-soy-heatmap2, fig.height=10}
plot_isolate_heatmap(vst_soy_host, res_host_soy$pergene,
                     "Soybean Host - Ordered by Isolate", n_top=200, cluster_cols=FALSE)
```

```{r viz-host-soy-con, fig.height=5, fig.width=14}
if (!is.null(res_host_soy$planned)) {
  grid.arrange(
    plot_planned_volcano(res_host_soy$planned, "High vs Low", "Host-Soy"),
    plot_planned_volcano(res_host_soy$planned, "SSPotter vs BN172", "Host-Soy"),
    plot_planned_volcano(res_host_soy$planned, "Specialists vs Generalists", "Host-Soy"),
    nrow=1)
}
```


# D. Sunflower Host Genes

```{r anova-host-sun}
res_host_sun <- run_isolate_anova(vst_sun_host, "Sunflower host genes", "host_sunflower")
```

```{r viz-host-sun-bar, fig.height=4}
print(plot_sig_bar(res_host_sun$pergene, "Sunflower Host"))
```

```{r viz-host-sun-pval, fig.height=4}
plot_pval_histograms(res_host_sun$pergene, "Host-Sun")
```

```{r viz-host-sun-heatmap, fig.height=10}
plot_isolate_heatmap(vst_sun_host, res_host_sun$pergene,
                     "Sunflower Host - Isolate DEGs (clustered)", n_top=200, cluster_cols=TRUE)
```

```{r viz-host-sun-heatmap2, fig.height=10}
plot_isolate_heatmap(vst_sun_host, res_host_sun$pergene,
                     "Sunflower Host - Ordered by Isolate", n_top=200, cluster_cols=FALSE)
```

```{r viz-host-sun-con, fig.height=5, fig.width=14}
if (!is.null(res_host_sun$planned)) {
  grid.arrange(
    plot_planned_volcano(res_host_sun$planned, "High vs Low", "Host-Sun"),
    plot_planned_volcano(res_host_sun$planned, "SSPotter vs BN172", "Host-Sun"),
    plot_planned_volcano(res_host_sun$planned, "Specialists vs Generalists", "Host-Sun"),
    nrow=1)
}
```


# Grand Summary

```{r grand-summary, fig.width=14, fig.height=6}
cat(paste(rep("=",80), collapse=""), "\n")
cat("HYPOTHESIS CONFIRMATION SUMMARY\n")
cat(paste(rep("=",80), collapse=""), "\n\n")

datasets <- list("Pathogen-Soybean"   = res_path_soy,
                 "Pathogen-Sunflower" = res_path_sun,
                 "Host-Soybean"       = res_host_soy,
                 "Host-Sunflower"     = res_host_sun)
summary_rows <- list()

for (nm in names(datasets)) {
  res <- datasets[[nm]]; pg <- res$pergene; conc <- res$concordance; plan <- res$planned
  if (is.null(pg) || nrow(pg)==0) { cat(sprintf("-- %s -- SKIPPED\n\n",nm)); next }

  n_iso <- sum(pg$padj_isolate<0.05, na.rm=T)
  n_blk <- sum(pg$padj_block<0.05, na.rm=T)
  nt    <- nrow(pg)

  cat(sprintf("-- %s --\n", nm))
  cat(sprintf("  Isolate DEGs: %d/%d (%.1f%%)\n", n_iso, nt, 100*n_iso/nt))
  cat(sprintf("  Block DEGs:   %d/%d (%.1f%%)\n", n_blk, nt, 100*n_blk/nt))

  if (!is.null(conc) && nrow(conc)>0 && sum(!is.na(conc$low_same))>0) {
    pl <- 100*mean(conc$low_same,na.rm=T); ph <- 100*mean(conc$high_same,na.rm=T)
    pd <- 100*mean(conc$low_vs_high_diff,na.rm=T)
    cat(sprintf("  WISS47~JS659 (Low):  %.1f%%\n", pl))
    cat(sprintf("  MNSS6~Xtra7 (High):  %.1f%%\n", ph))
    cat(sprintf("  Low!=High:           %.1f%%\n", pd))
    summary_rows[[nm]] <- data.frame(Dataset=nm, Isolate_DEGs=n_iso, Block_DEGs=n_blk,
                                      Pct_Low_same=round(pl,1), Pct_High_same=round(ph,1),
                                      Pct_LowVsHigh_diff=round(pd,1))
  }

  if (!is.null(plan) && nrow(plan)>0) {
    cat("  Planned contrasts (padj<0.05):\n")
    for (ct in unique(plan$contrast)) {
      sub <- plan[plan$contrast==ct,]
      cat(sprintf("    %-30s %d/%d\n", ct, sum(sub$padj<0.05,na.rm=T), nrow(sub)))
    }
  }
  cat("\n")
}

if (length(summary_rows)>0) {
  sdf <- do.call(rbind, summary_rows)
  cat("\n-- Concordance Summary --\n"); print(sdf, row.names=FALSE)
  write.csv(sdf, "isolate_anova_results/hypothesis_concordance_summary.csv", row.names=FALSE)
}

# Combined bar
bar_data <- do.call(rbind, lapply(names(datasets), function(nm) {
  pg <- datasets[[nm]]$pergene
  if (is.null(pg)||nrow(pg)==0) return(NULL)
  data.frame(Dataset=nm,
             Factor=c("Isolate","Timepoint","Block"),
             n_sig=c(sum(pg$padj_isolate<0.05,na.rm=T),
                     sum(pg$padj_timepoint<0.05,na.rm=T),
                     sum(pg$padj_block<0.05,na.rm=T)))
}))
if (!is.null(bar_data) && nrow(bar_data)>0) {
  bar_data$Factor  <- factor(bar_data$Factor, levels=c("Isolate","Timepoint","Block"))
  bar_data$Dataset <- factor(bar_data$Dataset, levels=names(datasets))
  print(ggplot(bar_data, aes(x=Factor, y=n_sig, fill=Dataset)) +
    geom_col(position=position_dodge(0.8), alpha=0.85, width=0.7) +
    geom_text(aes(label=n_sig), position=position_dodge(0.8), vjust=-0.4, size=3) +
    scale_fill_manual(values=c("#2ca02c","#17becf","#ffcb21","#5e3a04")) +
    labs(title="Significant Genes per Factor", y="Sig genes (padj<0.05)", x="") +
    theme_bw(base_size=14) + theme(legend.position="top"))
}
```

# Output File Index

```{r file-index}
cat("All files in isolate_anova_results/:\n\n")
cat("Per dataset (x4):\n")
cat("  *_PERMANOVA.csv         - whole-transcriptome variance\n")
cat("  *_pergene_ANOVA.csv     - per-gene F & p\n")
cat("  *_pairwise_isolate.csv  - Tukey pairwise\n")
cat("  *_CLD_isolate.csv       - compact letter display\n")
cat("  *_CLD_concordance.csv   - hypothesis match check\n")
cat("  *_planned_contrasts.csv - High vs Low, etc.\n\n")
cat("Summary:\n")
cat("  hypothesis_concordance_summary.csv\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
