---
title: "ANOVA & Pairwise Comparisons — Gene Expression"
subtitle: "PERMANOVA (whole-transcriptome) + per-gene ANOVA (FDR-corrected)"
author: "Analysis by wan00965"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
```

# Overview

This notebook runs two complementary ANOVA approaches on the RNA-seq
expression data:

| Approach | Scope | What it tells you |
|----------|-------|-------------------|
| **PERMANOVA** (vegan::adonis2) | Whole-transcriptome distance matrix | How much of the *overall* transcriptional variance each factor explains (fills the ANOVA table with one row per factor) |
| **Per-gene ANOVA** | Each gene independently, FDR-corrected | Which *individual genes* are significantly affected by each factor |

**Three analyses are run:**

1. **Pathogen (Ss) genes across both hosts** — factors: Host, Isolate, Timepoint, Block
2. **Soybean host (Gm) genes** — factors: Isolate, Timepoint, Block
3. **Sunflower host (Ha) genes** — factors: Isolate, Timepoint, Block

> **Prerequisite:** Run both individual species notebooks first so that
> count matrices and metadata exist, OR run the data-loading chunks from
> `Overall_cross_species_20260326.Rmd`.

# Setup

```{r load-libraries}
library(DESeq2)
library(vegan)       # PERMANOVA
library(car)         # Type III ANOVA
library(emmeans)     # pairwise comparisons
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(RColorBrewer)
library(gridExtra)

cat("All libraries loaded.\n")
```

```{r set-directory}
setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_data/New_salmon_round3/new_run_count")
dir.create("anova_results", showWarnings = FALSE)
```

# Load and Prepare Data

## Load count matrices

```{r load-counts}
# ── Soybean ──
counts_soy_host     <- read.table("gene_counts_soybean_SALMON_HOST_only.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)
counts_soy_pathogen <- read.table("gene_counts_soybean_SALMON_PATHOGEN_only.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)

# ── Sunflower ──
counts_sun_host     <- read.table("gene_counts_sunflower_SALMON_HOST_only.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)
counts_sun_pathogen <- read.table("gene_counts_sunflower_SALMON_PATHOGEN_only.txt",
                                   header = TRUE, row.names = 1, check.names = FALSE)

cat(sprintf("Soybean:   %d host genes,  %d pathogen genes, %d samples\n",
            nrow(counts_soy_host), nrow(counts_soy_pathogen), ncol(counts_soy_host)))
cat(sprintf("Sunflower: %d host genes,  %d pathogen genes, %d samples\n",
            nrow(counts_sun_host), nrow(counts_sun_pathogen), ncol(counts_sun_host)))
```

## Build metadata

```{r build-metadata}
# ── Helper function ──
build_metadata <- function(sample_names, host_label) {
  data.frame(
    sample    = sample_names,
    host      = host_label,
    timepoint = sapply(sample_names, function(x) {
      parts <- strsplit(x, "_")[[1]]
      if (parts[length(parts)] == "NC") "NC" else paste0(parts[1], "h")
    }),
    block     = sapply(strsplit(sample_names, "_"), `[`, 2),
    isolate   = sapply(strsplit(sample_names, "_"), `[`, 4),
    row.names = sample_names,
    stringsAsFactors = FALSE
  )
}

meta_soy <- build_metadata(colnames(counts_soy_host), "Soybean")
meta_sun <- build_metadata(colnames(counts_sun_host), "Sunflower")

# Combined metadata (for pathogen analysis across hosts)
meta_all <- rbind(meta_soy, meta_sun)
meta_all$host      <- factor(meta_all$host)
meta_all$timepoint <- factor(meta_all$timepoint, levels = c("NC","24h","48h","96h"))
meta_all$block     <- factor(meta_all$block)
meta_all$isolate   <- factor(meta_all$isolate,
                             levels = c("NC","WISS47","JS659","MNSS6","Xtra7","BN172","SSPotter"))

cat("Combined sample counts:\n")
print(table(meta_all$host, meta_all$timepoint))
```

## Prepare DESeq2 objects and VST

```{r prepare-deseq-objects}
# ── Helper: build DESeq2 + VST, excluding NC, returning filtered VST matrix + metadata ──
prepare_vst <- function(counts, metadata, design_formula, min_count = 10, min_samples = 3) {
  counts <- round(counts)
  keep   <- rowSums(counts >= min_count) >= min_samples
  counts <- counts[keep, ]
  
  dds <- DESeqDataSetFromMatrix(countData = counts,
                                colData   = metadata,
                                design    = design_formula)
  dds <- DESeq(dds)
  vst_obj <- vst(dds, blind = FALSE)
  
  # Exclude NC samples
  idx <- vst_obj$timepoint != "NC"
  vst_obj <- vst_obj[, idx]
  colData(vst_obj) <- droplevels(colData(vst_obj))
  
  return(vst_obj)
}

# ── 1. Pathogen genes: combine across hosts ──
common_pathogen_genes <- intersect(rownames(counts_soy_pathogen),
                                   rownames(counts_sun_pathogen))
counts_pathogen_combined <- cbind(counts_soy_pathogen[common_pathogen_genes, ],
                                  counts_sun_pathogen[common_pathogen_genes, ])
cat(sprintf("Common pathogen genes: %d\n", length(common_pathogen_genes)))

vst_pathogen <- prepare_vst(counts_pathogen_combined, meta_all, ~ timepoint + host)

# ── 2. Soybean host genes ──
vst_soy_host <- prepare_vst(counts_soy_host, meta_soy, ~ timepoint)

# ── 3. Sunflower host genes ──
vst_sun_host <- prepare_vst(counts_sun_host, meta_sun, ~ timepoint)

cat(sprintf("\nAfter filtering & excluding NC:\n"))
cat(sprintf("  Pathogen genes (combined): %d genes × %d samples\n",
            nrow(vst_pathogen), ncol(vst_pathogen)))
cat(sprintf("  Soybean host genes:        %d genes × %d samples\n",
            nrow(vst_soy_host), ncol(vst_soy_host)))
cat(sprintf("  Sunflower host genes:      %d genes × %d samples\n",
            nrow(vst_sun_host), ncol(vst_sun_host)))
```


# ═══════════════════════════════════════════════════════════════════════════
# PART A: Pathogen Genes Across Both Hosts
# ═══════════════════════════════════════════════════════════════════════════

# A1. PERMANOVA — Pathogen Genes

```{r permanova-pathogen}
# Extract VST expression matrix (genes × samples → transpose to samples × genes)
expr_pathogen <- t(assay(vst_pathogen))

# Build sample-level factor table
perm_meta_pathogen <- data.frame(
  host      = vst_pathogen$host,
  isolate   = vst_pathogen$isolate,
  timepoint = vst_pathogen$timepoint,
  block     = vst_pathogen$block,
  row.names = colnames(vst_pathogen)
)

# Drop unused factor levels
perm_meta_pathogen <- droplevels(perm_meta_pathogen)

# Euclidean distance on VST values
dist_pathogen <- vegdist(expr_pathogen, method = "euclidean")

# PERMANOVA — sequential (Type I) with 999 permutations
set.seed(42)
perm_pathogen <- adonis2(dist_pathogen ~ host + isolate + timepoint + block,
                         data       = perm_meta_pathogen,
                         permutations = 999,
                         method     = "euclidean",
                         by         = "margin")  # marginal (Type III-like)

cat("══════ PERMANOVA — Pathogen Genes (marginal) ══════\n")
print(perm_pathogen)
```

```{r permanova-pathogen-table}
# ── Build publication table from PERMANOVA ──
perm_df <- as.data.frame(perm_pathogen)

# Calculate eta-squared (R2 is already provided by adonis2)
pub_perm_pathogen <- data.frame(
  Source   = c("Host", "S. sclerotiorum isolate", "Inoculated timepoint",
               "Block (biological replicates)", "Residual", "Total"),
  df       = perm_df$Df,
  SS       = round(perm_df$SumOfSqs, 2),
  MS       = round(ifelse(perm_df$Df > 0, perm_df$SumOfSqs / perm_df$Df, NA), 2),
  F_value  = round(perm_df$F, 2),
  p_value  = perm_df$`Pr(>F)`,
  R2       = round(perm_df$R2, 4)
)

cat("\n══════ Publication-Ready PERMANOVA Table — Pathogen Genes ══════\n")
print(pub_perm_pathogen, row.names = FALSE, na.print = "")

write.csv(pub_perm_pathogen,
          "anova_results/PERMANOVA_pathogen_genes.csv", row.names = FALSE)
```

## A2. Per-Gene ANOVA — Pathogen Genes

```{r pergene-anova-pathogen}
# VST expression matrix: genes in rows, samples in columns
expr_mat <- assay(vst_pathogen)

# Factor table (same as PERMANOVA)
gene_meta <- perm_meta_pathogen

# Sum-to-zero contrasts for Type III
options(contrasts = c("contr.sum", "contr.poly"))

cat("Running per-gene ANOVA for", nrow(expr_mat), "pathogen genes...\n")

# ── Run ANOVA on each gene ──
pergene_results_pathogen <- do.call(rbind, lapply(seq_len(nrow(expr_mat)), function(i) {
  df_gene <- data.frame(expr = expr_mat[i, ], gene_meta)
  
  fit <- tryCatch(
    lm(expr ~ host + isolate + timepoint + block, data = df_gene),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    return(data.frame(gene = rownames(expr_mat)[i],
                      F_host = NA, p_host = NA,
                      F_isolate = NA, p_isolate = NA,
                      F_timepoint = NA, p_timepoint = NA,
                      F_block = NA, p_block = NA))
  }
  
  av <- tryCatch(Anova(fit, type = "III"), error = function(e) NULL)
  if (is.null(av)) {
    return(data.frame(gene = rownames(expr_mat)[i],
                      F_host = NA, p_host = NA,
                      F_isolate = NA, p_isolate = NA,
                      F_timepoint = NA, p_timepoint = NA,
                      F_block = NA, p_block = NA))
  }
  
  data.frame(
    gene        = rownames(expr_mat)[i],
    F_host      = av["host",      "F value"],
    p_host      = av["host",      "Pr(>F)"],
    F_isolate   = av["isolate",   "F value"],
    p_isolate   = av["isolate",   "Pr(>F)"],
    F_timepoint = av["timepoint", "F value"],
    p_timepoint = av["timepoint", "Pr(>F)"],
    F_block     = av["block",     "F value"],
    p_block     = av["block",     "Pr(>F)"]
  )
}))

# ── FDR correction (BH) per factor ──
pergene_results_pathogen$padj_host      <- p.adjust(pergene_results_pathogen$p_host,      method = "BH")
pergene_results_pathogen$padj_isolate   <- p.adjust(pergene_results_pathogen$p_isolate,   method = "BH")
pergene_results_pathogen$padj_timepoint <- p.adjust(pergene_results_pathogen$p_timepoint, method = "BH")
pergene_results_pathogen$padj_block     <- p.adjust(pergene_results_pathogen$p_block,     method = "BH")

cat("\n══════ Per-Gene ANOVA Summary — Pathogen Genes (padj < 0.05) ══════\n")
cat(sprintf("  Host effect:      %d genes\n",
            sum(pergene_results_pathogen$padj_host < 0.05, na.rm = TRUE)))
cat(sprintf("  Isolate effect:   %d genes\n",
            sum(pergene_results_pathogen$padj_isolate < 0.05, na.rm = TRUE)))
cat(sprintf("  Timepoint effect: %d genes\n",
            sum(pergene_results_pathogen$padj_timepoint < 0.05, na.rm = TRUE)))
cat(sprintf("  Block effect:     %d genes\n",
            sum(pergene_results_pathogen$padj_block < 0.05, na.rm = TRUE)))

write.csv(pergene_results_pathogen,
          "anova_results/pergene_ANOVA_pathogen_all.csv", row.names = FALSE)

# ── Export significant gene lists per factor ──
for (fac in c("host", "isolate", "timepoint")) {
  col_padj <- paste0("padj_", fac)
  sig <- pergene_results_pathogen %>%
    filter(.data[[col_padj]] < 0.05) %>%
    arrange(.data[[col_padj]])
  write.csv(sig,
            sprintf("anova_results/pergene_ANOVA_pathogen_sig_%s.csv", fac),
            row.names = FALSE)
  cat(sprintf("  Saved %d significant genes for %s effect\n", nrow(sig), fac))
}
```

## A3. Per-Gene Pairwise Comparisons — Pathogen Genes

For factors with >2 levels (isolate, timepoint), we use `emmeans` to run
Tukey-adjusted pairwise comparisons on genes that are significant for that
factor.

```{r pairwise-pathogen-isolate}
# ── Isolate pairwise (on significant genes only) ──
sig_isolate_genes <- pergene_results_pathogen$gene[
  which(pergene_results_pathogen$padj_isolate < 0.05)
]

cat(sprintf("\nRunning isolate pairwise comparisons for %d significant pathogen genes...\n",
            length(sig_isolate_genes)))

if (length(sig_isolate_genes) > 0) {
  # Limit to top 500 if very many (for speed)
  genes_to_test <- head(sig_isolate_genes, 500)
  
  pairwise_isolate_list <- lapply(genes_to_test, function(g) {
    df_gene <- data.frame(expr = expr_mat[g, ], gene_meta)
    fit <- lm(expr ~ host + isolate + timepoint + block, data = df_gene)
    emm <- emmeans(fit, pairwise ~ isolate, adjust = "tukey")
    contr <- as.data.frame(summary(emm$contrasts))
    contr$gene <- g
    return(contr)
  })
  
  pairwise_isolate_pathogen <- do.call(rbind, pairwise_isolate_list)
  write.csv(pairwise_isolate_pathogen,
            "anova_results/pairwise_isolate_pathogen.csv", row.names = FALSE)
  
  cat(sprintf("  Saved %d pairwise comparisons across %d genes\n",
              nrow(pairwise_isolate_pathogen), length(genes_to_test)))
}
```

```{r pairwise-pathogen-timepoint}
# ── Timepoint pairwise (on significant genes only) ──
sig_timepoint_genes <- pergene_results_pathogen$gene[
  which(pergene_results_pathogen$padj_timepoint < 0.05)
]

cat(sprintf("\nRunning timepoint pairwise comparisons for %d significant pathogen genes...\n",
            length(sig_timepoint_genes)))

if (length(sig_timepoint_genes) > 0) {
  genes_to_test <- head(sig_timepoint_genes, 500)
  
  pairwise_tp_list <- lapply(genes_to_test, function(g) {
    df_gene <- data.frame(expr = expr_mat[g, ], gene_meta)
    fit <- lm(expr ~ host + isolate + timepoint + block, data = df_gene)
    emm <- emmeans(fit, pairwise ~ timepoint, adjust = "tukey")
    contr <- as.data.frame(summary(emm$contrasts))
    contr$gene <- g
    return(contr)
  })
  
  pairwise_tp_pathogen <- do.call(rbind, pairwise_tp_list)
  write.csv(pairwise_tp_pathogen,
            "anova_results/pairwise_timepoint_pathogen.csv", row.names = FALSE)
  
  cat(sprintf("  Saved %d pairwise comparisons across %d genes\n",
              nrow(pairwise_tp_pathogen), length(genes_to_test)))
}
```

```{r pairwise-pathogen-host}
# ── Host pairwise (just 2 levels, but extract for all significant genes) ──
sig_host_genes <- pergene_results_pathogen$gene[
  which(pergene_results_pathogen$padj_host < 0.05)
]

cat(sprintf("\nRunning host pairwise comparisons for %d significant pathogen genes...\n",
            length(sig_host_genes)))

if (length(sig_host_genes) > 0) {
  genes_to_test <- head(sig_host_genes, 500)
  
  pairwise_host_list <- lapply(genes_to_test, function(g) {
    df_gene <- data.frame(expr = expr_mat[g, ], gene_meta)
    fit <- lm(expr ~ host + isolate + timepoint + block, data = df_gene)
    emm <- emmeans(fit, pairwise ~ host, adjust = "tukey")
    contr <- as.data.frame(summary(emm$contrasts))
    contr$gene <- g
    # Also get the means
    means <- as.data.frame(summary(emm$emmeans))
    contr$mean_soybean   <- means$emmean[means$host == "Soybean"]
    contr$mean_sunflower <- means$emmean[means$host == "Sunflower"]
    return(contr)
  })
  
  pairwise_host_pathogen <- do.call(rbind, pairwise_host_list)
  write.csv(pairwise_host_pathogen,
            "anova_results/pairwise_host_pathogen.csv", row.names = FALSE)
  
  cat(sprintf("  Saved %d pairwise comparisons\n", nrow(pairwise_host_pathogen)))
}
```

## A4. Summary Visualizations — Pathogen

```{r viz-pathogen-venn, fig.height=6}
# Bar chart of significant gene counts per factor
sig_counts_pathogen <- data.frame(
  Factor = c("Host", "Isolate", "Timepoint", "Block"),
  n_sig  = c(
    sum(pergene_results_pathogen$padj_host      < 0.05, na.rm = TRUE),
    sum(pergene_results_pathogen$padj_isolate   < 0.05, na.rm = TRUE),
    sum(pergene_results_pathogen$padj_timepoint < 0.05, na.rm = TRUE),
    sum(pergene_results_pathogen$padj_block     < 0.05, na.rm = TRUE)
  )
)
sig_counts_pathogen$Factor <- factor(sig_counts_pathogen$Factor,
                                     levels = c("Host","Isolate","Timepoint","Block"))

ggplot(sig_counts_pathogen, aes(x = Factor, y = n_sig, fill = Factor)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = n_sig), vjust = -0.5, size = 5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pathogen Genes: Significant Genes per Factor (padj < 0.05)",
       subtitle = "Per-gene Type III ANOVA, BH-corrected",
       y = "Number of significant genes", x = "") +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```

```{r viz-pathogen-pvalue-hist, fig.height=8}
# P-value histograms (diagnostic — should be uniform under null + spike at 0)
par(mfrow = c(2, 2))
hist(pergene_results_pathogen$p_host,      breaks = 50, main = "Host",      xlab = "p-value", col = "lightblue")
hist(pergene_results_pathogen$p_isolate,   breaks = 50, main = "Isolate",   xlab = "p-value", col = "lightgreen")
hist(pergene_results_pathogen$p_timepoint, breaks = 50, main = "Timepoint", xlab = "p-value", col = "lightyellow")
hist(pergene_results_pathogen$p_block,     breaks = 50, main = "Block",     xlab = "p-value", col = "lightgrey")
par(mfrow = c(1, 1))
```


# ═══════════════════════════════════════════════════════════════════════════
# PART B: Soybean Host Genes
# ═══════════════════════════════════════════════════════════════════════════

# B1. PERMANOVA — Soybean Host Genes

```{r permanova-soybean-host}
expr_soy_host <- t(assay(vst_soy_host))

perm_meta_soy <- data.frame(
  isolate   = vst_soy_host$isolate,
  timepoint = vst_soy_host$timepoint,
  block     = vst_soy_host$block,
  row.names = colnames(vst_soy_host)
)
perm_meta_soy <- droplevels(perm_meta_soy)

dist_soy_host <- vegdist(expr_soy_host, method = "euclidean")

set.seed(42)
perm_soy_host <- adonis2(dist_soy_host ~ isolate + timepoint + block,
                          data = perm_meta_soy,
                          permutations = 999,
                          method = "euclidean",
                          by = "margin")

cat("══════ PERMANOVA — Soybean Host Genes (marginal) ══════\n")
print(perm_soy_host)

# Publication table
perm_df_soy <- as.data.frame(perm_soy_host)
pub_perm_soy <- data.frame(
  Source   = c("S. sclerotiorum isolate", "Inoculated timepoint",
               "Block (biological replicates)", "Residual", "Total"),
  df       = perm_df_soy$Df,
  SS       = round(perm_df_soy$SumOfSqs, 2),
  MS       = round(ifelse(perm_df_soy$Df > 0, perm_df_soy$SumOfSqs / perm_df_soy$Df, NA), 2),
  F_value  = round(perm_df_soy$F, 2),
  p_value  = perm_df_soy$`Pr(>F)`,
  R2       = round(perm_df_soy$R2, 4)
)
print(pub_perm_soy, row.names = FALSE, na.print = "")
write.csv(pub_perm_soy,
          "anova_results/PERMANOVA_soybean_host_genes.csv", row.names = FALSE)
```

## B2. Per-Gene ANOVA — Soybean Host Genes

```{r pergene-anova-soybean-host}
expr_mat_soy <- assay(vst_soy_host)
gene_meta_soy <- perm_meta_soy

options(contrasts = c("contr.sum", "contr.poly"))

cat("Running per-gene ANOVA for", nrow(expr_mat_soy), "soybean host genes...\n")

pergene_results_soy <- do.call(rbind, lapply(seq_len(nrow(expr_mat_soy)), function(i) {
  df_gene <- data.frame(expr = expr_mat_soy[i, ], gene_meta_soy)
  
  fit <- tryCatch(
    lm(expr ~ isolate + timepoint + block, data = df_gene),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    return(data.frame(gene = rownames(expr_mat_soy)[i],
                      F_isolate = NA, p_isolate = NA,
                      F_timepoint = NA, p_timepoint = NA,
                      F_block = NA, p_block = NA))
  }
  
  av <- tryCatch(Anova(fit, type = "III"), error = function(e) NULL)
  if (is.null(av)) {
    return(data.frame(gene = rownames(expr_mat_soy)[i],
                      F_isolate = NA, p_isolate = NA,
                      F_timepoint = NA, p_timepoint = NA,
                      F_block = NA, p_block = NA))
  }
  
  data.frame(
    gene        = rownames(expr_mat_soy)[i],
    F_isolate   = av["isolate",   "F value"],
    p_isolate   = av["isolate",   "Pr(>F)"],
    F_timepoint = av["timepoint", "F value"],
    p_timepoint = av["timepoint", "Pr(>F)"],
    F_block     = av["block",     "F value"],
    p_block     = av["block",     "Pr(>F)"]
  )
}))

# FDR correction
pergene_results_soy$padj_isolate   <- p.adjust(pergene_results_soy$p_isolate,   method = "BH")
pergene_results_soy$padj_timepoint <- p.adjust(pergene_results_soy$p_timepoint, method = "BH")
pergene_results_soy$padj_block     <- p.adjust(pergene_results_soy$p_block,     method = "BH")

cat("\n══════ Per-Gene ANOVA Summary — Soybean Host (padj < 0.05) ══════\n")
cat(sprintf("  Isolate effect:   %d genes\n",
            sum(pergene_results_soy$padj_isolate < 0.05, na.rm = TRUE)))
cat(sprintf("  Timepoint effect: %d genes\n",
            sum(pergene_results_soy$padj_timepoint < 0.05, na.rm = TRUE)))
cat(sprintf("  Block effect:     %d genes\n",
            sum(pergene_results_soy$padj_block < 0.05, na.rm = TRUE)))

write.csv(pergene_results_soy,
          "anova_results/pergene_ANOVA_soybean_host_all.csv", row.names = FALSE)

for (fac in c("isolate", "timepoint")) {
  col_padj <- paste0("padj_", fac)
  sig <- pergene_results_soy %>%
    filter(.data[[col_padj]] < 0.05) %>%
    arrange(.data[[col_padj]])
  write.csv(sig,
            sprintf("anova_results/pergene_ANOVA_soybean_host_sig_%s.csv", fac),
            row.names = FALSE)
}
```

## B3. Pairwise Comparisons — Soybean Host Genes

```{r pairwise-soybean-isolate}
sig_iso_soy <- pergene_results_soy$gene[
  which(pergene_results_soy$padj_isolate < 0.05)
]

cat(sprintf("Running isolate pairwise for %d significant soybean host genes...\n",
            length(sig_iso_soy)))

if (length(sig_iso_soy) > 0) {
  genes_to_test <- head(sig_iso_soy, 500)
  
  pw_list <- lapply(genes_to_test, function(g) {
    df_gene <- data.frame(expr = expr_mat_soy[g, ], gene_meta_soy)
    fit <- lm(expr ~ isolate + timepoint + block, data = df_gene)
    emm <- emmeans(fit, pairwise ~ isolate, adjust = "tukey")
    contr <- as.data.frame(summary(emm$contrasts))
    contr$gene <- g
    return(contr)
  })
  
  pairwise_iso_soy <- do.call(rbind, pw_list)
  write.csv(pairwise_iso_soy,
            "anova_results/pairwise_isolate_soybean_host.csv", row.names = FALSE)
  cat(sprintf("  Saved %d comparisons for %d genes\n",
              nrow(pairwise_iso_soy), length(genes_to_test)))
}
```

```{r pairwise-soybean-timepoint}
sig_tp_soy <- pergene_results_soy$gene[
  which(pergene_results_soy$padj_timepoint < 0.05)
]

cat(sprintf("Running timepoint pairwise for %d significant soybean host genes...\n",
            length(sig_tp_soy)))

if (length(sig_tp_soy) > 0) {
  genes_to_test <- head(sig_tp_soy, 500)
  
  pw_list <- lapply(genes_to_test, function(g) {
    df_gene <- data.frame(expr = expr_mat_soy[g, ], gene_meta_soy)
    fit <- lm(expr ~ isolate + timepoint + block, data = df_gene)
    emm <- emmeans(fit, pairwise ~ timepoint, adjust = "tukey")
    contr <- as.data.frame(summary(emm$contrasts))
    contr$gene <- g
    return(contr)
  })
  
  pairwise_tp_soy <- do.call(rbind, pw_list)
  write.csv(pairwise_tp_soy,
            "anova_results/pairwise_timepoint_soybean_host.csv", row.names = FALSE)
  cat(sprintf("  Saved %d comparisons for %d genes\n",
              nrow(pairwise_tp_soy), length(genes_to_test)))
}
```

## B4. Summary Visualizations — Soybean

```{r viz-soybean-host, fig.height=4}
sig_counts_soy <- data.frame(
  Factor = c("Isolate", "Timepoint", "Block"),
  n_sig  = c(
    sum(pergene_results_soy$padj_isolate   < 0.05, na.rm = TRUE),
    sum(pergene_results_soy$padj_timepoint < 0.05, na.rm = TRUE),
    sum(pergene_results_soy$padj_block     < 0.05, na.rm = TRUE)
  )
)
sig_counts_soy$Factor <- factor(sig_counts_soy$Factor,
                                 levels = c("Isolate","Timepoint","Block"))

ggplot(sig_counts_soy, aes(x = Factor, y = n_sig, fill = Factor)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = n_sig), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("#E41A1C","#FFD700","#999999")) +
  labs(title = "Soybean Host Genes: Significant per Factor (padj < 0.05)",
       y = "Number of significant genes", x = "") +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```


# ═══════════════════════════════════════════════════════════════════════════
# PART C: Sunflower Host Genes
# ═══════════════════════════════════════════════════════════════════════════

# C1. PERMANOVA — Sunflower Host Genes

```{r permanova-sunflower-host}
expr_sun_host <- t(assay(vst_sun_host))

perm_meta_sun <- data.frame(
  isolate   = vst_sun_host$isolate,
  timepoint = vst_sun_host$timepoint,
  block     = vst_sun_host$block,
  row.names = colnames(vst_sun_host)
)
perm_meta_sun <- droplevels(perm_meta_sun)

dist_sun_host <- vegdist(expr_sun_host, method = "euclidean")

set.seed(42)
perm_sun_host <- adonis2(dist_sun_host ~ isolate + timepoint + block,
                          data = perm_meta_sun,
                          permutations = 999,
                          method = "euclidean",
                          by = "margin")

cat("══════ PERMANOVA — Sunflower Host Genes (marginal) ══════\n")
print(perm_sun_host)

perm_df_sun <- as.data.frame(perm_sun_host)
pub_perm_sun <- data.frame(
  Source   = c("S. sclerotiorum isolate", "Inoculated timepoint",
               "Block (biological replicates)", "Residual", "Total"),
  df       = perm_df_sun$Df,
  SS       = round(perm_df_sun$SumOfSqs, 2),
  MS       = round(ifelse(perm_df_sun$Df > 0, perm_df_sun$SumOfSqs / perm_df_sun$Df, NA), 2),
  F_value  = round(perm_df_sun$F, 2),
  p_value  = perm_df_sun$`Pr(>F)`,
  R2       = round(perm_df_sun$R2, 4)
)
print(pub_perm_sun, row.names = FALSE, na.print = "")
write.csv(pub_perm_sun,
          "anova_results/PERMANOVA_sunflower_host_genes.csv", row.names = FALSE)
```

## C2. Per-Gene ANOVA — Sunflower Host Genes

```{r pergene-anova-sunflower-host}
expr_mat_sun <- assay(vst_sun_host)
gene_meta_sun <- perm_meta_sun

options(contrasts = c("contr.sum", "contr.poly"))

cat("Running per-gene ANOVA for", nrow(expr_mat_sun), "sunflower host genes...\n")

pergene_results_sun <- do.call(rbind, lapply(seq_len(nrow(expr_mat_sun)), function(i) {
  df_gene <- data.frame(expr = expr_mat_sun[i, ], gene_meta_sun)
  
  fit <- tryCatch(
    lm(expr ~ isolate + timepoint + block, data = df_gene),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    return(data.frame(gene = rownames(expr_mat_sun)[i],
                      F_isolate = NA, p_isolate = NA,
                      F_timepoint = NA, p_timepoint = NA,
                      F_block = NA, p_block = NA))
  }
  
  av <- tryCatch(Anova(fit, type = "III"), error = function(e) NULL)
  if (is.null(av)) {
    return(data.frame(gene = rownames(expr_mat_sun)[i],
                      F_isolate = NA, p_isolate = NA,
                      F_timepoint = NA, p_timepoint = NA,
                      F_block = NA, p_block = NA))
  }
  
  data.frame(
    gene        = rownames(expr_mat_sun)[i],
    F_isolate   = av["isolate",   "F value"],
    p_isolate   = av["isolate",   "Pr(>F)"],
    F_timepoint = av["timepoint", "F value"],
    p_timepoint = av["timepoint", "Pr(>F)"],
    F_block     = av["block",     "F value"],
    p_block     = av["block",     "Pr(>F)"]
  )
}))

# FDR correction
pergene_results_sun$padj_isolate   <- p.adjust(pergene_results_sun$p_isolate,   method = "BH")
pergene_results_sun$padj_timepoint <- p.adjust(pergene_results_sun$p_timepoint, method = "BH")
pergene_results_sun$padj_block     <- p.adjust(pergene_results_sun$p_block,     method = "BH")

cat("\n══════ Per-Gene ANOVA Summary — Sunflower Host (padj < 0.05) ══════\n")
cat(sprintf("  Isolate effect:   %d genes\n",
            sum(pergene_results_sun$padj_isolate < 0.05, na.rm = TRUE)))
cat(sprintf("  Timepoint effect: %d genes\n",
            sum(pergene_results_sun$padj_timepoint < 0.05, na.rm = TRUE)))
cat(sprintf("  Block effect:     %d genes\n",
            sum(pergene_results_sun$padj_block < 0.05, na.rm = TRUE)))

write.csv(pergene_results_sun,
          "anova_results/pergene_ANOVA_sunflower_host_all.csv", row.names = FALSE)

for (fac in c("isolate", "timepoint")) {
  col_padj <- paste0("padj_", fac)
  sig <- pergene_results_sun %>%
    filter(.data[[col_padj]] < 0.05) %>%
    arrange(.data[[col_padj]])
  write.csv(sig,
            sprintf("anova_results/pergene_ANOVA_sunflower_host_sig_%s.csv", fac),
            row.names = FALSE)
}
```

## C3. Pairwise Comparisons — Sunflower Host Genes

```{r pairwise-sunflower-isolate}
sig_iso_sun <- pergene_results_sun$gene[
  which(pergene_results_sun$padj_isolate < 0.05)
]

cat(sprintf("Running isolate pairwise for %d significant sunflower host genes...\n",
            length(sig_iso_sun)))

if (length(sig_iso_sun) > 0) {
  genes_to_test <- head(sig_iso_sun, 500)
  
  pw_list <- lapply(genes_to_test, function(g) {
    df_gene <- data.frame(expr = expr_mat_sun[g, ], gene_meta_sun)
    fit <- lm(expr ~ isolate + timepoint + block, data = df_gene)
    emm <- emmeans(fit, pairwise ~ isolate, adjust = "tukey")
    contr <- as.data.frame(summary(emm$contrasts))
    contr$gene <- g
    return(contr)
  })
  
  pairwise_iso_sun <- do.call(rbind, pw_list)
  write.csv(pairwise_iso_sun,
            "anova_results/pairwise_isolate_sunflower_host.csv", row.names = FALSE)
  cat(sprintf("  Saved %d comparisons for %d genes\n",
              nrow(pairwise_iso_sun), length(genes_to_test)))
}
```

```{r pairwise-sunflower-timepoint}
sig_tp_sun <- pergene_results_sun$gene[
  which(pergene_results_sun$padj_timepoint < 0.05)
]

cat(sprintf("Running timepoint pairwise for %d significant sunflower host genes...\n",
            length(sig_tp_sun)))

if (length(sig_tp_sun) > 0) {
  genes_to_test <- head(sig_tp_sun, 500)
  
  pw_list <- lapply(genes_to_test, function(g) {
    df_gene <- data.frame(expr = expr_mat_sun[g, ], gene_meta_sun)
    fit <- lm(expr ~ isolate + timepoint + block, data = df_gene)
    emm <- emmeans(fit, pairwise ~ timepoint, adjust = "tukey")
    contr <- as.data.frame(summary(emm$contrasts))
    contr$gene <- g
    return(contr)
  })
  
  pairwise_tp_sun <- do.call(rbind, pw_list)
  write.csv(pairwise_tp_sun,
            "anova_results/pairwise_timepoint_sunflower_host.csv", row.names = FALSE)
  cat(sprintf("  Saved %d comparisons for %d genes\n",
              nrow(pairwise_tp_sun), length(genes_to_test)))
}
```

## C4. Summary Visualizations — Sunflower

```{r viz-sunflower-host, fig.height=4}
sig_counts_sun <- data.frame(
  Factor = c("Isolate", "Timepoint", "Block"),
  n_sig  = c(
    sum(pergene_results_sun$padj_isolate   < 0.05, na.rm = TRUE),
    sum(pergene_results_sun$padj_timepoint < 0.05, na.rm = TRUE),
    sum(pergene_results_sun$padj_block     < 0.05, na.rm = TRUE)
  )
)
sig_counts_sun$Factor <- factor(sig_counts_sun$Factor,
                                 levels = c("Isolate","Timepoint","Block"))

ggplot(sig_counts_sun, aes(x = Factor, y = n_sig, fill = Factor)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = n_sig), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("#377EB8","#FF6347","#999999")) +
  labs(title = "Sunflower Host Genes: Significant per Factor (padj < 0.05)",
       y = "Number of significant genes", x = "") +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```


# ═══════════════════════════════════════════════════════════════════════════
# PART D: Combined Summary
# ═══════════════════════════════════════════════════════════════════════════

```{r combined-summary, fig.width=12, fig.height=6}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("COMPLETE ANOVA ANALYSIS SUMMARY\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("PERMANOVA R² (proportion of variance explained):\n")
cat("─────────────────────────────────────────────────\n")

# Extract R2 from each PERMANOVA
cat("\n  Pathogen genes (across both hosts):\n")
for (i in 1:4) {
  cat(sprintf("    %-35s R² = %.4f, p = %s\n",
              pub_perm_pathogen$Source[i],
              pub_perm_pathogen$R2[i],
              ifelse(is.na(pub_perm_pathogen$p_value[i]), "NA",
                     sprintf("%.4f", pub_perm_pathogen$p_value[i]))))
}

cat("\n  Soybean host genes:\n")
for (i in 1:3) {
  cat(sprintf("    %-35s R² = %.4f, p = %s\n",
              pub_perm_soy$Source[i],
              pub_perm_soy$R2[i],
              ifelse(is.na(pub_perm_soy$p_value[i]), "NA",
                     sprintf("%.4f", pub_perm_soy$p_value[i]))))
}

cat("\n  Sunflower host genes:\n")
for (i in 1:3) {
  cat(sprintf("    %-35s R² = %.4f, p = %s\n",
              pub_perm_sun$Source[i],
              pub_perm_sun$R2[i],
              ifelse(is.na(pub_perm_sun$p_value[i]), "NA",
                     sprintf("%.4f", pub_perm_sun$p_value[i]))))
}

cat("\n\nPER-GENE ANOVA (significant genes, padj < 0.05):\n")
cat("─────────────────────────────────────────────────\n")
summary_table <- data.frame(
  Dataset   = c("Pathogen","Pathogen","Pathogen","Pathogen",
                "Soy Host","Soy Host","Soy Host",
                "Sun Host","Sun Host","Sun Host"),
  Factor    = c("Host","Isolate","Timepoint","Block",
                "Isolate","Timepoint","Block",
                "Isolate","Timepoint","Block"),
  n_sig     = c(
    sum(pergene_results_pathogen$padj_host      < 0.05, na.rm = TRUE),
    sum(pergene_results_pathogen$padj_isolate   < 0.05, na.rm = TRUE),
    sum(pergene_results_pathogen$padj_timepoint < 0.05, na.rm = TRUE),
    sum(pergene_results_pathogen$padj_block     < 0.05, na.rm = TRUE),
    sum(pergene_results_soy$padj_isolate   < 0.05, na.rm = TRUE),
    sum(pergene_results_soy$padj_timepoint < 0.05, na.rm = TRUE),
    sum(pergene_results_soy$padj_block     < 0.05, na.rm = TRUE),
    sum(pergene_results_sun$padj_isolate   < 0.05, na.rm = TRUE),
    sum(pergene_results_sun$padj_timepoint < 0.05, na.rm = TRUE),
    sum(pergene_results_sun$padj_block     < 0.05, na.rm = TRUE)
  )
)
print(summary_table, row.names = FALSE)

# ── Combined bar plot ──
summary_table$Dataset <- factor(summary_table$Dataset,
                                 levels = c("Pathogen","Soy Host","Sun Host"))

ggplot(summary_table, aes(x = Factor, y = n_sig, fill = Dataset)) +
  geom_col(position = position_dodge(width = 0.8), alpha = 0.85) +
  geom_text(aes(label = n_sig),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Pathogen" = "#2ca02c",
                                "Soy Host" = "#ffcb21",
                                "Sun Host" = "#5e3a04")) +
  labs(title = "Significant Genes per Factor Across All Datasets",
       subtitle = "Per-gene Type III ANOVA, BH-corrected (padj < 0.05)",
       y = "Number of significant genes", x = "") +
  theme_bw(base_size = 14) +
  theme(legend.position = "top")
```

# Output File Summary

```{r file-summary}
cat("All output files in anova_results/:\n")
cat("────────────────────────────────────\n\n")
cat("PERMANOVA tables:\n")
cat("  PERMANOVA_pathogen_genes.csv\n")
cat("  PERMANOVA_soybean_host_genes.csv\n")
cat("  PERMANOVA_sunflower_host_genes.csv\n")
cat("\nPer-gene ANOVA (all genes):\n")
cat("  pergene_ANOVA_pathogen_all.csv\n")
cat("  pergene_ANOVA_soybean_host_all.csv\n")
cat("  pergene_ANOVA_sunflower_host_all.csv\n")
cat("\nPer-gene ANOVA (significant only):\n")
cat("  pergene_ANOVA_pathogen_sig_host.csv\n")
cat("  pergene_ANOVA_pathogen_sig_isolate.csv\n")
cat("  pergene_ANOVA_pathogen_sig_timepoint.csv\n")
cat("  pergene_ANOVA_soybean_host_sig_isolate.csv\n")
cat("  pergene_ANOVA_soybean_host_sig_timepoint.csv\n")
cat("  pergene_ANOVA_sunflower_host_sig_isolate.csv\n")
cat("  pergene_ANOVA_sunflower_host_sig_timepoint.csv\n")
cat("\nPairwise comparisons (Tukey HSD):\n")
cat("  pairwise_host_pathogen.csv\n")
cat("  pairwise_isolate_pathogen.csv\n")
cat("  pairwise_timepoint_pathogen.csv\n")
cat("  pairwise_isolate_soybean_host.csv\n")
cat("  pairwise_timepoint_soybean_host.csv\n")
cat("  pairwise_isolate_sunflower_host.csv\n")
cat("  pairwise_timepoint_sunflower_host.csv\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
