---
title: "Cross-Host Visualization: Venn Diagrams, UpSet Plot & Tanglegram"
subtitle: "Sclerotinia Pathogen DEGs — Sunflower (Ha) vs Soybean (Gm)"
author: "Analysis by wan00965"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 8, dpi = 300)
```

# 1. Setup

## Install Required Packages (Run Once)

```{r install-packages, eval=FALSE}
install.packages(c("VennDiagram", "UpSetR", "dendextend", "ggplot2",
                    "dplyr", "RColorBrewer", "gridExtra", "eulerr", "pheatmap"))
```

## Load Libraries

```{r load-libraries}
library(DESeq2)
library(VennDiagram)
library(UpSetR)
library(dendextend)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(gridExtra)
library(grid)
library(eulerr)
library(pheatmap)

cat("All libraries loaded successfully!\n")
```

## Set Working Directory

```{r set-wd}
# UPDATE this path to your data folder
setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_data/New_salmon_round3/new_run_count")
cat("Working directory:", getwd(), "\n")
```

# 2. Load DEG Data from Both Hosts

```{r load-degs}
timepoints <- c("24h", "48h", "96h")

# Helper: safely load a DEG file, return NULL if not found
safe_load_deg <- function(filepath) {
  if (file.exists(filepath)) {
    df <- read.csv(filepath, stringsAsFactors = FALSE, row.names = 1)
    if (nrow(df) > 0) return(df)
  }
  return(NULL)
}

# --- Sunflower (Ha) pathogen DEGs ---
ha_degs <- list()
for (tp in timepoints) {
  f <- sprintf("Ha_pathogen_aggressive_vs_low_DEGs_%s.csv", tp)
  ha_degs[[tp]] <- safe_load_deg(f)
  if (!is.null(ha_degs[[tp]])) {
    cat(sprintf("Ha %s: %d DEGs loaded\n", tp, nrow(ha_degs[[tp]])))
  } else {
    cat(sprintf("WARNING: Ha %s not found (%s)\n", tp, f))
  }
}

# --- Soybean (Gm) pathogen DEGs ---
gm_degs <- list()
for (tp in timepoints) {
  f <- sprintf("Gm_pathogen_aggressive_vs_low_DEGs_%s.csv", tp)
  gm_degs[[tp]] <- safe_load_deg(f)
  if (!is.null(gm_degs[[tp]])) {
    cat(sprintf("Gm %s: %d DEGs loaded\n", tp, nrow(gm_degs[[tp]])))
  } else {
    cat(sprintf("WARNING: Gm %s not found (%s)\n", tp, f))
  }
}

# --- Show available CSV files for debugging ---
cat("\n=== DEG CSV files in working directory ===\n")
csv_files <- list.files(pattern = "aggressive_vs_low.*\\.csv$", ignore.case = TRUE)
if (length(csv_files) > 0) {
  print(csv_files)
} else {
  cat("No matching CSV files found! Check filenames.\n")
  cat("Expected pattern: Ha_pathogen_aggressive_vs_low_DEGs_24h.csv\n")
  cat("All CSV files:\n")
  print(list.files(pattern = "\\.csv$"))
}
```

## Extract Gene IDs (with safe fallbacks)

```{r extract-gene-ids}
lfc_col <- "log2FoldChange"

# Safe extraction: returns character(0) if data is NULL or empty
safe_rownames <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(character(0))
  return(rownames(df))
}

safe_filter_rows <- function(df, direction = "up") {
  if (is.null(df) || nrow(df) == 0) return(character(0))
  if (!(lfc_col %in% colnames(df))) return(character(0))
  if (direction == "up") {
    return(rownames(df[df[[lfc_col]] > 0, ]))
  } else {
    return(rownames(df[df[[lfc_col]] < 0, ]))
  }
}

# Build all gene sets with safe fallbacks
ha_gene_sets <- list(); gm_gene_sets <- list()
ha_up_sets <- list(); ha_down_sets <- list()
gm_up_sets <- list(); gm_down_sets <- list()

for (tp in timepoints) {
  ha_gene_sets[[tp]] <- safe_rownames(ha_degs[[tp]])
  gm_gene_sets[[tp]] <- safe_rownames(gm_degs[[tp]])
  ha_up_sets[[tp]]   <- safe_filter_rows(ha_degs[[tp]], "up")
  ha_down_sets[[tp]] <- safe_filter_rows(ha_degs[[tp]], "down")
  gm_up_sets[[tp]]   <- safe_filter_rows(gm_degs[[tp]], "up")
  gm_down_sets[[tp]] <- safe_filter_rows(gm_degs[[tp]], "down")
}

# Which timepoints actually have data?
ha_available <- timepoints[sapply(timepoints, function(tp) length(ha_gene_sets[[tp]]) > 0)]
gm_available <- timepoints[sapply(timepoints, function(tp) length(gm_gene_sets[[tp]]) > 0)]

cat("\n=== DATA AVAILABILITY ===\n")
cat("Ha timepoints with data:", if (length(ha_available) > 0) paste(ha_available, collapse=", ") else "NONE", "\n")
cat("Gm timepoints with data:", if (length(gm_available) > 0) paste(gm_available, collapse=", ") else "NONE", "\n")

cat("\n=== GENE SET SIZES ===\n")
for (tp in timepoints) {
  cat(sprintf("%s  Ha: %d (Up:%d Down:%d)  |  Gm: %d (Up:%d Down:%d)\n", tp,
              length(ha_gene_sets[[tp]]), length(ha_up_sets[[tp]]), length(ha_down_sets[[tp]]),
              length(gm_gene_sets[[tp]]), length(gm_up_sets[[tp]]), length(gm_down_sets[[tp]])))
}
```

---

# 3. Venn Diagrams — DEGs Across 3 Timepoints

```{r venn-helper}
# Helper function to compute all 7 regions of a 3-set Venn
compute_venn3_stats <- function(set_a, set_b, set_c, names = c("A","B","C")) {
  all_genes <- unique(c(set_a, set_b, set_c))
  n_total <- length(all_genes)
  
  only_a  <- setdiff(set_a, union(set_b, set_c))
  only_b  <- setdiff(set_b, union(set_a, set_c))
  only_c  <- setdiff(set_c, union(set_a, set_b))
  only_ab <- setdiff(intersect(set_a, set_b), set_c)
  only_ac <- setdiff(intersect(set_a, set_c), set_b)
  only_bc <- setdiff(intersect(set_b, set_c), set_a)
  abc     <- Reduce(intersect, list(set_a, set_b, set_c))
  
  counts <- c(length(only_a), length(only_b), length(only_c),
              length(only_ab), length(only_ac), length(only_bc), length(abc))
  pcts <- if (n_total > 0) round(100 * counts / n_total, 1) else rep(0, 7)
  
  region_names <- c(paste0("Only ",names[1]), paste0("Only ",names[2]), paste0("Only ",names[3]),
                    paste0(names[1]," & ",names[2]," only"),
                    paste0(names[1]," & ",names[3]," only"),
                    paste0(names[2]," & ",names[3]," only"),
                    "All three")
  
  euler_input <- setNames(counts, c(
    names[1], names[2], names[3],
    paste0(names[1],"&",names[2]),
    paste0(names[1],"&",names[3]),
    paste0(names[2],"&",names[3]),
    paste0(names[1],"&",names[2],"&",names[3])
  ))
  
  list(counts=counts, pcts=pcts, n_total=n_total,
       region_names=region_names, euler_input=euler_input)
}
```

## 3.1 Sunflower (Ha) Pathogen DEGs — 3-Way Venn

```{r venn-ha-eulerr, fig.width=9, fig.height=8}
venn_colors <- c("#6495ED", "#FFD700", "#32CD32")
venn_fills  <- adjustcolor(venn_colors, alpha.f = 0.45)

if (length(ha_available) >= 3) {
  
  ha_stats <- compute_venn3_stats(
    ha_gene_sets[["24h"]], ha_gene_sets[["48h"]], ha_gene_sets[["96h"]],
    names = c("24 HPI", "48 HPI", "96 HPI"))
  
  cat("=== SUNFLOWER (Ha) VENN SUMMARY ===\n")
  cat(sprintf("Total unique pathogen DEGs: %d\n\n", ha_stats$n_total))
  for (i in seq_along(ha_stats$counts)) {
    cat(sprintf("  %s: %d (%.1f%%)\n", ha_stats$region_names[i],
                ha_stats$counts[i], ha_stats$pcts[i]))
  }
  
  fit_ha <- euler(ha_stats$euler_input)
  plot(fit_ha,
       fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.3, font = 2, col = venn_colors),
       quantities = list(cex = 1.2, font = 2, type = c("counts", "percent")),
       main = list(label = "Pathogen DEGs in Ha89 (Aggressive vs Low Isolates)",
                   cex = 1.5, font = 2))
  
} else if (length(ha_available) == 2) {
  cat("Only 2 Ha timepoints with data — drawing 2-set Venn\n")
  tp1 <- ha_available[1]; tp2 <- ha_available[2]
  fit_ha <- euler(c(
    setNames(length(setdiff(ha_gene_sets[[tp1]], ha_gene_sets[[tp2]])), tp1),
    setNames(length(setdiff(ha_gene_sets[[tp2]], ha_gene_sets[[tp1]])), tp2),
    setNames(length(intersect(ha_gene_sets[[tp1]], ha_gene_sets[[tp2]])),
             paste0(tp1,"&",tp2))))
  plot(fit_ha, fills = list(fill = venn_fills[1:2], alpha = 0.7),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Sunflower (Ha89)", cex = 1.5, font = 2))
} else {
  cat("Fewer than 2 Ha timepoints — cannot draw Venn.\n")
  cat("Available files:\n")
  print(list.files(pattern = "Ha_pathogen"))
}
```

```{r venn-ha-classic, fig.width=8, fig.height=8}
# Classic VennDiagram package (non-proportional circles, like your Fig. 5)
if (length(ha_available) >= 3) {
  
  futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
  
  venn.plot.ha <- venn.diagram(
    x = list("24 HPI" = ha_gene_sets[["24h"]],
             "48 HPI" = ha_gene_sets[["48h"]],
             "96 HPI" = ha_gene_sets[["96h"]]),
    filename = NULL,
    fill = venn_fills, col = venn_colors, lwd = 2.5,
    category.names = c("24 HPI","48 HPI","96 HPI"),
    cat.cex = 1.8, cat.fontface = "bold", cat.col = venn_colors,
    cat.dist = c(0.06, 0.06, 0.06),
    cex = 1.6, fontface = "bold", fontfamily = "sans",
    main = "Pathogen DEGs in Sunflower (Ha89)",
    main.cex = 2, main.fontface = "bold",
    euler.d = FALSE, scaled = FALSE, margin = 0.1)
  
  grid.newpage()
  grid.draw(venn.plot.ha)
}
```

## 3.2 Soybean (Gm) Pathogen DEGs — 3-Way Venn

```{r venn-gm-eulerr, fig.width=9, fig.height=8}
if (length(gm_available) >= 3) {
  
  gm_stats <- compute_venn3_stats(
    gm_gene_sets[["24h"]], gm_gene_sets[["48h"]], gm_gene_sets[["96h"]],
    names = c("24 HPI", "48 HPI", "96 HPI"))
  
  cat("=== SOYBEAN (Gm) VENN SUMMARY ===\n")
  cat(sprintf("Total unique pathogen DEGs: %d\n\n", gm_stats$n_total))
  for (i in seq_along(gm_stats$counts)) {
    cat(sprintf("  %s: %d (%.1f%%)\n", gm_stats$region_names[i],
                gm_stats$counts[i], gm_stats$pcts[i]))
  }
  
  fit_gm <- euler(gm_stats$euler_input)
  plot(fit_gm,
       fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.3, font = 2, col = venn_colors),
       quantities = list(cex = 1.2, font = 2, type = c("counts", "percent")),
       main = list(label = "Pathogen DEGs in Williams 82 (Aggressive vs Low Isolates)",
                   cex = 1.5, font = 2))
  
} else if (length(gm_available) == 2) {
  cat("Only 2 Gm timepoints with data — drawing 2-set Venn\n")
  tp1 <- gm_available[1]; tp2 <- gm_available[2]
  fit_gm <- euler(c(
    setNames(length(setdiff(gm_gene_sets[[tp1]], gm_gene_sets[[tp2]])), tp1),
    setNames(length(setdiff(gm_gene_sets[[tp2]], gm_gene_sets[[tp1]])), tp2),
    setNames(length(intersect(gm_gene_sets[[tp1]], gm_gene_sets[[tp2]])),
             paste0(tp1,"&",tp2))))
  plot(fit_gm, fills = list(fill = venn_fills[1:2], alpha = 0.7),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Soybean (Wm82)", cex = 1.5, font = 2))
} else {
  cat("Fewer than 2 Gm timepoints — cannot draw Venn.\n")
  cat("Available files:\n")
  print(list.files(pattern = "Gm_pathogen"))
}
```

```{r venn-gm-classic, fig.width=8, fig.height=8}
if (length(gm_available) >= 3) {
  
  futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
  
  venn.plot.gm <- venn.diagram(
    x = list("24 HPI" = gm_gene_sets[["24h"]],
             "48 HPI" = gm_gene_sets[["48h"]],
             "96 HPI" = gm_gene_sets[["96h"]]),
    filename = NULL,
    fill = venn_fills, col = venn_colors, lwd = 2.5,
    category.names = c("24 HPI","48 HPI","96 HPI"),
    cat.cex = 1.8, cat.fontface = "bold", cat.col = venn_colors,
    cat.dist = c(0.06, 0.06, 0.06),
    cex = 1.6, fontface = "bold", fontfamily = "sans",
    main = "Pathogen DEGs in Soybean (Williams 82)",
    main.cex = 2, main.fontface = "bold",
    euler.d = FALSE, scaled = FALSE, margin = 0.1)
  
  grid.newpage()
  grid.draw(venn.plot.gm)
}
```

## 3.3 Save Venns as PDF

```{r save-venns, eval=FALSE}
if (exists("fit_ha")) {
  pdf("Fig_Venn_Ha_pathogen_DEGs.pdf", width = 8, height = 8)
  plot(fit_ha, fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.2, font = 2),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Sunflower (Ha89)", cex = 1.5, font = 2))
  dev.off(); cat("Saved: Fig_Venn_Ha_pathogen_DEGs.pdf\n")
}
if (exists("fit_gm")) {
  pdf("Fig_Venn_Gm_pathogen_DEGs.pdf", width = 8, height = 8)
  plot(fit_gm, fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.2, font = 2),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Soybean (Wm82)", cex = 1.5, font = 2))
  dev.off(); cat("Saved: Fig_Venn_Gm_pathogen_DEGs.pdf\n")
}
```

---

# 4. UpSet Plot — Cross-Host DEG Overlap

```{r upset-cross-host, fig.width=14, fig.height=8}
# Build named list of all available sets
all_sets <- list()
for (tp in ha_available) all_sets[[paste0("Ha_",tp)]] <- ha_gene_sets[[tp]]
for (tp in gm_available) all_sets[[paste0("Gm_",tp)]] <- gm_gene_sets[[tp]]

if (length(all_sets) >= 2) {
  
  all_genes <- unique(unlist(all_sets))
  upset_df <- data.frame(row.names = all_genes)
  for (nm in names(all_sets)) {
    upset_df[[nm]] <- as.integer(all_genes %in% all_sets[[nm]])
  }
  
  # ---------------------------------------------------------------
  # NOTE: UpSetR's intersects() query is fragile — it crashes if
  # the highlighted intersection doesn't appear among the top N
  # displayed bars. We use a safe approach: color bars by set
  # membership using the "sets.bar.color" parameter instead,
  # and skip the problematic queries entirely.
  # ---------------------------------------------------------------
  
  # Color the set size bars: orange for Ha, blue for Gm
  set_names_ordered <- rev(names(all_sets))
  bar_colors <- ifelse(grepl("^Ha_", set_names_ordered), "#E69F00", "#56B4E9")
  
  upset(upset_df,
        sets = set_names_ordered,
        order.by = "freq",
        nsets = length(all_sets),
        nintersects = 30,
        mb.ratio = c(0.6, 0.4),
        point.size = 3.5,
        line.size = 1.5,
        sets.bar.color = bar_colors,
        mainbar.y.label = "Intersection Size",
        sets.x.label = "Number of DEGs",
        text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
        set_size.show = TRUE)
  
  # Print cross-host overlap stats for the UpSet context
  cat("\n=== UPSET INTERSECTION SUMMARY ===\n")
  cat(sprintf("Total unique pathogen DEGs across all conditions: %d\n", length(all_genes)))
  cat(sprintf("Sets included: %s\n", paste(names(all_sets), collapse = ", ")))
  
  # Show top intersections manually
  # Each gene's membership pattern as a string
  patterns <- apply(upset_df[, names(all_sets), drop = FALSE], 1,
                    function(row) paste(names(all_sets)[row == 1], collapse = " + "))
  pattern_counts <- sort(table(patterns), decreasing = TRUE)
  cat("\nTop 10 intersection patterns:\n")
  print(head(pattern_counts, 10))
  
} else {
  cat("Need at least 2 gene sets for UpSet plot.\n")
}
```

## 4.2 Cross-Host Venn: Ha vs Gm (All Timepoints Combined)

```{r venn-2host, fig.width=9, fig.height=7}
ha_all_genes <- unique(unlist(ha_gene_sets))
gm_all_genes <- unique(unlist(gm_gene_sets))

if (length(ha_all_genes) > 0 && length(gm_all_genes) > 0) {
  
  shared_hosts <- intersect(ha_all_genes, gm_all_genes)
  ha_only <- setdiff(ha_all_genes, gm_all_genes)
  gm_only <- setdiff(gm_all_genes, ha_all_genes)
  
  cat("\n=== CROSS-HOST OVERLAP (All Timepoints Combined) ===\n")
  cat(sprintf("Ha-only pathogen DEGs: %d\n", length(ha_only)))
  cat(sprintf("Gm-only pathogen DEGs: %d\n", length(gm_only)))
  cat(sprintf("Shared between hosts:  %d\n", length(shared_hosts)))
  
  venn_2host <- euler(c(
    "Sunflower\n(Ha89)" = length(ha_only),
    "Soybean\n(Wm82)"   = length(gm_only),
    "Sunflower\n(Ha89)&Soybean\n(Wm82)" = length(shared_hosts)))
  
  plot(venn_2host,
       fills = list(fill = c("#FF9933","#6699CC"), alpha = 0.6),
       edges = list(col = c("#CC6600","#336699"), lwd = 3),
       labels = list(cex = 1.4, font = 2),
       quantities = list(cex = 1.5, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs: Sunflower vs Soybean (All Timepoints Combined)",
                   cex = 1.5, font = 2))
} else {
  cat("Need data from both hosts for cross-host comparison.\n")
}
```

## 4.3 Per-Timepoint Cross-Host Comparison

```{r venn-per-tp, fig.width=15, fig.height=5}
both_available <- intersect(ha_available, gm_available)

if (length(both_available) > 0) {
  
  tp_summaries <- data.frame()
  plots_per_tp <- list()
  
  for (tp in both_available) {
    shared_tp <- intersect(ha_gene_sets[[tp]], gm_gene_sets[[tp]])
    ha_only_tp <- setdiff(ha_gene_sets[[tp]], gm_gene_sets[[tp]])
    gm_only_tp <- setdiff(gm_gene_sets[[tp]], ha_gene_sets[[tp]])
    total_union <- length(union(ha_gene_sets[[tp]], gm_gene_sets[[tp]]))
    
    tp_summaries <- rbind(tp_summaries, data.frame(
      Timepoint = tp,
      Ha_only = length(ha_only_tp), Shared = length(shared_tp),
      Gm_only = length(gm_only_tp),
      Ha_total = length(ha_gene_sets[[tp]]), Gm_total = length(gm_gene_sets[[tp]]),
      Overlap_pct = if(total_union>0) round(100*length(shared_tp)/total_union,1) else 0))
    
    fit_tp <- euler(c("Ha"=length(ha_only_tp), "Gm"=length(gm_only_tp),
                       "Ha&Gm"=length(shared_tp)))
    plots_per_tp[[tp]] <- plot(fit_tp,
         fills = list(fill = c("#FF9933","#6699CC"), alpha = 0.6),
         edges = list(col = c("#CC6600","#336699"), lwd = 2.5),
         labels = list(cex = 1.1, font = 2),
         quantities = list(cex = 1.2, font = 2, type = c("counts","percent")),
         main = list(label = tp, cex = 1.4, font = 2))
  }
  
  cat("\n=== PER-TIMEPOINT CROSS-HOST OVERLAP ===\n")
  print(tp_summaries)
  
  gridExtra::grid.arrange(
    grobs = plots_per_tp, ncol = length(both_available),
    top = textGrob("Pathogen DEGs: Sunflower vs Soybean per Timepoint",
                   gp = gpar(fontsize = 16, fontface = "bold")))
} else {
  cat("No shared timepoints between hosts.\n")
}
```

---

# 5. Tanglegram — Comparing DEG Profiles Between Hosts

```{r tanglegram, fig.width=14, fig.height=10}
shared_genes_any <- if (length(ha_all_genes) > 0 && length(gm_all_genes) > 0) {
  intersect(ha_all_genes, gm_all_genes)
} else { character(0) }

tps_both <- intersect(ha_available, gm_available)

cat(sprintf("Shared DEGs: %d genes | Shared timepoints: %s\n",
            length(shared_genes_any), paste(tps_both, collapse=", ")))

if (length(shared_genes_any) >= 20 && length(tps_both) >= 2) {
  
  # Build fold-change matrices
  build_fc_matrix <- function(degs_list, genes, tps) {
    mat <- matrix(0, nrow=length(genes), ncol=length(tps),
                  dimnames=list(genes, tps))
    for (tp in tps) {
      df <- degs_list[[tp]]
      if (!is.null(df) && lfc_col %in% colnames(df)) {
        shared <- intersect(rownames(df), genes)
        if (length(shared) > 0) mat[shared, tp] <- df[shared, lfc_col]
      }
    }
    return(mat)
  }
  
  fc_ha <- build_fc_matrix(ha_degs, shared_genes_any, tps_both)
  fc_gm <- build_fc_matrix(gm_degs, shared_genes_any, tps_both)
  
  # Select top N genes by max absolute fold change
  max_fc <- pmax(apply(abs(fc_ha), 1, max, na.rm=TRUE),
                  apply(abs(fc_gm), 1, max, na.rm=TRUE))
  
  n_top <- min(50, length(shared_genes_any))
  top_genes <- names(sort(max_fc, decreasing=TRUE))[1:n_top]
  
  fc_ha_top <- fc_ha[top_genes, , drop=FALSE]
  fc_gm_top <- fc_gm[top_genes, , drop=FALSE]
  
  # Short gene labels
  short_names <- sapply(top_genes, function(id) {
    parts <- strsplit(id, "\\|")[[1]]
    if (length(parts) >= 4) return(parts[4])
    if (length(parts) >= 3) return(parts[3])
    return(substr(id, 1, 20))
  })
  rownames(fc_ha_top) <- short_names
  rownames(fc_gm_top) <- short_names
  
  # Cluster and color
  dend_ha <- as.dendrogram(hclust(dist(fc_ha_top), method="ward.D2"))
  dend_gm <- as.dendrogram(hclust(dist(fc_gm_top), method="ward.D2"))
  
  k_cl <- min(4, max(2, n_top %/% 8))
  cols <- c("#E69F00","#56B4E9","#009E73","#CC79A7")
  dend_ha <- color_branches(dend_ha, k=k_cl, col=cols[1:k_cl])
  dend_gm <- color_branches(dend_gm, k=k_cl, col=cols[1:k_cl])
  
  dl <- dendlist(dend_ha, dend_gm) %>% untangle(method="step2side")
  
  tanglegram(dl,
             main_left = "Sunflower (Ha89)",
             main_right = "Soybean (Wm82)",
             common_subtrees_color_branches = TRUE,
             highlight_distinct_edges = TRUE,
             highlight_branches_lwd = FALSE,
             margin_inner = 8, lwd = 1.8,
             columns_width = c(5, 3, 5),
             cex_main = 1.5, cex_main_left = 1.4, cex_main_right = 1.4,
             lab.cex = 0.7,
             main = paste0("Top ", n_top, " Shared Pathogen DEGs: Sunflower vs Soybean"))
  
  cat(sprintf("\nEntanglement: %.3f (0=identical, 1=max crossing)\n", entanglement(dl)))
  
} else {
  cat("Not enough shared genes or timepoints for tanglegram.\n")
}
```

---

# 6. Heatmap — Cross-Host Fold Change Comparison

```{r heatmap, fig.width=12, fig.height=10}
if (exists("fc_ha") && exists("fc_gm") && length(shared_genes_any) >= 10) {

  # ── Use ALL shared significant genes (no top-N cap) ───────────────────────
  all_heat <- shared_genes_any

  # Build combined fold-change matrix across both hosts
  build_fc_matrix_all <- function(degs_list, genes, tps) {
    mat <- matrix(0, nrow = length(genes), ncol = length(tps),
                  dimnames = list(genes, tps))
    for (tp in tps) {
      df <- degs_list[[tp]]
      if (!is.null(df) && lfc_col %in% colnames(df)) {
        shared <- intersect(rownames(df), genes)
        if (length(shared) > 0) mat[shared, tp] <- df[shared, lfc_col]
      }
    }
    return(mat)
  }

  fc_ha_heat <- build_fc_matrix_all(ha_degs, all_heat, tps_both)
  fc_gm_heat <- build_fc_matrix_all(gm_degs, all_heat, tps_both)

  fc_combined <- cbind(fc_ha_heat, fc_gm_heat)
  colnames(fc_combined) <- c(paste0("Ha_", tps_both), paste0("Gm_", tps_both))

  # Short gene labels
  rownames(fc_combined) <- sapply(all_heat, function(id) {
    parts <- strsplit(id, "\\|")[[1]]
    if (length(parts) >= 4) return(parts[4])
    if (length(parts) >= 3) return(parts[3])
    return(substr(id, 1, 20))
  })

  # Column annotations
  col_annot <- data.frame(
    Host      = rep(c("Sunflower", "Soybean"), each = length(tps_both)),
    Timepoint = rep(tps_both, 2),
    row.names = colnames(fc_combined))

  annot_colors <- list(
    Host      = c("Sunflower" = "#FF9933", "Soybean" = "#6699CC"),
    Timepoint = c("24h" = "#66C2A5", "48h" = "#FC8D62", "96h" = "#8DA0CB")[tps_both])

  max_abs <- max(abs(fc_combined), na.rm = TRUE)
  if (max_abs == 0) max_abs <- 1

  # Scale figure height dynamically: ~0.18 inches per gene, min 8, max 60
  n_genes   <- nrow(fc_combined)
  fig_h     <- max(8, min(60, n_genes * 0.18))
  font_row  <- ifelse(n_genes <= 60, 7, ifelse(n_genes <= 120, 5, 4))

  cat(sprintf("Plotting heatmap for all %d shared significant genes\n", n_genes))

  pheatmap(fc_combined,
           color            = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
           breaks           = seq(-max_abs, max_abs, length.out = 101),
           cluster_cols     = FALSE,
           clustering_method = "ward.D2",
           annotation_col   = col_annot,
           annotation_colors = annot_colors,
           fontsize_row     = font_row,
           fontsize_col     = 11,
           main             = paste0("All ", n_genes,
                                     " Shared Pathogen DEGs: log2FC (Aggressive vs Low)"),
           border_color     = NA,
           cellwidth        = 35,
           height           = fig_h)

} else {
  cat("Not enough shared data for heatmap.\n")
}
```

---

# 6b. Heatmap — Genes Shared in Both Hosts at ≥2 Timepoints

```{r heatmap-2tp, fig.width=12, fig.height=10}
if (length(shared_genes_any) >= 5 && length(tps_both) >= 2) {

  # ── Filter: keep only genes that are a DEG in BOTH hosts at ≥2 timepoints ─
  n_tp_shared <- sapply(shared_genes_any, function(g) {
    sum(sapply(tps_both,
               function(tp) g %in% ha_gene_sets[[tp]] && g %in% gm_gene_sets[[tp]]))
  })

  genes_2tp <- shared_genes_any[n_tp_shared >= 2]

  cat(sprintf("Genes shared in both hosts at >=2 timepoints: %d / %d total shared\n",
              length(genes_2tp), length(shared_genes_any)))

  if (length(genes_2tp) >= 5) {

    # ── Build FC matrix for filtered genes ──────────────────────────────────
    build_fc_mat <- function(degs_list, genes, tps) {
      mat <- matrix(0, nrow = length(genes), ncol = length(tps),
                    dimnames = list(genes, tps))
      for (tp in tps) {
        df <- degs_list[[tp]]
        if (!is.null(df) && lfc_col %in% colnames(df)) {
          sh <- intersect(rownames(df), genes)
          if (length(sh) > 0) mat[sh, tp] <- df[sh, lfc_col]
        }
      }
      return(mat)
    }

    fc_ha_2tp <- build_fc_mat(ha_degs, genes_2tp, tps_both)
    fc_gm_2tp <- build_fc_mat(gm_degs, genes_2tp, tps_both)

    fc_comb_2tp <- cbind(fc_ha_2tp, fc_gm_2tp)
    colnames(fc_comb_2tp) <- c(paste0("Ha_", tps_both), paste0("Gm_", tps_both))

    # Short gene labels
    rownames(fc_comb_2tp) <- sapply(genes_2tp, function(id) {
      parts <- strsplit(id, "\\|")[[1]]
      if (length(parts) >= 4) return(parts[4])
      if (length(parts) >= 3) return(parts[3])
      return(substr(id, 1, 20))
    })

    # Add row annotation: how many timepoints shared
    row_annot <- data.frame(
      Shared_TPs = factor(n_tp_shared[genes_2tp]),
      row.names  = rownames(fc_comb_2tp))

    # Column annotation
    col_annot_2tp <- data.frame(
      Host      = rep(c("Sunflower", "Soybean"), each = length(tps_both)),
      Timepoint = rep(tps_both, 2),
      row.names = colnames(fc_comb_2tp))

    n_tp_levels <- as.character(sort(unique(n_tp_shared[genes_2tp])))
    tp_pal <- c("2" = "#FDAE61", "3" = "#D7191C")

    annot_colors_2tp <- list(
      Host       = c("Sunflower" = "#FF9933", "Soybean" = "#6699CC"),
      Timepoint  = c("24h" = "#66C2A5", "48h" = "#FC8D62", "96h" = "#8DA0CB")[tps_both],
      Shared_TPs = tp_pal[n_tp_levels])

    max_abs_2tp <- max(abs(fc_comb_2tp), na.rm = TRUE)
    if (max_abs_2tp == 0) max_abs_2tp <- 1

    n_g       <- nrow(fc_comb_2tp)
    fig_h_2tp <- max(6, min(60, n_g * 0.18))
    font_r    <- ifelse(n_g <= 60, 7, ifelse(n_g <= 120, 5, 4))

    pheatmap(fc_comb_2tp,
             color             = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
             breaks            = seq(-max_abs_2tp, max_abs_2tp, length.out = 101),
             cluster_cols      = FALSE,
             clustering_method = "ward.D2",
             annotation_col    = col_annot_2tp,
             annotation_row    = row_annot,
             annotation_colors = annot_colors_2tp,
             fontsize_row      = font_r,
             fontsize_col      = 11,
             main              = paste0(n_g,
               " Shared Pathogen DEGs (Both Hosts, ≥2 Timepoints): log2FC (Aggressive vs Low)"),
             border_color      = NA,
             cellwidth         = 35,
             height            = fig_h_2tp)

  } else {
    cat("Fewer than 5 genes pass the >=2 shared timepoints filter — skipping heatmap.\n")
  }

} else {
  cat("Not enough shared data for this heatmap.\n")
}
```

# 6c. Export Genes Shared at ≥2 Timepoints to CSV

```{r export-shared-2tp}
if (exists("genes_2tp") && length(genes_2tp) >= 1) {

  # ── Build full FC and padj matrices for the filtered genes ────────────────
  build_fc_export <- function(degs_list, genes, tps) {
    mat <- matrix(NA_real_, nrow = length(genes), ncol = length(tps),
                  dimnames = list(genes, tps))
    for (tp in tps) {
      df <- degs_list[[tp]]
      if (!is.null(df) && lfc_col %in% colnames(df)) {
        sh <- intersect(rownames(df), genes)
        if (length(sh) > 0) mat[sh, tp] <- df[sh, lfc_col]
      }
    }
    return(mat)
  }

  get_padj_export <- function(degs_list, genes, tps, col) {
    mat <- matrix(NA_real_, nrow = length(genes), ncol = length(tps),
                  dimnames = list(genes, tps))
    if (is.null(col)) return(mat)
    for (tp in tps) {
      df <- degs_list[[tp]]
      if (!is.null(df) && col %in% colnames(df)) {
        sh <- intersect(rownames(df), genes)
        if (length(sh) > 0) mat[sh, tp] <- df[sh, col]
      }
    }
    return(mat)
  }

  # Detect padj column
  padj_col_2tp <- NULL
  for (tp in timepoints) {
    df_check <- ha_degs[[tp]]
    if (!is.null(df_check)) {
      if ("padj"      %in% colnames(df_check)) { padj_col_2tp <- "padj";      break }
      if ("FDR"       %in% colnames(df_check)) { padj_col_2tp <- "FDR";       break }
      if ("adj.P.Val" %in% colnames(df_check)) { padj_col_2tp <- "adj.P.Val"; break }
    }
  }

  fc_ha_e  <- build_fc_export(ha_degs, genes_2tp, timepoints)
  fc_gm_e  <- build_fc_export(gm_degs, genes_2tp, timepoints)
  padj_ha_e <- get_padj_export(ha_degs, genes_2tp, timepoints, padj_col_2tp)
  padj_gm_e <- get_padj_export(gm_degs, genes_2tp, timepoints, padj_col_2tp)

  # ── Assemble data frame ───────────────────────────────────────────────────
  df_2tp <- data.frame(gene_id = genes_2tp, row.names = genes_2tp,
                       stringsAsFactors = FALSE)

  # Per-timepoint DEG flags
  for (tp in timepoints) {
    df_2tp[[paste0("Ha_DEG_", tp)]] <- genes_2tp %in% ha_gene_sets[[tp]]
    df_2tp[[paste0("Gm_DEG_", tp)]] <- genes_2tp %in% gm_gene_sets[[tp]]
  }

  # Number of timepoints shared in both hosts simultaneously
  df_2tp$n_timepoints_shared <- n_tp_shared[genes_2tp]

  # log2FC columns
  for (tp in timepoints) {
    df_2tp[[paste0("Ha_log2FC_", tp)]] <- fc_ha_e[genes_2tp, tp]
    df_2tp[[paste0("Gm_log2FC_", tp)]] <- fc_gm_e[genes_2tp, tp]
    if (!is.null(padj_col_2tp)) {
      df_2tp[[paste0("Ha_padj_", tp)]] <- padj_ha_e[genes_2tp, tp]
      df_2tp[[paste0("Gm_padj_", tp)]] <- padj_gm_e[genes_2tp, tp]
    }
  }

  # Max absolute LFC across all conditions
  lfc_cols_2tp <- grep("_log2FC_", colnames(df_2tp), value = TRUE)
  df_2tp$max_abs_log2FC <- apply(abs(df_2tp[, lfc_cols_2tp, drop = FALSE]),
                                  1, max, na.rm = TRUE)

  # Sort: most timepoints shared first, then strongest fold change
  df_2tp <- df_2tp[order(-df_2tp$n_timepoints_shared, -df_2tp$max_abs_log2FC), ]

  # ── Write CSV ─────────────────────────────────────────────────────────────
  out_2tp <- "shared_significant_genes_min2tp.csv"
  write.csv(df_2tp, file = out_2tp, row.names = FALSE)

  cat(sprintf("Exported %d genes (shared in both hosts at >=2 timepoints) to: %s\n",
              nrow(df_2tp), out_2tp))
  cat(sprintf("Columns: %s\n\n", paste(colnames(df_2tp), collapse = ", ")))

  preview_cols_2tp <- c("gene_id", "n_timepoints_shared", "max_abs_log2FC",
                         grep("log2FC", colnames(df_2tp), value = TRUE))
  cat("Preview (first 6 rows):\n")
  print(head(df_2tp[, preview_cols_2tp], 6), digits = 3)

} else {
  cat("No genes passed the >=2 shared timepoints filter — nothing to export.\n")
}
```

---

# 6d. Heatmap — ≥2 Shared Timepoints, Row-wise Z-score of VST Expression

```{r heatmap-zscore-vst, fig.width=28, fig.height=14}
# ── UPDATE these count file names if yours differ ─────────────────────────
ha_counts_file <- "gene_counts_sunflower_SALMON_PATHOGEN_only.txt"
gm_counts_file <- "gene_counts_soybean_SALMON_PATHOGEN_only.txt"

# ── Helper: build metadata from sample names ──────────────────────────────
# Sample name pattern: timepoint_replicate_species_isolate
# e.g. "24_1_Ha_Xtra7", "48_2_Gm_WISS47"
build_meta <- function(sample_names) {
  meta <- data.frame(
    sample      = sample_names,
    row.names   = sample_names,
    stringsAsFactors = FALSE)
  meta$timepoint_num <- sapply(strsplit(sample_names, "_"), `[`, 1)
  meta$replicate     <- sapply(strsplit(sample_names, "_"), `[`, 2)
  meta$isolate       <- sapply(strsplit(sample_names, "_"), function(x) x[length(x)])
  meta$timepoint     <- sapply(sample_names, function(x) {
    parts <- strsplit(x, "_")[[1]]
    if (parts[length(parts)] == "NC") "NC" else paste0(parts[1], "h")
  })
  meta$aggressiveness <- sapply(meta$isolate, function(iso) {
    if (iso == "NC")                          return("Control")
    if (iso %in% c("WISS47", "JS659"))        return("Low")
    if (iso %in% c("Xtra7", "MNSS6"))         return("High")
    if (iso == "SSPotter")                    return("Soy-specific")
    if (iso == "BN172")                       return("Sun-specific")
    return("Unknown")
  })
  meta$timepoint     <- factor(meta$timepoint,
                                levels = c("NC","24h","48h","96h"))
  meta$isolate       <- factor(meta$isolate)
  meta$aggressiveness <- factor(meta$aggressiveness,
                                 levels = c("Control","Low","High",
                                            "Soy-specific","Sun-specific"))
  return(meta)
}

# ── Helper: load counts, build DESeq2 object, return VST matrix ───────────
build_vst <- function(counts_file, label) {
  if (!file.exists(counts_file)) {
    cat(sprintf("Count file not found: %s\n", counts_file))
    return(NULL)
  }
  cnt  <- round(read.table(counts_file, header = TRUE, row.names = 1,
                            check.names = FALSE))
  meta <- build_meta(colnames(cnt))

  # Keep only infected (non-NC) samples
  meta_inf <- meta[meta$timepoint != "NC", ]
  cnt_inf  <- cnt[, rownames(meta_inf)]

  # Filter lowly expressed genes
  keep <- rowSums(cnt_inf >= 5) >= 3
  cnt_filt <- cnt_inf[keep, ]

  cat(sprintf("%s: %d pathogen genes (filtered), %d samples\n",
              label, nrow(cnt_filt), ncol(cnt_filt)))

  dds <- DESeqDataSetFromMatrix(
    countData = cnt_filt,
    colData   = meta_inf,
    design    = ~ isolate + timepoint)
  dds <- DESeq(dds, quiet = TRUE)

  vst_obj <- vst(dds, blind = FALSE)
  vst_mat <- assay(vst_obj)
  attr(vst_mat, "meta") <- meta_inf
  return(vst_mat)
}

# ── Build VST matrices ────────────────────────────────────────────────────
cat("Building VST for Sunflower (Ha) pathogen...\n")
vst_ha_mat <- build_vst(ha_counts_file, "Ha")

cat("\nBuilding VST for Soybean (Gm) pathogen...\n")
vst_gm_mat <- build_vst(gm_counts_file, "Gm")

# ── Proceed only if both VST matrices and genes_2tp are available ─────────
if (!is.null(vst_ha_mat) && !is.null(vst_gm_mat) &&
    exists("genes_2tp") && length(genes_2tp) >= 5) {

  genes_in_both <- intersect(genes_2tp,
                              intersect(rownames(vst_ha_mat), rownames(vst_gm_mat)))
  cat(sprintf("\nGenes (>=2 shared TPs) present in both VST matrices: %d\n",
              length(genes_in_both)))

  if (length(genes_in_both) >= 5) {

    # ── Combine VST matrices, adding host prefix to sample names ────────────
    meta_ha <- attr(vst_ha_mat, "meta")
    meta_gm <- attr(vst_gm_mat, "meta")

    colnames(vst_ha_mat) <- paste0("Ha_", colnames(vst_ha_mat))
    colnames(vst_gm_mat) <- paste0("Gm_", colnames(vst_gm_mat))
    rownames(meta_ha)    <- paste0("Ha_", rownames(meta_ha))
    rownames(meta_gm)    <- paste0("Gm_", rownames(meta_gm))

    vst_comb <- cbind(vst_ha_mat[genes_in_both, ],
                      vst_gm_mat[genes_in_both, ])

    # ── Row-wise z-score computed SEPARATELY per host to avoid read-depth bias ─
    # Z-scoring across both hosts combined would inflate/deflate scores based on
    # whichever host has higher read counts. Instead, each host is z-scored
    # independently so the mean and SD are host-specific.
    zscore_host <- function(mat) {
      t(apply(mat, 1, function(x) {
        s <- sd(x, na.rm = TRUE)
        if (is.na(s) || s == 0) return(rep(0, length(x)))
        (x - mean(x, na.rm = TRUE)) / s
      }))
    }

    zscore_ha  <- zscore_host(vst_ha_mat[genes_in_both, ])
    zscore_gm  <- zscore_host(vst_gm_mat[genes_in_both, ])
    zscore_mat <- cbind(zscore_ha, zscore_gm)
    colnames(zscore_mat) <- c(colnames(vst_ha_mat), colnames(vst_gm_mat))

    # Clamp to [-2, +2]
    zscore_clamped <- pmin(pmax(zscore_mat, -2), 2)

    # ── Column annotations: Host, Aggressiveness, Isolate, Timepoint ────────
    meta_combined <- rbind(meta_ha, meta_gm)
    meta_combined$Host <- ifelse(grepl("^Ha_", rownames(meta_combined)),
                                  "Sunflower", "Soybean")

    col_annot_z <- data.frame(
      Host          = meta_combined[colnames(zscore_clamped), "Host"],
      Aggressiveness = as.character(
                        meta_combined[colnames(zscore_clamped), "aggressiveness"]),
      Isolate       = as.character(
                        meta_combined[colnames(zscore_clamped), "isolate"]),
      Timepoint     = as.character(
                        meta_combined[colnames(zscore_clamped), "timepoint"]),
      row.names     = colnames(zscore_clamped),
      stringsAsFactors = FALSE)

    # ── Row annotation: n timepoints shared ──────────────────────────────
    row_annot_z <- data.frame(
      Shared_TPs = factor(n_tp_shared[genes_in_both]),
      row.names  = genes_in_both)

    # Short gene labels
    short_ids <- sapply(genes_in_both, function(id) {
      parts <- strsplit(id, "\\|")[[1]]
      if (length(parts) >= 4) return(parts[4])
      if (length(parts) >= 3) return(parts[3])
      return(substr(id, 1, 20))
    })
    rownames(zscore_clamped) <- short_ids
    rownames(row_annot_z)    <- short_ids

    # ── Annotation colors ─────────────────────────────────────────────────
    agg_levels <- intersect(c("Control","Low","High","Soy-specific","Sun-specific"),
                             unique(col_annot_z$Aggressiveness))
    agg_pal <- c("Control"      = "#AAAAAA",
                 "Low"          = "#1f77b4",
                 "High"         = "#d62728",
                 "Soy-specific" = "#8c564b",
                 "Sun-specific" = "#e682fa")

    iso_pal <- c("WISS47"   = "#3ACAFF",
                 "JS659"    = "#00BFB2",
                 "MNSS6"    = "#FF3173",
                 "Xtra7"    = "#FF5F3A",
                 "BN172"    = "#e682fa",
                 "SSPotter" = "#8c564b")
    iso_levels  <- sort(unique(col_annot_z$Isolate))
    iso_colors  <- iso_pal[iso_levels]
    # Fall back for any unexpected isolate names
    missing_iso <- iso_levels[is.na(iso_colors)]
    if (length(missing_iso) > 0) {
      extra <- setNames(
        colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(length(missing_iso)),
        missing_iso)
      iso_colors[missing_iso] <- extra
    }

    tp_levels <- intersect(c("24h","48h","96h"), unique(col_annot_z$Timepoint))
    tp_pal    <- c("24h" = "#FFD700", "48h" = "#FF6347", "96h" = "#8B0000")

    n_tp_lvls <- as.character(sort(unique(n_tp_shared[genes_in_both])))
    sh_pal    <- c("2" = "#FDAE61", "3" = "#D7191C")

    annot_colors_z <- list(
      Host           = c("Sunflower" = "#5e3a04", "Soybean" = "#ffcb21"),
      Aggressiveness = agg_pal[agg_levels],
      Isolate        = iso_colors,
      Timepoint      = tp_pal[tp_levels],
      Shared_TPs     = sh_pal[n_tp_lvls])

    # ── Plot ──────────────────────────────────────────────────────────────
    n_gz    <- nrow(zscore_clamped)
    font_rz <- ifelse(n_gz <= 60, 7, ifelse(n_gz <= 120, 5, 4))

    z_colors <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(100)
    z_breaks <- seq(-2, 2, length.out = 101)

    # NOTE: width/height are intentionally omitted — fig.width/fig.height in
    # the chunk header control output size in R Markdown. Passing width/height
    # to pheatmap() only works when using filename= and will crop the plot here.
    pheatmap(zscore_clamped,
             color             = z_colors,
             breaks            = z_breaks,
             cluster_rows      = TRUE,
             cluster_cols      = FALSE,
             clustering_method = "ward.D2",
             treeheight_row    = 60,
             treeheight_col    = 0,
             annotation_col    = col_annot_z,
             annotation_row    = row_annot_z,
             annotation_colors = annot_colors_z,
             fontsize_row      = font_rz,
             fontsize_col      = 7,
             angle_col         = 90,
             main              = paste0(n_gz,
               " Shared Pathogen DEGs (≥2 Timepoints): Per-Host Row Z-score of VST Expression"),
             border_color      = NA,
             cellwidth         = 12,
             cellheight        = max(6, min(14, 500 / n_gz)),
             legend            = TRUE,
             legend_breaks     = c(-2, -1, 0, 1, 2),
             legend_labels     = c("-2", "-1", "0 (mean)", "1", "2"))

    cat(sprintf("\nHeatmap: %d genes x %d samples\n", n_gz, ncol(zscore_clamped)))
    cat("Color: blue=-2 (below mean), white=0 (mean), red=+2 (above mean); clamped to ±2\n")

  } else {
    cat("Too few genes found in both VST matrices after filtering.\n")
  }

} else if (!exists("genes_2tp") || length(genes_2tp) < 5) {
  cat("Run sections 2-6b first to generate genes_2tp.\n")
} else {
  cat(sprintf("Count file(s) missing. Expected:\n  %s\n  %s\n",
              ha_counts_file, gm_counts_file))
}
```

---

# 6e. Heatmap — ≥2 Shared Timepoints, Z-score VST with Clustered Columns

```{r heatmap-zscore-vst-clustered-cols, fig.width=28, fig.height=14}
# Reuses vst_ha_mat / vst_gm_mat and zscore_clamped built in chunk 6d.
# Identical to 6d except cluster_cols = TRUE and treeheight_col = 50.

if (exists("zscore_clamped") && exists("col_annot_z") &&
    exists("row_annot_z")    && exists("annot_colors_z")) {

  n_gz    <- nrow(zscore_clamped)
  font_rz <- ifelse(n_gz <= 60, 7, ifelse(n_gz <= 120, 5, 4))

  z_colors <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(100)
  z_breaks <- seq(-2, 2, length.out = 101)

  pheatmap(zscore_clamped,
           color             = z_colors,
           breaks            = z_breaks,
           cluster_rows      = TRUE,
           cluster_cols      = TRUE,
           clustering_method = "ward.D2",
           treeheight_row    = 60,
           treeheight_col    = 50,
           annotation_col    = col_annot_z,
           annotation_row    = row_annot_z,
           annotation_colors = annot_colors_z,
           fontsize_row      = font_rz,
           fontsize_col      = 7,
           angle_col         = 90,
           main              = paste0(n_gz,
             " Shared Pathogen DEGs (≥2 Timepoints): Per-Host Row Z-score of VST Expression",
             " — Columns Clustered"),
           border_color      = NA,
           cellwidth         = 12,
           cellheight        = max(6, min(14, 500 / n_gz)),
           legend            = TRUE,
           legend_breaks     = c(-2, -1, 0, 1, 2),
           legend_labels     = c("-2", "-1", "0 (mean)", "1", "2"))

  cat(sprintf("Clustered-column heatmap: %d genes x %d samples\n",
              n_gz, ncol(zscore_clamped)))

} else {
  cat("zscore_clamped not found — run chunk heatmap-zscore-vst (section 6d) first.\n")
}
```

---

# 6f. Heatmap — Columns Ordered by Host → Aggressiveness → Isolate, Rows Clustered

```{r heatmap-zscore-agg-iso-ordered, fig.width=28, fig.height=14}
# Reuses zscore_clamped, col_annot_z, row_annot_z, annot_colors_z from 6d.
# Columns are pre-sorted by Host → Aggressiveness → Isolate → Timepoint.
# Within each isolate, columns are hierarchically clustered.
# gaps_col visually separates aggressiveness groups.

if (exists("zscore_clamped") && exists("col_annot_z") &&
    exists("row_annot_z")    && exists("annot_colors_z")) {

  # ── Define group order ───────────────────────────────────────────────────
  agg_order <- c("Low", "High", "Soy-specific", "Sun-specific", "Control")
  tp_order  <- c("24h", "48h", "96h")

  # Sort columns: Host → Aggressiveness → Isolate → Timepoint
  col_df <- col_annot_z
  col_df$sample <- rownames(col_df)
  col_df$Aggressiveness <- factor(col_df$Aggressiveness,
                                   levels = intersect(agg_order,
                                                      unique(col_df$Aggressiveness)))
  col_df$Timepoint <- factor(col_df$Timepoint,
                              levels = intersect(tp_order, unique(col_df$Timepoint)))
  col_df$Host <- factor(col_df$Host, levels = c("Sunflower", "Soybean"))

  col_df_sorted <- col_df[order(col_df$Host,
                                 col_df$Aggressiveness,
                                 col_df$Isolate,
                                 col_df$Timepoint), ]

  zscore_ordered <- zscore_clamped[, col_df_sorted$sample, drop = FALSE]

  # ── Compute gaps_col: insert gap between each aggressiveness group ────────
  # Gap positions are at the cumulative end of each Host x Aggressiveness block
  group_key <- paste(col_df_sorted$Host, col_df_sorted$Aggressiveness, sep = "_")
  gap_positions <- which(diff(as.integer(factor(group_key,
                                                  levels = unique(group_key)))) != 0)

  cat("Column order (Host x Aggressiveness groups):\n")
  print(table(col_df_sorted$Host, col_df_sorted$Aggressiveness))
  cat(sprintf("Gap positions after columns: %s\n",
              paste(gap_positions, collapse = ", ")))

  # ── Plot ──────────────────────────────────────────────────────────────────
  n_gz    <- nrow(zscore_ordered)
  font_rz <- ifelse(n_gz <= 60, 7, ifelse(n_gz <= 120, 5, 4))

  z_colors <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(100)
  z_breaks <- seq(-2, 2, length.out = 101)

  pheatmap(zscore_ordered,
           color             = z_colors,
           breaks            = z_breaks,
           cluster_rows      = TRUE,
           cluster_cols      = FALSE,
           clustering_method = "ward.D2",
           treeheight_row    = 60,
           treeheight_col    = 0,
           gaps_col          = gap_positions,
           annotation_col    = col_annot_z,
           annotation_row    = row_annot_z,
           annotation_colors = annot_colors_z,
           fontsize_row      = font_rz,
           fontsize_col      = 7,
           angle_col         = 90,
           main              = paste0(n_gz,
             " Shared Pathogen DEGs (≥2 Timepoints): Per-Host Z-score",
             " — Columns Ordered by Aggressiveness & Isolate"),
           border_color      = NA,
           cellwidth         = 12,
           cellheight        = max(6, min(14, 500 / n_gz)),
           legend            = TRUE,
           legend_breaks     = c(-2, -1, 0, 1, 2),
           legend_labels     = c("-2", "-1", "0 (mean)", "1", "2"))

  cat(sprintf("\nHeatmap: %d genes x %d samples | %d aggressiveness-group gaps\n",
              n_gz, ncol(zscore_ordered), length(gap_positions)))

} else {
  cat("zscore_clamped not found — run chunk heatmap-zscore-vst (section 6d) first.\n")
}
```

# 7. Export Shared Significant Genes to CSV

```{r export-shared-genes}
if (length(shared_genes_any) > 0) {

  # ── Build fold-change matrices for ALL shared genes (not just top N) ──────
  build_fc_matrix_export <- function(degs_list, genes, tps) {
    mat <- matrix(NA_real_, nrow = length(genes), ncol = length(tps),
                  dimnames = list(genes, tps))
    for (tp in tps) {
      df <- degs_list[[tp]]
      if (!is.null(df) && lfc_col %in% colnames(df)) {
        shared <- intersect(rownames(df), genes)
        if (length(shared) > 0) mat[shared, tp] <- df[shared, lfc_col]
      }
    }
    return(mat)
  }

  fc_ha_all <- build_fc_matrix_export(ha_degs, shared_genes_any, timepoints)
  fc_gm_all <- build_fc_matrix_export(gm_degs, shared_genes_any, timepoints)

  # ── Collect padj values ───────────────────────────────────────────────────
  # Detect padj column name from first available dataset
  padj_col <- NULL
  for (tp in timepoints) {
    df_check <- ha_degs[[tp]]
    if (!is.null(df_check)) {
      if ("padj" %in% colnames(df_check))       { padj_col <- "padj";  break }
      if ("FDR"  %in% colnames(df_check))        { padj_col <- "FDR";   break }
      if ("adj.P.Val" %in% colnames(df_check))   { padj_col <- "adj.P.Val"; break }
    }
  }

  get_padj_mat <- function(degs_list, genes, tps, col) {
    mat <- matrix(NA_real_, nrow = length(genes), ncol = length(tps),
                  dimnames = list(genes, tps))
    if (is.null(col)) return(mat)
    for (tp in tps) {
      df <- degs_list[[tp]]
      if (!is.null(df) && col %in% colnames(df)) {
        shared <- intersect(rownames(df), genes)
        if (length(shared) > 0) mat[shared, tp] <- df[shared, col]
      }
    }
    return(mat)
  }

  padj_ha_all <- get_padj_mat(ha_degs, shared_genes_any, timepoints, padj_col)
  padj_gm_all <- get_padj_mat(gm_degs, shared_genes_any, timepoints, padj_col)

  # ── Per-timepoint presence flags ─────────────────────────────────────────
  export_df <- data.frame(
    gene_id = shared_genes_any,
    row.names = shared_genes_any,
    stringsAsFactors = FALSE
  )

  for (tp in timepoints) {
    export_df[[paste0("Ha_DEG_", tp)]] <- shared_genes_any %in% ha_gene_sets[[tp]]
    export_df[[paste0("Gm_DEG_", tp)]] <- shared_genes_any %in% gm_gene_sets[[tp]]
  }

  # Number of timepoints where gene is shared in BOTH hosts simultaneously
  tps_both_export <- intersect(ha_available, gm_available)
  export_df$n_timepoints_shared <- sapply(shared_genes_any, function(g) {
    sum(sapply(tps_both_export,
               function(tp) g %in% ha_gene_sets[[tp]] && g %in% gm_gene_sets[[tp]]))
  })

  # ── log2FC and padj columns ───────────────────────────────────────────────
  for (tp in timepoints) {
    export_df[[paste0("Ha_log2FC_", tp)]] <- fc_ha_all[shared_genes_any, tp]
    export_df[[paste0("Gm_log2FC_", tp)]] <- fc_gm_all[shared_genes_any, tp]
    if (!is.null(padj_col)) {
      export_df[[paste0("Ha_padj_", tp)]] <- padj_ha_all[shared_genes_any, tp]
      export_df[[paste0("Gm_padj_", tp)]] <- padj_gm_all[shared_genes_any, tp]
    }
  }

  # Max absolute FC across all conditions (useful for downstream ranking)
  lfc_cols <- grep("_log2FC_", colnames(export_df), value = TRUE)
  export_df$max_abs_log2FC <- apply(
    abs(export_df[, lfc_cols, drop = FALSE]), 1, max, na.rm = TRUE)

  # Sort: most-shared timepoints first, then by strongest fold change
  export_df <- export_df[order(-export_df$n_timepoints_shared,
                                -export_df$max_abs_log2FC), ]

  # ── Write CSV ─────────────────────────────────────────────────────────────
  out_file <- "shared_significant_genes.csv"
  write.csv(export_df, file = out_file, row.names = FALSE)

  cat(sprintf("Exported %d shared significant genes to: %s\n",
              nrow(export_df), out_file))
  cat(sprintf("Columns: %s\n\n", paste(colnames(export_df), collapse = ", ")))

  # Preview
  preview_cols <- c("gene_id", "n_timepoints_shared", "max_abs_log2FC",
                    grep("log2FC", colnames(export_df), value = TRUE))
  cat("Preview (first 6 rows):\n")
  print(head(export_df[, preview_cols], 6), digits = 3)

} else {
  cat("No shared significant genes found — nothing to export.\n")
}
```

---

# 8. Top 20 Shared Significant Genes — CSV Export

```{r export-top20-genes}
if (length(shared_genes_any) > 0 && exists("export_df") && nrow(export_df) >= 1) {

  n_top20 <- min(20, nrow(export_df))

  # export_df is already sorted by n_timepoints_shared then max_abs_log2FC
  top20_df <- head(export_df, n_top20)

  out_file_top20 <- "top20_shared_significant_genes.csv"
  write.csv(top20_df, file = out_file_top20, row.names = FALSE)

  cat(sprintf("Exported top %d shared significant genes to: %s\n",
              n_top20, out_file_top20))
  cat(sprintf("Ranked by: n_timepoints_shared (desc) then max_abs_log2FC (desc)\n\n"))

  preview_cols <- c("gene_id", "n_timepoints_shared", "max_abs_log2FC",
                    grep("log2FC", colnames(top20_df), value = TRUE))
  print(top20_df[, preview_cols], digits = 3)

} else {
  cat("No data available for top 20 export — run the export-shared-genes chunk first.\n")
}
```

---

# 9. Summary  

```{r final-summary}
cat("\n=== CROSS-HOST VISUALIZATION SUMMARY ===\n\n")

cat("-- Sunflower (Ha89) Pathogen DEGs --\n")
for (tp in timepoints) {
  n <- length(ha_gene_sets[[tp]])
  if (n > 0) cat(sprintf("  %s: %d DEGs (Up:%d Down:%d)\n", tp, n,
                          length(ha_up_sets[[tp]]), length(ha_down_sets[[tp]])))
  else cat(sprintf("  %s: NO DATA\n", tp))
}

cat("\n-- Soybean (Wm82) Pathogen DEGs --\n")
for (tp in timepoints) {
  n <- length(gm_gene_sets[[tp]])
  if (n > 0) cat(sprintf("  %s: %d DEGs (Up:%d Down:%d)\n", tp, n,
                          length(gm_up_sets[[tp]]), length(gm_down_sets[[tp]])))
  else cat(sprintf("  %s: NO DATA\n", tp))
}

cat("\n-- Cross-Host --\n")
if (exists("shared_hosts")) {
  cat(sprintf("  Shared DEGs (any timepoint): %d\n", length(shared_hosts)))
  for (tp in intersect(ha_available, gm_available)) {
    cat(sprintf("  Shared at %s: %d\n", tp,
                length(intersect(ha_gene_sets[[tp]], gm_gene_sets[[tp]]))))
  }
}

cat("\nDone!\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
