---
title: "Cross-Host Visualization: Venn Diagrams, UpSet Plot & Tanglegram"
subtitle: "Sclerotinia Pathogen DEGs — Sunflower (Ha) vs Soybean (Gm)"
author: "Analysis by wan00965"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 8, dpi = 300)
```

# 1. Setup

## Install Required Packages (Run Once)

```{r install-packages, eval=FALSE}
install.packages(c("VennDiagram", "UpSetR", "dendextend", "ggplot2",
                    "dplyr", "RColorBrewer", "gridExtra", "eulerr", "pheatmap"))
```

## Load Libraries

```{r load-libraries}
library(VennDiagram)
library(UpSetR)
library(dendextend)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(gridExtra)
library(grid)
library(eulerr)
library(pheatmap)

cat("All libraries loaded successfully!\n")
```

## Set Working Directory

```{r set-wd}
# UPDATE this path to your data folder
setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_data/New_salmon_round3/new_run_count")
cat("Working directory:", getwd(), "\n")
```

# 2. Load DEG Data from Both Hosts

```{r load-degs}
timepoints <- c("24h", "48h", "96h")

# Helper: safely load a DEG file, return NULL if not found
safe_load_deg <- function(filepath) {
  if (file.exists(filepath)) {
    df <- read.csv(filepath, stringsAsFactors = FALSE, row.names = 1)
    if (nrow(df) > 0) return(df)
  }
  return(NULL)
}

# --- Sunflower (Ha) pathogen DEGs ---
ha_degs <- list()
for (tp in timepoints) {
  f <- sprintf("Ha_pathogen_aggressive_vs_low_DEGs_%s.csv", tp)
  ha_degs[[tp]] <- safe_load_deg(f)
  if (!is.null(ha_degs[[tp]])) {
    cat(sprintf("Ha %s: %d DEGs loaded\n", tp, nrow(ha_degs[[tp]])))
  } else {
    cat(sprintf("WARNING: Ha %s not found (%s)\n", tp, f))
  }
}

# --- Soybean (Gm) pathogen DEGs ---
gm_degs <- list()
for (tp in timepoints) {
  f <- sprintf("Gm_pathogen_aggressive_vs_low_DEGs_%s.csv", tp)
  gm_degs[[tp]] <- safe_load_deg(f)
  if (!is.null(gm_degs[[tp]])) {
    cat(sprintf("Gm %s: %d DEGs loaded\n", tp, nrow(gm_degs[[tp]])))
  } else {
    cat(sprintf("WARNING: Gm %s not found (%s)\n", tp, f))
  }
}

# --- Show available CSV files for debugging ---
cat("\n=== DEG CSV files in working directory ===\n")
csv_files <- list.files(pattern = "aggressive_vs_low.*\\.csv$", ignore.case = TRUE)
if (length(csv_files) > 0) {
  print(csv_files)
} else {
  cat("No matching CSV files found! Check filenames.\n")
  cat("Expected pattern: Ha_pathogen_aggressive_vs_low_DEGs_24h.csv\n")
  cat("All CSV files:\n")
  print(list.files(pattern = "\\.csv$"))
}
```

## Extract Gene IDs (with safe fallbacks)

```{r extract-gene-ids}
lfc_col <- "log2FoldChange"

# Safe extraction: returns character(0) if data is NULL or empty
safe_rownames <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(character(0))
  return(rownames(df))
}

safe_filter_rows <- function(df, direction = "up") {
  if (is.null(df) || nrow(df) == 0) return(character(0))
  if (!(lfc_col %in% colnames(df))) return(character(0))
  if (direction == "up") {
    return(rownames(df[df[[lfc_col]] > 0, ]))
  } else {
    return(rownames(df[df[[lfc_col]] < 0, ]))
  }
}

# Build all gene sets with safe fallbacks
ha_gene_sets <- list(); gm_gene_sets <- list()
ha_up_sets <- list(); ha_down_sets <- list()
gm_up_sets <- list(); gm_down_sets <- list()

for (tp in timepoints) {
  ha_gene_sets[[tp]] <- safe_rownames(ha_degs[[tp]])
  gm_gene_sets[[tp]] <- safe_rownames(gm_degs[[tp]])
  ha_up_sets[[tp]]   <- safe_filter_rows(ha_degs[[tp]], "up")
  ha_down_sets[[tp]] <- safe_filter_rows(ha_degs[[tp]], "down")
  gm_up_sets[[tp]]   <- safe_filter_rows(gm_degs[[tp]], "up")
  gm_down_sets[[tp]] <- safe_filter_rows(gm_degs[[tp]], "down")
}

# Which timepoints actually have data?
ha_available <- timepoints[sapply(timepoints, function(tp) length(ha_gene_sets[[tp]]) > 0)]
gm_available <- timepoints[sapply(timepoints, function(tp) length(gm_gene_sets[[tp]]) > 0)]

cat("\n=== DATA AVAILABILITY ===\n")
cat("Ha timepoints with data:", if (length(ha_available) > 0) paste(ha_available, collapse=", ") else "NONE", "\n")
cat("Gm timepoints with data:", if (length(gm_available) > 0) paste(gm_available, collapse=", ") else "NONE", "\n")

cat("\n=== GENE SET SIZES ===\n")
for (tp in timepoints) {
  cat(sprintf("%s  Ha: %d (Up:%d Down:%d)  |  Gm: %d (Up:%d Down:%d)\n", tp,
              length(ha_gene_sets[[tp]]), length(ha_up_sets[[tp]]), length(ha_down_sets[[tp]]),
              length(gm_gene_sets[[tp]]), length(gm_up_sets[[tp]]), length(gm_down_sets[[tp]])))
}
```

---

# 3. Venn Diagrams — DEGs Across 3 Timepoints

```{r venn-helper}
# Helper function to compute all 7 regions of a 3-set Venn
compute_venn3_stats <- function(set_a, set_b, set_c, names = c("A","B","C")) {
  all_genes <- unique(c(set_a, set_b, set_c))
  n_total <- length(all_genes)
  
  only_a  <- setdiff(set_a, union(set_b, set_c))
  only_b  <- setdiff(set_b, union(set_a, set_c))
  only_c  <- setdiff(set_c, union(set_a, set_b))
  only_ab <- setdiff(intersect(set_a, set_b), set_c)
  only_ac <- setdiff(intersect(set_a, set_c), set_b)
  only_bc <- setdiff(intersect(set_b, set_c), set_a)
  abc     <- Reduce(intersect, list(set_a, set_b, set_c))
  
  counts <- c(length(only_a), length(only_b), length(only_c),
              length(only_ab), length(only_ac), length(only_bc), length(abc))
  pcts <- if (n_total > 0) round(100 * counts / n_total, 1) else rep(0, 7)
  
  region_names <- c(paste0("Only ",names[1]), paste0("Only ",names[2]), paste0("Only ",names[3]),
                    paste0(names[1]," & ",names[2]," only"),
                    paste0(names[1]," & ",names[3]," only"),
                    paste0(names[2]," & ",names[3]," only"),
                    "All three")
  
  euler_input <- setNames(counts, c(
    names[1], names[2], names[3],
    paste0(names[1],"&",names[2]),
    paste0(names[1],"&",names[3]),
    paste0(names[2],"&",names[3]),
    paste0(names[1],"&",names[2],"&",names[3])
  ))
  
  list(counts=counts, pcts=pcts, n_total=n_total,
       region_names=region_names, euler_input=euler_input)
}
```

## 3.1 Sunflower (Ha) Pathogen DEGs — 3-Way Venn

```{r venn-ha-eulerr, fig.width=9, fig.height=8}
venn_colors <- c("#6495ED", "#FFD700", "#32CD32")
venn_fills  <- adjustcolor(venn_colors, alpha.f = 0.45)

if (length(ha_available) >= 3) {
  
  ha_stats <- compute_venn3_stats(
    ha_gene_sets[["24h"]], ha_gene_sets[["48h"]], ha_gene_sets[["96h"]],
    names = c("24 HPI", "48 HPI", "96 HPI"))
  
  cat("=== SUNFLOWER (Ha) VENN SUMMARY ===\n")
  cat(sprintf("Total unique pathogen DEGs: %d\n\n", ha_stats$n_total))
  for (i in seq_along(ha_stats$counts)) {
    cat(sprintf("  %s: %d (%.1f%%)\n", ha_stats$region_names[i],
                ha_stats$counts[i], ha_stats$pcts[i]))
  }
  
  fit_ha <- euler(ha_stats$euler_input)
  plot(fit_ha,
       fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.3, font = 2, col = venn_colors),
       quantities = list(cex = 1.2, font = 2, type = c("counts", "percent")),
       main = list(label = "Pathogen DEGs in Ha89 (Aggressive vs Low Isolates)",
                   cex = 1.5, font = 2))
  
} else if (length(ha_available) == 2) {
  cat("Only 2 Ha timepoints with data — drawing 2-set Venn\n")
  tp1 <- ha_available[1]; tp2 <- ha_available[2]
  fit_ha <- euler(c(
    setNames(length(setdiff(ha_gene_sets[[tp1]], ha_gene_sets[[tp2]])), tp1),
    setNames(length(setdiff(ha_gene_sets[[tp2]], ha_gene_sets[[tp1]])), tp2),
    setNames(length(intersect(ha_gene_sets[[tp1]], ha_gene_sets[[tp2]])),
             paste0(tp1,"&",tp2))))
  plot(fit_ha, fills = list(fill = venn_fills[1:2], alpha = 0.7),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Sunflower (Ha89)", cex = 1.5, font = 2))
} else {
  cat("Fewer than 2 Ha timepoints — cannot draw Venn.\n")
  cat("Available files:\n")
  print(list.files(pattern = "Ha_pathogen"))
}
```

```{r venn-ha-classic, fig.width=8, fig.height=8}
# Classic VennDiagram package (non-proportional circles, like your Fig. 5)
if (length(ha_available) >= 3) {
  
  futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
  
  venn.plot.ha <- venn.diagram(
    x = list("24 HPI" = ha_gene_sets[["24h"]],
             "48 HPI" = ha_gene_sets[["48h"]],
             "96 HPI" = ha_gene_sets[["96h"]]),
    filename = NULL,
    fill = venn_fills, col = venn_colors, lwd = 2.5,
    category.names = c("24 HPI","48 HPI","96 HPI"),
    cat.cex = 1.8, cat.fontface = "bold", cat.col = venn_colors,
    cat.dist = c(0.06, 0.06, 0.06),
    cex = 1.6, fontface = "bold", fontfamily = "sans",
    main = "Pathogen DEGs in Sunflower (Ha89)",
    main.cex = 2, main.fontface = "bold",
    euler.d = FALSE, scaled = FALSE, margin = 0.1)
  
  grid.newpage()
  grid.draw(venn.plot.ha)
}
```

## 3.2 Soybean (Gm) Pathogen DEGs — 3-Way Venn

```{r venn-gm-eulerr, fig.width=9, fig.height=8}
if (length(gm_available) >= 3) {
  
  gm_stats <- compute_venn3_stats(
    gm_gene_sets[["24h"]], gm_gene_sets[["48h"]], gm_gene_sets[["96h"]],
    names = c("24 HPI", "48 HPI", "96 HPI"))
  
  cat("=== SOYBEAN (Gm) VENN SUMMARY ===\n")
  cat(sprintf("Total unique pathogen DEGs: %d\n\n", gm_stats$n_total))
  for (i in seq_along(gm_stats$counts)) {
    cat(sprintf("  %s: %d (%.1f%%)\n", gm_stats$region_names[i],
                gm_stats$counts[i], gm_stats$pcts[i]))
  }
  
  fit_gm <- euler(gm_stats$euler_input)
  plot(fit_gm,
       fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.3, font = 2, col = venn_colors),
       quantities = list(cex = 1.2, font = 2, type = c("counts", "percent")),
       main = list(label = "Pathogen DEGs in Williams 82 (Aggressive vs Low Isolates)",
                   cex = 1.5, font = 2))
  
} else if (length(gm_available) == 2) {
  cat("Only 2 Gm timepoints with data — drawing 2-set Venn\n")
  tp1 <- gm_available[1]; tp2 <- gm_available[2]
  fit_gm <- euler(c(
    setNames(length(setdiff(gm_gene_sets[[tp1]], gm_gene_sets[[tp2]])), tp1),
    setNames(length(setdiff(gm_gene_sets[[tp2]], gm_gene_sets[[tp1]])), tp2),
    setNames(length(intersect(gm_gene_sets[[tp1]], gm_gene_sets[[tp2]])),
             paste0(tp1,"&",tp2))))
  plot(fit_gm, fills = list(fill = venn_fills[1:2], alpha = 0.7),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Soybean (Wm82)", cex = 1.5, font = 2))
} else {
  cat("Fewer than 2 Gm timepoints — cannot draw Venn.\n")
  cat("Available files:\n")
  print(list.files(pattern = "Gm_pathogen"))
}
```

```{r venn-gm-classic, fig.width=8, fig.height=8}
if (length(gm_available) >= 3) {
  
  futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
  
  venn.plot.gm <- venn.diagram(
    x = list("24 HPI" = gm_gene_sets[["24h"]],
             "48 HPI" = gm_gene_sets[["48h"]],
             "96 HPI" = gm_gene_sets[["96h"]]),
    filename = NULL,
    fill = venn_fills, col = venn_colors, lwd = 2.5,
    category.names = c("24 HPI","48 HPI","96 HPI"),
    cat.cex = 1.8, cat.fontface = "bold", cat.col = venn_colors,
    cat.dist = c(0.06, 0.06, 0.06),
    cex = 1.6, fontface = "bold", fontfamily = "sans",
    main = "Pathogen DEGs in Soybean (Williams 82)",
    main.cex = 2, main.fontface = "bold",
    euler.d = FALSE, scaled = FALSE, margin = 0.1)
  
  grid.newpage()
  grid.draw(venn.plot.gm)
}
```

## 3.3 Save Venns as PDF

```{r save-venns, eval=FALSE}
if (exists("fit_ha")) {
  pdf("Fig_Venn_Ha_pathogen_DEGs.pdf", width = 8, height = 8)
  plot(fit_ha, fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.2, font = 2),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Sunflower (Ha89)", cex = 1.5, font = 2))
  dev.off(); cat("Saved: Fig_Venn_Ha_pathogen_DEGs.pdf\n")
}
if (exists("fit_gm")) {
  pdf("Fig_Venn_Gm_pathogen_DEGs.pdf", width = 8, height = 8)
  plot(fit_gm, fills = list(fill = venn_fills, alpha = 0.7),
       edges = list(col = venn_colors, lwd = 2.5),
       labels = list(cex = 1.2, font = 2),
       quantities = list(cex = 1.3, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs in Soybean (Wm82)", cex = 1.5, font = 2))
  dev.off(); cat("Saved: Fig_Venn_Gm_pathogen_DEGs.pdf\n")
}
```

---

# 4. UpSet Plot — Cross-Host DEG Overlap

```{r upset-cross-host, fig.width=14, fig.height=8}
# Build named list of all available sets
all_sets <- list()
for (tp in ha_available) all_sets[[paste0("Ha_",tp)]] <- ha_gene_sets[[tp]]
for (tp in gm_available) all_sets[[paste0("Gm_",tp)]] <- gm_gene_sets[[tp]]

if (length(all_sets) >= 2) {
  
  all_genes <- unique(unlist(all_sets))
  upset_df <- data.frame(row.names = all_genes)
  for (nm in names(all_sets)) {
    upset_df[[nm]] <- as.integer(all_genes %in% all_sets[[nm]])
  }
  
  # ---------------------------------------------------------------
  # NOTE: UpSetR's intersects() query is fragile — it crashes if
  # the highlighted intersection doesn't appear among the top N
  # displayed bars. We use a safe approach: color bars by set
  # membership using the "sets.bar.color" parameter instead,
  # and skip the problematic queries entirely.
  # ---------------------------------------------------------------
  
  # Color the set size bars: orange for Ha, blue for Gm
  set_names_ordered <- rev(names(all_sets))
  bar_colors <- ifelse(grepl("^Ha_", set_names_ordered), "#E69F00", "#56B4E9")
  
  upset(upset_df,
        sets = set_names_ordered,
        order.by = "freq",
        nsets = length(all_sets),
        nintersects = 30,
        mb.ratio = c(0.6, 0.4),
        point.size = 3.5,
        line.size = 1.5,
        sets.bar.color = bar_colors,
        mainbar.y.label = "Intersection Size",
        sets.x.label = "Number of DEGs",
        text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
        set_size.show = TRUE)
  
  # Print cross-host overlap stats for the UpSet context
  cat("\n=== UPSET INTERSECTION SUMMARY ===\n")
  cat(sprintf("Total unique pathogen DEGs across all conditions: %d\n", length(all_genes)))
  cat(sprintf("Sets included: %s\n", paste(names(all_sets), collapse = ", ")))
  
  # Show top intersections manually
  # Each gene's membership pattern as a string
  patterns <- apply(upset_df[, names(all_sets), drop = FALSE], 1,
                    function(row) paste(names(all_sets)[row == 1], collapse = " + "))
  pattern_counts <- sort(table(patterns), decreasing = TRUE)
  cat("\nTop 10 intersection patterns:\n")
  print(head(pattern_counts, 10))
  
} else {
  cat("Need at least 2 gene sets for UpSet plot.\n")
}
```

## 4.2 Cross-Host Venn: Ha vs Gm (All Timepoints Combined)

```{r venn-2host, fig.width=9, fig.height=7}
ha_all_genes <- unique(unlist(ha_gene_sets))
gm_all_genes <- unique(unlist(gm_gene_sets))

if (length(ha_all_genes) > 0 && length(gm_all_genes) > 0) {
  
  shared_hosts <- intersect(ha_all_genes, gm_all_genes)
  ha_only <- setdiff(ha_all_genes, gm_all_genes)
  gm_only <- setdiff(gm_all_genes, ha_all_genes)
  
  cat("\n=== CROSS-HOST OVERLAP (All Timepoints Combined) ===\n")
  cat(sprintf("Ha-only pathogen DEGs: %d\n", length(ha_only)))
  cat(sprintf("Gm-only pathogen DEGs: %d\n", length(gm_only)))
  cat(sprintf("Shared between hosts:  %d\n", length(shared_hosts)))
  
  venn_2host <- euler(c(
    "Sunflower\n(Ha89)" = length(ha_only),
    "Soybean\n(Wm82)"   = length(gm_only),
    "Sunflower\n(Ha89)&Soybean\n(Wm82)" = length(shared_hosts)))
  
  plot(venn_2host,
       fills = list(fill = c("#FF9933","#6699CC"), alpha = 0.6),
       edges = list(col = c("#CC6600","#336699"), lwd = 3),
       labels = list(cex = 1.4, font = 2),
       quantities = list(cex = 1.5, font = 2, type = c("counts","percent")),
       main = list(label = "Pathogen DEGs: Sunflower vs Soybean (All Timepoints Combined)",
                   cex = 1.5, font = 2))
} else {
  cat("Need data from both hosts for cross-host comparison.\n")
}
```

## 4.3 Per-Timepoint Cross-Host Comparison

```{r venn-per-tp, fig.width=15, fig.height=5}
both_available <- intersect(ha_available, gm_available)

if (length(both_available) > 0) {
  
  tp_summaries <- data.frame()
  plots_per_tp <- list()
  
  for (tp in both_available) {
    shared_tp <- intersect(ha_gene_sets[[tp]], gm_gene_sets[[tp]])
    ha_only_tp <- setdiff(ha_gene_sets[[tp]], gm_gene_sets[[tp]])
    gm_only_tp <- setdiff(gm_gene_sets[[tp]], ha_gene_sets[[tp]])
    total_union <- length(union(ha_gene_sets[[tp]], gm_gene_sets[[tp]]))
    
    tp_summaries <- rbind(tp_summaries, data.frame(
      Timepoint = tp,
      Ha_only = length(ha_only_tp), Shared = length(shared_tp),
      Gm_only = length(gm_only_tp),
      Ha_total = length(ha_gene_sets[[tp]]), Gm_total = length(gm_gene_sets[[tp]]),
      Overlap_pct = if(total_union>0) round(100*length(shared_tp)/total_union,1) else 0))
    
    fit_tp <- euler(c("Ha"=length(ha_only_tp), "Gm"=length(gm_only_tp),
                       "Ha&Gm"=length(shared_tp)))
    plots_per_tp[[tp]] <- plot(fit_tp,
         fills = list(fill = c("#FF9933","#6699CC"), alpha = 0.6),
         edges = list(col = c("#CC6600","#336699"), lwd = 2.5),
         labels = list(cex = 1.1, font = 2),
         quantities = list(cex = 1.2, font = 2, type = c("counts","percent")),
         main = list(label = tp, cex = 1.4, font = 2))
  }
  
  cat("\n=== PER-TIMEPOINT CROSS-HOST OVERLAP ===\n")
  print(tp_summaries)
  
  gridExtra::grid.arrange(
    grobs = plots_per_tp, ncol = length(both_available),
    top = textGrob("Pathogen DEGs: Sunflower vs Soybean per Timepoint",
                   gp = gpar(fontsize = 16, fontface = "bold")))
} else {
  cat("No shared timepoints between hosts.\n")
}
```

---

# 5. Tanglegram — Comparing DEG Profiles Between Hosts

```{r tanglegram, fig.width=14, fig.height=10}
shared_genes_any <- if (length(ha_all_genes) > 0 && length(gm_all_genes) > 0) {
  intersect(ha_all_genes, gm_all_genes)
} else { character(0) }

tps_both <- intersect(ha_available, gm_available)

cat(sprintf("Shared DEGs: %d genes | Shared timepoints: %s\n",
            length(shared_genes_any), paste(tps_both, collapse=", ")))

if (length(shared_genes_any) >= 20 && length(tps_both) >= 2) {
  
  # Build fold-change matrices
  build_fc_matrix <- function(degs_list, genes, tps) {
    mat <- matrix(0, nrow=length(genes), ncol=length(tps),
                  dimnames=list(genes, tps))
    for (tp in tps) {
      df <- degs_list[[tp]]
      if (!is.null(df) && lfc_col %in% colnames(df)) {
        shared <- intersect(rownames(df), genes)
        if (length(shared) > 0) mat[shared, tp] <- df[shared, lfc_col]
      }
    }
    return(mat)
  }
  
  fc_ha <- build_fc_matrix(ha_degs, shared_genes_any, tps_both)
  fc_gm <- build_fc_matrix(gm_degs, shared_genes_any, tps_both)
  
  # Select top N genes by max absolute fold change
  max_fc <- pmax(apply(abs(fc_ha), 1, max, na.rm=TRUE),
                  apply(abs(fc_gm), 1, max, na.rm=TRUE))
  
  n_top <- min(50, length(shared_genes_any))
  top_genes <- names(sort(max_fc, decreasing=TRUE))[1:n_top]
  
  fc_ha_top <- fc_ha[top_genes, , drop=FALSE]
  fc_gm_top <- fc_gm[top_genes, , drop=FALSE]
  
  # Short gene labels
  short_names <- sapply(top_genes, function(id) {
    parts <- strsplit(id, "\\|")[[1]]
    if (length(parts) >= 4) return(parts[4])
    if (length(parts) >= 3) return(parts[3])
    return(substr(id, 1, 20))
  })
  rownames(fc_ha_top) <- short_names
  rownames(fc_gm_top) <- short_names
  
  # Cluster and color
  dend_ha <- as.dendrogram(hclust(dist(fc_ha_top), method="ward.D2"))
  dend_gm <- as.dendrogram(hclust(dist(fc_gm_top), method="ward.D2"))
  
  k_cl <- min(4, max(2, n_top %/% 8))
  cols <- c("#E69F00","#56B4E9","#009E73","#CC79A7")
  dend_ha <- color_branches(dend_ha, k=k_cl, col=cols[1:k_cl])
  dend_gm <- color_branches(dend_gm, k=k_cl, col=cols[1:k_cl])
  
  dl <- dendlist(dend_ha, dend_gm) %>% untangle(method="step2side")
  
  tanglegram(dl,
             main_left = "Sunflower (Ha89)",
             main_right = "Soybean (Wm82)",
             common_subtrees_color_branches = TRUE,
             highlight_distinct_edges = TRUE,
             highlight_branches_lwd = FALSE,
             margin_inner = 8, lwd = 1.8,
             columns_width = c(5, 3, 5),
             cex_main = 1.5, cex_main_left = 1.4, cex_main_right = 1.4,
             lab.cex = 0.7,
             main = paste0("Top ", n_top, " Shared Pathogen DEGs: Sunflower vs Soybean"))
  
  cat(sprintf("\nEntanglement: %.3f (0=identical, 1=max crossing)\n", entanglement(dl)))
  
} else {
  cat("Not enough shared genes or timepoints for tanglegram.\n")
}
```

---

# 6. Heatmap — Cross-Host Fold Change Comparison

```{r heatmap, fig.width=12, fig.height=10}
if (exists("fc_ha") && exists("fc_gm") && length(shared_genes_any) >= 10) {
  
  n_heat <- min(60, length(shared_genes_any))
  top_heat <- names(sort(max_fc, decreasing=TRUE))[1:n_heat]
  
  fc_combined <- cbind(fc_ha[top_heat,,drop=FALSE], fc_gm[top_heat,,drop=FALSE])
  colnames(fc_combined) <- c(paste0("Ha_",tps_both), paste0("Gm_",tps_both))
  
  rownames(fc_combined) <- sapply(top_heat, function(id) {
    parts <- strsplit(id, "\\|")[[1]]
    if (length(parts) >= 4) return(parts[4])
    if (length(parts) >= 3) return(parts[3])
    return(substr(id, 1, 20))
  })
  
  col_annot <- data.frame(
    Host = rep(c("Sunflower","Soybean"), each=length(tps_both)),
    Timepoint = rep(tps_both, 2),
    row.names = colnames(fc_combined))
  
  annot_colors <- list(
    Host = c("Sunflower"="#FF9933", "Soybean"="#6699CC"),
    Timepoint = c("24h"="#66C2A5", "48h"="#FC8D62", "96h"="#8DA0CB")[tps_both])
  
  max_abs <- max(abs(fc_combined), na.rm=TRUE)
  if (max_abs == 0) max_abs <- 1
  
  pheatmap(fc_combined,
           color = colorRampPalette(c("#2166AC","white","#B2182B"))(100),
           breaks = seq(-max_abs, max_abs, length.out=101),
           cluster_cols = FALSE, clustering_method = "ward.D2",
           annotation_col = col_annot, annotation_colors = annot_colors,
           fontsize_row = 7, fontsize_col = 11,
           main = paste0("Top ",n_heat," Shared Pathogen DEGs: log2FC (Aggressive vs Low)"),
           border_color = NA, cellwidth = 35)
} else {
  cat("Not enough shared data for heatmap.\n")
}
```

---

# 7. Summary

```{r final-summary}
cat("\n=== CROSS-HOST VISUALIZATION SUMMARY ===\n\n")

cat("-- Sunflower (Ha89) Pathogen DEGs --\n")
for (tp in timepoints) {
  n <- length(ha_gene_sets[[tp]])
  if (n > 0) cat(sprintf("  %s: %d DEGs (Up:%d Down:%d)\n", tp, n,
                          length(ha_up_sets[[tp]]), length(ha_down_sets[[tp]])))
  else cat(sprintf("  %s: NO DATA\n", tp))
}

cat("\n-- Soybean (Wm82) Pathogen DEGs --\n")
for (tp in timepoints) {
  n <- length(gm_gene_sets[[tp]])
  if (n > 0) cat(sprintf("  %s: %d DEGs (Up:%d Down:%d)\n", tp, n,
                          length(gm_up_sets[[tp]]), length(gm_down_sets[[tp]])))
  else cat(sprintf("  %s: NO DATA\n", tp))
}

cat("\n-- Cross-Host --\n")
if (exists("shared_hosts")) {
  cat(sprintf("  Shared DEGs (any timepoint): %d\n", length(shared_hosts)))
  for (tp in intersect(ha_available, gm_available)) {
    cat(sprintf("  Shared at %s: %d\n", tp,
                length(intersect(ha_gene_sets[[tp]], gm_gene_sets[[tp]]))))
  }
}

cat("\nDone!\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
