---
title: "Host GO Enrichment Analysis - All Timepoints (Based on Proven Approach)"
author: "RNA-seq Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Overview

This analysis adapts your **proven successful approach** to analyze host GO enrichment across all timepoints (24h, 48h, 96h) using the same gene ID fixes and annotation files that worked in your previous analysis.

**Key elements from your successful analysis:**
- ✅ Gene ID fix: Remove `Ha_`/`Gm_` prefix + `.1/.2/.3` transcript suffix
- ✅ Phytozome annotation files with `locusName` and `GO` columns
- ✅ Coverage verification after ID fix
- ✅ `enricher()` function with `term2gene` mapping

---

# 1. Setup and Load Libraries

```{r load_libraries}
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(pheatmap)
library(RColorBrewer)
library(cowplot)

theme_set(theme_bw(base_size = 12))
cat("✓ Libraries loaded successfully\n")
```

```{r configuration}
# ===== CONFIGURATION - UPDATE THESE PATHS =====

# Choose organism: "Ha" for Helianthus, "Gm" for Glycine max
ORGANISM <- "Gm"  # Change to "Gm" for soybean analysis

# Set working directory - UPDATE THIS PATH
setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_data/New_salmon_round3/new_run_count")

# Define file paths based on organism
if(ORGANISM == "Ha") {
  organism_name <- "Helianthus annuus"
  deg_file_pattern <- "./Ha_host_aggressive_vs_low_DEGs_%s.csv"
  annotation_file <- "./Hannuus_494_r1.2.P14.annotation_info.txt"
  output_prefix <- "Ha_host"
} else if(ORGANISM == "Gm") {
  organism_name <- "Glycine max"
  deg_file_pattern <- "./Gm_host_aggressive_vs_low_DEGs_%s.csv"
  annotation_file <- "./Gmax_880_Wm82.a6.v1.P14.annotation_info.txt"
  output_prefix <- "Gm_host"
}

timepoints <- c("24h", "48h", "96h")

cat("=== ANALYSIS CONFIGURATION ===\n")
cat("Organism:", organism_name, "\n")
cat("Annotation file:", annotation_file, "\n")
cat("Timepoints:", paste(timepoints, collapse = ", "), "\n\n")

# Create output directories
dir.create(paste0(output_prefix, "_enrichment_results"), showWarnings = FALSE)
dir.create("plots", showWarnings = FALSE)
```

---

# 2. Load and Check Annotation (Your Proven Approach)

```{r load_annotation}
cat("=== LOADING ANNOTATION (PROVEN APPROACH) ===\n\n")

if(file.exists(annotation_file)) {
  cat("✓ Loading Phytozome annotation...\n")
  
  # Read annotation using your proven approach
  annot <- read.delim(annotation_file, header = TRUE, stringsAsFactors = FALSE)
  
  cat("File dimensions:\n")
  cat("  Rows:", nrow(annot), "\n")
  cat("  Columns:", ncol(annot), "\n")
  cat("  Column names:", paste(colnames(annot), collapse = ", "), "\n\n")
  
  # Extract GO terms using your exact approach
  go_mapping <- annot %>%
    select(locusName, GO) %>%
    filter(!is.na(GO) & GO != "" & GO != "none") %>%
    separate_rows(GO, sep = ",") %>%
    mutate(GO = trimws(GO)) %>%
    filter(grepl("^GO:", GO))
  
  # Create term2gene dataframe
  term2gene <- go_mapping %>%
    select(GO, locusName) %>%
    distinct()
  
  cat("✓ Annotation processing complete:\n")
  cat("  Genes with GO annotation:", length(unique(term2gene$locusName)), "\n")
  cat("  Total gene-GO pairs:", nrow(term2gene), "\n\n")
  
  annotation_available <- TRUE
  
} else {
  cat("❌ Annotation file not found:", annotation_file, "\n")
  cat("Please ensure the Phytozome annotation file is in the working directory.\n")
  
  annotation_available <- FALSE
  term2gene <- data.frame(GO = character(), locusName = character())
}
```

---

# 3. Gene ID Fixing Function (Your Proven Approach)

```{r gene_id_functions}
cat("=== GENE ID FIXING (YOUR PROVEN APPROACH) ===\n\n")

# Exact same gene ID fix that worked in your successful analysis
fix_gene_ids_proven_approach <- function(original_ids, organism) {
  
  cat(sprintf("Applying proven gene ID fix for %s...\n", organism))
  
  if(organism == "Ha") {
    # For Helianthus: Ha_HanXRQChr01g0000011.1 -> HanXRQChr01g0000011
    cat("Step 1: Removing 'Ha_' prefix...\n")
    step1 <- gsub("^Ha_", "", original_ids)
    
    cat("Step 2: Removing transcript suffix (.1, .2, etc.)...\n")
    fixed <- gsub("\\.[0-9]+$", "", step1)
    
  } else if(organism == "Gm") {
    # For Glycine max: Gm_Glyma.01G008500.1 -> Glyma.01G008500
    cat("Step 1: Removing 'Gm_' prefix...\n")
    step1 <- gsub("^Gm_", "", original_ids)
    
    cat("Step 2: Removing transcript suffix (.1, .2, etc.)...\n")
    fixed <- gsub("\\.[0-9]+$", "", step1)
  }
  
  # Show transformation examples
  cat("\nID transformation examples:\n")
  for(i in 1:min(3, length(original_ids))) {
    if(organism == "Ha") {
      step1_example <- gsub("^Ha_", "", original_ids[i])
    } else {
      step1_example <- gsub("^Gm_", "", original_ids[i])
    }
    cat(sprintf("  Original:  %s\n", original_ids[i]))
    cat(sprintf("  Step 1:    %s\n", step1_example))
    cat(sprintf("  Final:     %s ✓\n\n", fixed[i]))
  }
  
  return(fixed)
}

cat("✓ Gene ID fixing function defined\n")
```

---

# 4. Process Each Timepoint

```{r process_timepoints, results='asis', fig.width=12, fig.height=8}
cat("=== PROCESSING ALL TIMEPOINTS ===\n\n")

# Storage for all results
all_results <- list()

for (tp in timepoints) {
  
  cat("\n\n---\n\n")
  cat(sprintf("## %s Analysis\n\n", tp))
  
  cat(sprintf("=== PROCESSING TIMEPOINT: %s ===\n\n", tp))
  
  # Load DEG file
  deg_file <- sprintf(deg_file_pattern, tp)
  
  if(!file.exists(deg_file)) {
    cat(sprintf("❌ DEG file not found: %s — SKIPPING\n", deg_file))
    next
  }
  
  host_degs <- read.csv(deg_file, stringsAsFactors = FALSE, row.names = 1)
  cat(sprintf("✓ Loaded %d DEGs from %s\n", nrow(host_degs), basename(deg_file)))
  
  # Store original gene IDs
  original_gene_ids <- rownames(host_degs)
  
  # Apply proven gene ID fix
  fixed_gene_ids <- fix_gene_ids_proven_approach(original_gene_ids, ORGANISM)
  
  # Create gene lists
  host_all <- unique(fixed_gene_ids)
  
  # Identify log fold change column
  lfc_col <- grep("log.*fold|lfc", colnames(host_degs), ignore.case = TRUE, value = TRUE)[1]
  
  if(is.na(lfc_col)) {
    cat("❌ No log fold change column found — skipping\n")
    next
  }
  
  # Separate up/down regulated
  upregulated_idx <- host_degs[[lfc_col]] > 0
  downregulated_idx <- host_degs[[lfc_col]] < 0
  
  host_up <- unique(fixed_gene_ids[upregulated_idx])
  host_down <- unique(fixed_gene_ids[downregulated_idx])
  
  cat(sprintf("Gene counts after fix:\n"))
  cat(sprintf("  Total unique genes: %d\n", length(host_all)))
  cat(sprintf("  Up-regulated: %d\n", length(host_up)))
  cat(sprintf("  Down-regulated: %d\n\n", length(host_down)))
  
  # Check annotation coverage using proven approach
  if(annotation_available) {
    matched_genes <- intersect(host_all, unique(term2gene$locusName))
    coverage <- round(100 * length(matched_genes) / length(host_all), 1)
    
    cat("=== COVERAGE CHECK (PROVEN APPROACH) ===\n")
    cat(sprintf("  Unique DEGs: %d\n", length(host_all)))
    cat(sprintf("  Genes with GO annotation: %d\n", length(matched_genes)))
    cat(sprintf("  Coverage: %.1f%%\n\n", coverage))
    
    if(coverage > 10) {
      cat("✓ Sufficient coverage for enrichment analysis!\n\n")
      proceed_with_enrichment <- TRUE
    } else {
      cat("⚠️ Low coverage — enrichment may have limited power\n\n")
      proceed_with_enrichment <- FALSE
    }
  } else {
    coverage <- 0
    proceed_with_enrichment <- FALSE
  }
  
  # Run GO enrichment if coverage is good
  ego_all <- NULL
  ego_up <- NULL
  ego_down <- NULL
  
  if(proceed_with_enrichment) {
    
    cat("Running GO enrichment analysis...\n")
    
    # All DEGs
    if(length(host_all) >= 10) {
      tryCatch({
        ego_all <- enricher(
          gene = host_all,
          TERM2GENE = term2gene,
          pvalueCutoff = 0.05,
          pAdjustMethod = "BH",
          qvalueCutoff = 0.2,
          minGSSize = 10,
          maxGSSize = 500
        )
        
        if(!is.null(ego_all) && nrow(ego_all@result) > 0) {
          cat(sprintf("  ✓ All DEGs: %d enriched GO terms\n", nrow(ego_all@result)))
        } else {
          cat("  ⚠️ All DEGs: No enriched terms found\n")
        }
      }, error = function(e) {
        cat("  ❌ All DEGs enrichment failed:", e$message, "\n")
      })
    }
    
    # Up-regulated DEGs
    if(length(host_up) >= 10) {
      tryCatch({
        ego_up <- enricher(
          gene = host_up,
          TERM2GENE = term2gene,
          pvalueCutoff = 0.05,
          pAdjustMethod = "BH", 
          qvalueCutoff = 0.2,
          minGSSize = 10,
          maxGSSize = 500
        )
        
        if(!is.null(ego_up) && nrow(ego_up@result) > 0) {
          cat(sprintf("  ✓ Up-regulated: %d enriched GO terms\n", nrow(ego_up@result)))
        } else {
          cat("  ⚠️ Up-regulated: No enriched terms found\n")
        }
      }, error = function(e) {
        cat("  ❌ Up-regulated enrichment failed:", e$message, "\n")
      })
    }
    
    # Down-regulated DEGs  
    if(length(host_down) >= 10) {
      tryCatch({
        ego_down <- enricher(
          gene = host_down,
          TERM2GENE = term2gene,
          pvalueCutoff = 0.05,
          pAdjustMethod = "BH",
          qvalueCutoff = 0.2,
          minGSSize = 10,
          maxGSSize = 500
        )
        
        if(!is.null(ego_down) && nrow(ego_down@result) > 0) {
          cat(sprintf("  ✓ Down-regulated: %d enriched GO terms\n", nrow(ego_down@result)))
        } else {
          cat("  ⚠️ Down-regulated: No enriched terms found\n")
        }
      }, error = function(e) {
        cat("  ❌ Down-regulated enrichment failed:", e$message, "\n")
      })
    }
  }
  
  # Store results
  all_results[[tp]] <- list(
    degs = host_degs,
    original_ids = original_gene_ids,
    fixed_ids = fixed_gene_ids,
    host_all = host_all,
    host_up = host_up,
    host_down = host_down,
    coverage = coverage,
    ego_all = ego_all,
    ego_up = ego_up,
    ego_down = ego_down
  )
  
  # Create plots if enrichment successful
  if(!is.null(ego_all) && nrow(ego_all@result) > 0) {
    
    # Dot plot for all DEGs
    p1 <- dotplot(ego_all, showCategory = 20, font.size = 10) +
      ggtitle(paste0(organism_name, " Host DEGs (", tp, ")")) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    
    print(p1)
    
    # Save plot
    ggsave(filename = paste0("plots/", output_prefix, "_dotplot_", tp, ".pdf"), 
           plot = p1, width = 10, height = 8)
  }
  
  # Up vs Down comparison
  if(!is.null(ego_up) && !is.null(ego_down)) {
    if(nrow(ego_up@result) > 0 && nrow(ego_down@result) > 0) {
      
      p_up <- dotplot(ego_up, showCategory = 15, font.size = 9) +
        ggtitle(paste0("Up-regulated (", tp, ")")) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold"))
      
      p_down <- dotplot(ego_down, showCategory = 15, font.size = 9) +
        ggtitle(paste0("Down-regulated (", tp, ")")) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold"))
      
      p_combined <- plot_grid(p_up, p_down, ncol = 1)
      print(p_combined)
      
      ggsave(filename = paste0("plots/", output_prefix, "_up_down_", tp, ".pdf"), 
             plot = p_combined, width = 10, height = 12)
    }
  }
  
  cat(sprintf("\n✓ Timepoint %s analysis complete\n", tp))
}
```

---

# 5. Cross-Timepoint Summary

```{r cross_timepoint_summary}
cat("\n=== CROSS-TIMEPOINT SUMMARY ===\n\n")

if(length(all_results) > 0) {
  
  # Create summary table
  summary_df <- data.frame(
    Timepoint = character(),
    Total_DEGs = integer(),
    Up_regulated = integer(),
    Down_regulated = integer(),
    Coverage_pct = numeric(),
    GO_all = integer(),
    GO_up = integer(),
    GO_down = integer(),
    stringsAsFactors = FALSE
  )
  
  for (tp in names(all_results)) {
    r <- all_results[[tp]]
    summary_df <- rbind(summary_df, data.frame(
      Timepoint = tp,
      Total_DEGs = length(r$host_all),
      Up_regulated = length(r$host_up),
      Down_regulated = length(r$host_down),
      Coverage_pct = r$coverage,
      GO_all = ifelse(!is.null(r$ego_all) && nrow(r$ego_all@result) > 0, nrow(r$ego_all@result), 0),
      GO_up = ifelse(!is.null(r$ego_up) && nrow(r$ego_up@result) > 0, nrow(r$ego_up@result), 0),
      GO_down = ifelse(!is.null(r$ego_down) && nrow(r$ego_down@result) > 0, nrow(r$ego_down@result), 0),
      stringsAsFactors = FALSE
    ))
  }
  
  print(summary_df)
  
  # Plot summary
  summary_long <- summary_df %>%
    select(Timepoint, Up_regulated, Down_regulated) %>%
    pivot_longer(cols = c(Up_regulated, Down_regulated), 
                names_to = "Direction", values_to = "Count")
  
  p_summary <- ggplot(summary_long, aes(x = Timepoint, y = Count, fill = Direction)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = c("Up_regulated" = "red", "Down_regulated" = "blue")) +
    labs(title = paste(organism_name, "Host Response Across Time"),
         subtitle = "Number of Differentially Expressed Genes",
         y = "Number of DEGs") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  print(p_summary)
  
  ggsave(paste0("plots/", output_prefix, "_temporal_summary.pdf"), 
         plot = p_summary, width = 8, height = 6)
  
} else {
  cat("No results to summarize\n")
}
```

---

# 6. Save Results

```{r save_results}
cat("\n=== SAVING RESULTS ===\n\n")

if(length(all_results) > 0) {
  
  output_dir <- paste0(output_prefix, "_enrichment_results")
  
  for (tp in names(all_results)) {
    r <- all_results[[tp]]
    
    cat(sprintf("--- Saving %s results ---\n", tp))
    
    # Save enrichment results
    if(!is.null(r$ego_all) && nrow(r$ego_all@result) > 0) {
      fname <- paste0(output_prefix, "_all_DEGs_GO_enrichment_", tp, ".csv")
      write.csv(r$ego_all@result, file.path(output_dir, fname), row.names = FALSE)
      cat("  ✓", fname, "\n")
    }
    
    if(!is.null(r$ego_up) && nrow(r$ego_up@result) > 0) {
      fname <- paste0(output_prefix, "_upregulated_GO_enrichment_", tp, ".csv")
      write.csv(r$ego_up@result, file.path(output_dir, fname), row.names = FALSE)
      cat("  ✓", fname, "\n")
    }
    
    if(!is.null(r$ego_down) && nrow(r$ego_down@result) > 0) {
      fname <- paste0(output_prefix, "_downregulated_GO_enrichment_", tp, ".csv")
      write.csv(r$ego_down@result, file.path(output_dir, fname), row.names = FALSE)
      cat("  ✓", fname, "\n")
    }
    
    # Save processed gene lists for web tools
    writeLines(r$host_all, file.path(output_dir, paste0(output_prefix, "_all_genes_", tp, ".txt")))
    writeLines(r$host_up, file.path(output_dir, paste0(output_prefix, "_up_genes_", tp, ".txt")))
    writeLines(r$host_down, file.path(output_dir, paste0(output_prefix, "_down_genes_", tp, ".txt")))
    cat("  ✓ Gene lists saved\n")
  }
  
  # Save summary
  write.csv(summary_df, file.path(output_dir, paste0(output_prefix, "_timepoint_summary.csv")), row.names = FALSE)
  cat("\n✓ Summary saved\n")
  
  cat("\nAll results saved to:", output_dir, "\n")
  
} else {
  cat("No results to save\n")
}
```

---

# 7. Biological Interpretation

```{r interpretation}
cat("\n=== BIOLOGICAL INTERPRETATION ===\n\n")

cat(sprintf("Analysis for %s host response to Sclerotinia:\n", organism_name))
cat(paste(rep("=", nchar(organism_name) + 38), collapse=""), "\n\n")

if(length(all_results) > 0) {
  
  for (tp in names(all_results)) {
    r <- all_results[[tp]]
    
    cat(sprintf("--- %s ---\n", tp))
    cat(sprintf("DEGs: %d (Up: %d, Down: %d)\n", 
                length(r$host_all), length(r$host_up), length(r$host_down)))
    cat(sprintf("GO annotation coverage: %.1f%%\n", r$coverage))
    
    n_all <- ifelse(!is.null(r$ego_all) && nrow(r$ego_all@result) > 0, nrow(r$ego_all@result), 0)
    n_up <- ifelse(!is.null(r$ego_up) && nrow(r$ego_up@result) > 0, nrow(r$ego_up@result), 0)
    n_down <- ifelse(!is.null(r$ego_down) && nrow(r$ego_down@result) > 0, nrow(r$ego_down@result), 0)
    
    cat(sprintf("Enriched GO terms: All=%d, Up=%d, Down=%d\n\n", n_all, n_up, n_down))
    
    # Show top enriched terms
    if(!is.null(r$ego_up) && nrow(r$ego_up@result) > 0) {
      cat("Top up-regulated pathways:\n")
      top_up <- r$ego_up@result %>% arrange(pvalue) %>% head(3)
      for(i in 1:nrow(top_up)) {
        cat(sprintf("  • %s (p=%.2e)\n", top_up$Description[i], top_up$pvalue[i]))
      }
      cat("\n")
    }
  }
  
  cat("KEY INSIGHTS:\n")
  cat("• Up-regulated genes = Defense responses to pathogen\n")
  cat("• Down-regulated genes = Suppressed growth/metabolism\n")
  cat("• Time-course = Progressive infection response\n")
  cat("• Compare with pathogen analysis for interaction insights\n\n")
  
} else {
  cat("No enrichment results for interpretation\n")
}
```

---

# 8. Next Steps

```{r next_steps}
cat("=== RECOMMENDED NEXT STEPS ===\n\n")

cat("IMMEDIATE ACTIONS:\n")
cat("1. Run same analysis for the other organism (change ORGANISM variable)\n")
cat("2. Compare results between Helianthus and Glycine max\n")
cat("3. Use gene lists with web tools (g:Profiler, AgriGO) for additional insights\n\n")

cat("WEB ENRICHMENT TOOLS:\n")
cat("• g:Profiler: https://biit.cs.ut.ee/gprofiler/\n")
cat("• AgriGO v2.0: http://systemsbiology.cau.edu.cn/agriGOv2/\n")
cat("• ShinyGO: http://bioinformatics.sdstate.edu/go/\n\n")

cat("COMPARATIVE ANALYSIS:\n")
cat("• Identify conserved defense mechanisms across hosts\n")
cat("• Find host-specific response patterns\n") 
cat("• Link temporal patterns to infection progression\n\n")

if(length(all_results) > 0) {
  total_enriched <- sum(sapply(all_results, function(x) {
    ifelse(!is.null(x$ego_all) && nrow(x$ego_all@result) > 0, nrow(x$ego_all@result), 0)
  }))
  cat(sprintf("✓ SUCCESS: %d enriched GO terms found across timepoints\n", total_enriched))
} else {
  cat("⚠️ Consider using web tools if local enrichment had low coverage\n")
}
```

---

# Session Information

```{r session_info}
sessionInfo()
```

---

**Analysis complete!** This approach uses your exact successful gene ID fixing and annotation processing methods, extended to work with individual timepoint files.
