---
title: "Simplified Host GO Enrichment Analysis (No Bioconductor Required)"
author: "RNA-seq Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# 1. Setup and Library Loading

```{r libraries}
# Load only the essential libraries that are easily available
library(clusterProfiler)
library(ggplot2)
library(dplyr)
library(tidyr)
library(enrichplot)

cat("✓ Essential libraries loaded successfully\n")
cat("Note: This simplified version works without Bioconductor organism packages\n")
```

---

# 2. Manual Setup Instructions

**Before running this analysis, you need to:**

1. **Download GO annotation files** for your organisms from:
   - [Gene Ontology website](http://geneontology.org/docs/download-ontology/)
   - [UniProt-GOA annotations](https://www.ebi.ac.uk/GOA/downloads)
   - Organism-specific databases (Phytozome, Ensembl Plants, etc.)

2. **Prepare annotation files** in the format:
   ```
   gene_id    go_id       go_name
   Gene1      GO:0008150  biological_process
   Gene1      GO:0003674  molecular_function
   Gene2      GO:0009607  response to biotic stimulus
   ```

3. **Update file paths** in the script below to match your data location

---

# 3. Configuration

```{r config}
# ===== UPDATE THESE PATHS =====
# Set your working directory
setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_data/New_salmon_round2")

# Define your organism (change to "Gm" for soybean)
ORGANISM <- "Ha"  # "Ha" for Helianthus, "Gm" for Glycine max

# Define file patterns for DEG files
if(ORGANISM == "Ha") {
  deg_file_pattern <- "./Ha_host_aggressive_vs_low_DEGs_%s.csv"
  organism_name <- "Helianthus"
  go_annotation_file <- "./Ha_GO_annotation.txt"  # CREATE THIS FILE
} else if(ORGANISM == "Gm") {
  deg_file_pattern <- "./Gm_host_aggressive_vs_low_DEGs_%s.csv"
  organism_name <- "Glycine max"
  go_annotation_file <- "./Gm_GO_annotation.txt"  # CREATE THIS FILE
}

timepoints <- c("24h", "48h", "96h")

cat(sprintf("Analysis configured for: %s\n", organism_name))
cat(sprintf("GO annotation file: %s\n", go_annotation_file))
```

---

# 4. Load GO Annotation (Manual)

```{r load_annotation}
cat("\n=== LOADING GO ANNOTATION ===\n\n")

# Check if annotation file exists
if(file.exists(go_annotation_file)) {
  
  # Try to load custom GO annotation
  tryCatch({
    go_annot <- read.delim(go_annotation_file, header = TRUE, stringsAsFactors = FALSE)
    
    cat("✓ Loaded GO annotation file\n")
    cat("  Rows:", nrow(go_annot), "\n")
    cat("  Columns:", paste(colnames(go_annot), collapse = ", "), "\n")
    
    # Show expected format
    if(nrow(go_annot) > 0) {
      print(head(go_annot, 3))
    }
    
    annotation_available <- TRUE
    
  }, error = function(e) {
    cat("❌ Error reading annotation file:", e$message, "\n")
    annotation_available <- FALSE
  })
  
} else {
  cat("❌ GO annotation file not found:", go_annotation_file, "\n\n")
  
  cat("To create this file, you can:\n")
  cat("1. Download from organism databases (Phytozome, Ensembl)\n")
  cat("2. Use online tools like g:Profiler or DAVID\n") 
  cat("3. Extract from GFF3/GTF files with GO annotations\n")
  cat("4. Use biomaRt to query online databases\n\n")
  
  cat("Expected format (tab-separated):\n")
  cat("gene_id\tgo_id\tgo_name\n")
  cat("Gene001\tGO:0008150\tbiological_process\n")
  cat("Gene001\tGO:0009607\tresponse to biotic stimulus\n")
  cat("Gene002\tGO:0006952\tdefense response\n\n")
  
  annotation_available <- FALSE
}
```

---

# 5. Alternative: Use Pre-defined GO Categories

```{r predefined_categories}
if(!annotation_available) {
  cat("\n=== USING PRE-DEFINED GO CATEGORIES ===\n\n")
  
  # Define key plant defense-related GO terms manually
  # These are the most relevant for host-pathogen interactions
  
  defense_go_terms <- data.frame(
    GO_ID = c(
      "GO:0006952",  # defense response
      "GO:0009607",  # response to biotic stimulus
      "GO:0009626",  # plant-type hypersensitive response
      "GO:0009627",  # systemic acquired resistance
      "GO:0009611",  # response to wounding
      "GO:0071555",  # cell wall organization
      "GO:0009809",  # lignin biosynthetic process
      "GO:0052544",  # defense response by callose deposition
      "GO:0006950",  # response to stress
      "GO:0042542",  # response to hydrogen peroxide
      "GO:0009737",  # response to abscisic acid
      "GO:0009751",  # response to salicylic acid
      "GO:0009753",  # response to jasmonic acid
      "GO:0010200",  # response to chitin
      "GO:0002679",  # respiratory burst involved in defense response
      "GO:0009595",  # detection of biotic stimulus
      "GO:0052033",  # pathogen-associated molecular pattern dependent induction
      "GO:0043900"   # regulation of multi-organism process
    ),
    GO_Name = c(
      "defense response",
      "response to biotic stimulus", 
      "plant-type hypersensitive response",
      "systemic acquired resistance",
      "response to wounding",
      "cell wall organization",
      "lignin biosynthetic process",
      "defense response by callose deposition in cell wall",
      "response to stress",
      "response to hydrogen peroxide",
      "response to abscisic acid",
      "response to salicylic acid",
      "response to jasmonic acid",
      "response to chitin",
      "respiratory burst involved in defense response",
      "detection of biotic stimulus",
      "pathogen-associated molecular pattern dependent induction by symbiont",
      "regulation of multi-organism process"
    ),
    stringsAsFactors = FALSE
  )
  
  cat("✓ Using pre-defined defense-related GO categories\n")
  cat("  Number of categories:", nrow(defense_go_terms), "\n")
  print(defense_go_terms)
}
```

---

# 6. Gene List Analysis (Without Enrichment)

```{r gene_analysis, results='asis', fig.width=12, fig.height=8}
cat("\n=== ANALYZING DEG FILES ===\n\n")

# Function to clean gene IDs based on organism
clean_gene_id <- function(gene_ids, organism) {
  if(organism == "Ha") {
    # Helianthus gene ID cleaning
    cleaned <- gsub("^[Hh]a_", "", gene_ids)
    cleaned <- gsub("^HanXRQ_", "", cleaned)
    cleaned <- gsub("\\.\\d+$", "", cleaned)
    cleaned <- gsub("_.*$", "", cleaned)
  } else if(organism == "Gm") {
    # Glycine max gene ID cleaning  
    cleaned <- gsub("^[Gg]lyma\\.?", "", gene_ids)
    cleaned <- gsub("^[Gg]m_", "", cleaned)
    cleaned <- gsub("\\.\\d+$", "", cleaned)
    cleaned <- gsub("_.*$", "", cleaned)
  } else {
    cleaned <- gene_ids
  }
  return(cleaned)
}

# Store results for each timepoint
all_results <- list()

for (tp in timepoints) {
  
  cat("\n\n---\n\n")
  cat(sprintf("## %s Analysis for %s {.tabset}\n\n", tp, organism_name))
  
  cat(sprintf("\n=== PROCESSING TIMEPOINT: %s ===\n\n", tp))
  
  # Load DEG file
  deg_file <- sprintf(deg_file_pattern, tp)
  
  if(!file.exists(deg_file)) {
    cat(sprintf("❌ DEG file not found: %s — SKIPPING\n", deg_file))
    next
  }
  
  host_degs <- read.csv(deg_file, stringsAsFactors = FALSE, row.names = 1)
  cat(sprintf("✓ Loaded %d DEGs from %s\n", nrow(host_degs), basename(deg_file)))
  
  # Process gene IDs
  original_gene_ids <- rownames(host_degs)
  host_degs$clean_gene_id <- clean_gene_id(original_gene_ids, ORGANISM)
  host_degs$original_gene_id <- original_gene_ids
  
  # Find log fold change column
  lfc_col <- grep("log.*fold|lfc", colnames(host_degs), ignore.case = TRUE, value = TRUE)[1]
  
  if(is.na(lfc_col)) {
    cat("  ❌ No log fold change column found — skipping\n")
    next
  }
  
  # Separate up/down-regulated
  upregulated   <- host_degs[host_degs[[lfc_col]] > 0, ]
  downregulated <- host_degs[host_degs[[lfc_col]] < 0, ]
  
  cat(sprintf("  Up-regulated: %d | Down-regulated: %d\n", 
              nrow(upregulated), nrow(downregulated)))
  
  # Summary statistics
  cat(sprintf("  Log fold change range: %.2f to %.2f\n",
              min(host_degs[[lfc_col]], na.rm = TRUE),
              max(host_degs[[lfc_col]], na.rm = TRUE)))
  
  # Look for defense-related genes by gene ID patterns (if available)
  defense_patterns <- c("PR", "PR1", "PR2", "PR3", "PR4", "PR5",  # Pathogenesis-related
                       "PAL", "CAD", "4CL",                        # Phenylpropanoid pathway
                       "LOX", "AOS", "AOC",                        # Jasmonic acid pathway
                       "NPR", "TGA",                               # SA signaling
                       "WRKY", "ERF",                              # Defense transcription factors
                       "GST", "CYP",                               # Detoxification
                       "CHI", "CHS",                               # Flavonoid pathway
                       "MLO", "RPM", "RPS", "RPP")                 # Resistance genes
  
  defense_genes <- host_degs[grepl(paste(defense_patterns, collapse = "|"), 
                                  host_degs$clean_gene_id, ignore.case = TRUE), ]
  
  if(nrow(defense_genes) > 0) {
    cat(sprintf("  Defense-related genes found: %d\n", nrow(defense_genes)))
    cat("  Examples:", paste(head(defense_genes$clean_gene_id, 5), collapse = ", "), "\n")
  }
  
  # Create plots
  # Volcano plot
  if("pvalue" %in% colnames(host_degs) || "PValue" %in% colnames(host_degs)) {
    pval_col <- ifelse("pvalue" %in% colnames(host_degs), "pvalue", "PValue")
    
    volcano_data <- data.frame(
      logFC = host_degs[[lfc_col]],
      negLogP = -log10(host_degs[[pval_col]]),
      significant = (host_degs[[pval_col]] < 0.05 & abs(host_degs[[lfc_col]]) > 1)
    )
    
    p_volcano <- ggplot(volcano_data, aes(x = logFC, y = negLogP, color = significant)) +
      geom_point(alpha = 0.6) +
      scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
      labs(title = sprintf("%s Host DEGs - %s", organism_name, tp),
           x = "Log2 Fold Change", y = "-log10(p-value)") +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    
    print(p_volcano)
  }
  
  # Expression distribution
  p_dist <- ggplot(host_degs, aes(x = !!sym(lfc_col))) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    labs(title = sprintf("Log Fold Change Distribution - %s", tp),
         x = "Log2 Fold Change", y = "Number of Genes") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  print(p_dist)
  
  # Store results
  all_results[[tp]] <- list(
    degs = host_degs,
    n_up = nrow(upregulated),
    n_down = nrow(downregulated),
    defense_genes = defense_genes
  )
  
  cat(sprintf("\n✓ Timepoint %s analysis complete\n", tp))
}
```

---

# 7. Cross-Timepoint Summary

```{r summary_analysis}
cat("\n=== CROSS-TIMEPOINT SUMMARY ===\n\n")

if(length(all_results) > 0) {
  
  # Create summary table
  summary_df <- data.frame(
    Timepoint = character(),
    Total_DEGs = integer(),
    Up_regulated = integer(),
    Down_regulated = integer(),
    Defense_genes = integer(),
    stringsAsFactors = FALSE
  )
  
  for (tp in names(all_results)) {
    r <- all_results[[tp]]
    summary_df <- rbind(summary_df, data.frame(
      Timepoint = tp,
      Total_DEGs = nrow(r$degs),
      Up_regulated = r$n_up,
      Down_regulated = r$n_down,
      Defense_genes = nrow(r$defense_genes),
      stringsAsFactors = FALSE
    ))
  }
  
  print(summary_df)
  
  # Plot summary
  summary_long <- summary_df %>%
    select(-Total_DEGs, -Defense_genes) %>%
    pivot_longer(cols = c(Up_regulated, Down_regulated), 
                names_to = "Direction", values_to = "Count")
  
  p_summary <- ggplot(summary_long, aes(x = Timepoint, y = Count, fill = Direction)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = c("Up_regulated" = "red", "Down_regulated" = "blue")) +
    labs(title = sprintf("%s Host Response Across Time", organism_name),
         subtitle = "Number of Differentially Expressed Genes",
         y = "Number of DEGs") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  print(p_summary)
  
} else {
  cat("No results to summarize\n")
}
```

---

# 8. Save Results

```{r save_results}
cat("\n=== SAVING RESULTS ===\n\n")

if(length(all_results) > 0) {
  
  output_dir <- sprintf("./%s_host_analysis_simplified", ORGANISM)
  if(!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
    cat("✓ Created output directory:", output_dir, "\n")
  }
  
  # Save processed DEG files
  for (tp in names(all_results)) {
    r <- all_results[[tp]]
    
    # Save all DEGs with processed IDs
    fname <- sprintf("%s_host_DEGs_processed_%s.csv", ORGANISM, tp)
    write.csv(r$degs, file.path(output_dir, fname), row.names = FALSE)
    cat("  ✓", fname, "\n")
    
    # Save defense genes if found
    if(nrow(r$defense_genes) > 0) {
      fname <- sprintf("%s_host_defense_genes_%s.csv", ORGANISM, tp)
      write.csv(r$defense_genes, file.path(output_dir, fname), row.names = FALSE)
      cat("  ✓", fname, "\n")
    }
  }
  
  # Save summary
  if(exists("summary_df")) {
    fname <- sprintf("%s_host_summary.csv", ORGANISM)
    write.csv(summary_df, file.path(output_dir, fname), row.names = FALSE)
    cat("  ✓", fname, "\n")
  }
  
  cat("\nAll results saved to:", output_dir, "\n")
  
} else {
  cat("No results to save\n")
}
```

---

# 9. Next Steps

```{r next_steps}
cat("\n=== NEXT STEPS FOR FULL GO ENRICHMENT ===\n\n")

cat("This simplified analysis provides an overview of your host DEGs.\n")
cat("For full GO enrichment analysis, you have several options:\n\n")

cat("OPTION 1 - Install Bioconductor packages:\n")
cat("```r\n")
cat("if (!require('BiocManager', quietly = TRUE))\n")
cat("    install.packages('BiocManager')\n")
cat("BiocManager::install('org.At.tair.db')  # For Arabidopsis reference\n")
cat("BiocManager::install('GO.db')           # For GO term descriptions\n")
cat("```\n\n")

cat("OPTION 2 - Use online enrichment tools:\n")
cat("- g:Profiler (https://biit.cs.ut.ee/gprofiler/)\n")
cat("- DAVID (https://david.ncifcrf.gov/)\n")
cat("- Enrichr (https://maayanlab.cloud/Enrichr/)\n")
cat("- AgriGO (http://systemsbiology.cau.edu.cn/agriGOv2/)\n\n")

cat("OPTION 3 - Create custom GO annotation files:\n")
cat("- Download from Phytozome, Ensembl Plants, or other databases\n")
cat("- Use biomaRt to query online databases\n")
cat("- Extract from organism-specific annotation files\n\n")

cat("OPTION 4 - Use alternative functional annotation:\n")
cat("- KEGG pathway analysis\n")
cat("- InterPro domain analysis\n") 
cat("- eggNOG functional annotation\n")
cat("- Mercator4 plant-specific annotation\n\n")

if(length(all_results) > 0) {
  total_degs <- sum(sapply(all_results, function(x) nrow(x$degs)))
  cat(sprintf("Your dataset contains %d total DEGs across timepoints.\n", total_degs))
  cat("This is a good size for functional enrichment analysis!\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
