---
title: "Soybean-Sclerotinia Isolate Comparison Analysis (Salmon)"
author: "Analysis by wan00965"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 12, fig.height = 8)
```

# Overview

This notebook analyzes soybean (Williams 82) responses to different **Sclerotinia sclerotiorum** isolates with varying aggressiveness levels.

**Isolate Categories:**
- **Low aggressive:** WISS47, JS659
- **High aggressive:** Xtra7, MNSS6
- **Soybean-specific:** SsPotter
- **Sunflower-specific:** BN172

**Timepoints:** NC (negative control), 24h, 48h, 96h

# Setup

## Install Required Packages (Run Once)

```{r install-packages, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("DESeq2", "ComplexHeatmap", "EnhancedVolcano"))
install.packages(c("ggplot2", "pheatmap", "RColorBrewer", "dplyr", "tidyr", "gridExtra"))
```

## Load Libraries

```{r load-libraries}
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(EnhancedVolcano)
library(gridExtra)

cat("All libraries loaded successfully!\n")
```

## Set Working Directory

```{r set-directory}
# Update this path to where your count files are located
setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_data/New_salmon_round3/new_run_count")

# Verify we're in the right place
list.files(pattern = "gene_counts.*soybean.*SALMON")
```

# Load Data

```{r load-data}
# Load count matrices
counts_all <- read.table("gene_counts_soybean_SALMON.txt", 
                         header = TRUE, row.names = 1, check.names = FALSE)
counts_host <- read.table("gene_counts_soybean_SALMON_HOST_only.txt", 
                          header = TRUE, row.names = 1, check.names = FALSE)
counts_pathogen <- read.table("gene_counts_soybean_SALMON_PATHOGEN_only.txt", 
                              header = TRUE, row.names = 1, check.names = FALSE)

cat(sprintf("Total genes: %d\n", nrow(counts_all)))
cat(sprintf("Host genes (Gm_): %d\n", nrow(counts_host)))
cat(sprintf("Pathogen genes (Ss_): %d\n", nrow(counts_pathogen)))
cat(sprintf("Samples: %d\n", ncol(counts_all)))
```

## Create Metadata with Isolate Information

```{r create-metadata}
# Sample name pattern: timepoint_replicate_species_isolate
# Examples: "0_1_Gm_NC", "24_1_Gm_BN172", "48_2_Gm_MNSS6"

sample_names <- colnames(counts_all)

metadata <- data.frame(
  sample = sample_names,
  row.names = sample_names,
  stringsAsFactors = FALSE
)

# Extract timepoint (first number)
metadata$timepoint_num <- sapply(strsplit(sample_names, "_"), `[`, 1)
metadata$replicate <- sapply(strsplit(sample_names, "_"), `[`, 2)
metadata$batch <- sapply(strsplit(sample_names, "_"), `[`, 2)
metadata$species <- sapply(strsplit(sample_names, "_"), `[`, 3)

# Extract isolate name (last element)
metadata$isolate <- sapply(strsplit(sample_names, "_"), function(x) x[length(x)])

# Create timepoint with hours
metadata$timepoint <- sapply(sample_names, function(x) {
  parts <- strsplit(x, "_")[[1]]
  if(parts[length(parts)] == "NC") {
    return("NC")
  } else {
    return(paste0(parts[1], "h"))
  }
})

# Assign aggressiveness categories
metadata$aggressiveness <- sapply(metadata$isolate, function(iso) {
  if(iso == "NC") return("Control")
  if(iso %in% c("WISS47", "JS659")) return("Low")
  if(iso %in% c("Xtra7", "MNSS6")) return("High")
  if(iso == "SSPotter") return("Soy-specific")
  if(iso == "BN172") return("Sun-specific")
  return("Unknown")
})

# Convert to factors with proper order
metadata$timepoint <- factor(metadata$timepoint, levels = c("NC", "24h", "48h", "96h"))
metadata$batch <- factor(metadata$batch)
metadata$isolate <- factor(metadata$isolate)
metadata$aggressiveness <- factor(metadata$aggressiveness, 
                                  levels = c("Control", "Low", "High", 
                                            "Soy-specific", "Sun-specific"))

# Display summary
cat("\n=== METADATA SUMMARY ===\n")
print(head(metadata, 10))

cat("\n=== Samples per timepoint ===\n")
print(table(metadata$timepoint, useNA = "always"))

cat("\n=== Samples per isolate ===\n")
print(table(metadata$isolate, useNA = "always"))

cat("\n=== Samples per aggressiveness category ===\n")
print(table(metadata$aggressiveness, useNA = "always"))

cat("\n=== Isolates by timepoint ===\n")
print(table(metadata$timepoint, metadata$isolate))
```

# Quality Control

## Library Sizes

```{r qc-library-sizes}
lib_sizes <- colSums(counts_host)

lib_df <- data.frame(
  sample = names(lib_sizes),
  size = lib_sizes / 1e6,
  timepoint = metadata$timepoint,
  isolate = metadata$isolate
)

ggplot(lib_df, aes(x = reorder(sample, size), y = size, fill = isolate)) +
  geom_bar(stat = "identity") +
  facet_wrap(~timepoint, scales = "free_x") +
  labs(title = "Library Sizes by Isolate",
       x = "Sample", y = "Million Reads",
       fill = "Isolate") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

## Pathogen Load Analysis

```{r qc-pathogen-load}
host_reads <- colSums(counts_host)
pathogen_reads <- colSums(counts_pathogen)

pathogen_load_df <- data.frame(
  sample = names(host_reads),
  host = host_reads,
  pathogen = pathogen_reads,
  pathogen_pct = (pathogen_reads / (host_reads + pathogen_reads)) * 100,
  timepoint = metadata$timepoint,
  isolate = metadata$isolate,
  aggressiveness = metadata$aggressiveness
)

# Plot pathogen load by isolate
ggplot(pathogen_load_df[pathogen_load_df$timepoint != "NC",], 
       aes(x = timepoint, y = pathogen_pct, fill = isolate)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.5) +
  labs(title = "Pathogen Load by Isolate Over Time",
       x = "Timepoint", y = "Pathogen Reads (%)") +
  theme_minimal() +
  theme(legend.position = "right")

# Summary statistics by isolate
cat("\n=== Pathogen Load Summary by Isolate ===\n")
pathogen_summary <- pathogen_load_df %>%
  filter(timepoint != "NC") %>%
  group_by(isolate, timepoint) %>%
  summarise(
    mean_pct = mean(pathogen_pct),
    sd_pct = sd(pathogen_pct),
    n = n(),
    .groups = "drop"
  )
print(pathogen_summary)
```

## Pathogen Load by Aggressiveness

```{r qc-pathogen-aggressiveness}
# Compare pathogen load between aggressive categories
ggplot(pathogen_load_df[pathogen_load_df$aggressiveness != "Control",], 
       aes(x = timepoint, y = pathogen_pct, fill = aggressiveness)) +
  geom_boxplot() +
  labs(title = "Pathogen Load by Aggressiveness Category",
       x = "Timepoint", y = "Pathogen Reads (%)") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

# Statistical summary
cat("\n=== Pathogen Load by Aggressiveness ===\n")
agg_summary <- pathogen_load_df %>%
  filter(aggressiveness != "Control") %>%
  group_by(aggressiveness, timepoint) %>%
  summarise(
    mean_pct = mean(pathogen_pct),
    sd_pct = sd(pathogen_pct),
    n = n(),
    .groups = "drop"
  )
print(agg_summary)
```

# HOST GENE ANALYSIS

## Prepare DESeq2 Dataset - Host Genes

```{r deseq2-host-setup}
# IMPORTANT: Salmon produces non-integer counts, round them for DESeq2
counts_host <- round(counts_host)

# Filter low-expressed genes
keep <- rowSums(counts_host >= 10) >= 3
counts_host_filtered <- counts_host[keep, ]

cat(sprintf("Genes after filtering: %d (from %d)\n", 
            nrow(counts_host_filtered), nrow(counts_host)))

# ===================================================================
# CRITICAL: Remove NC samples to avoid isolate/timepoint confounding
# NC samples are negative controls (no pathogen) and shouldn't be
# compared as an "isolate" alongside real pathogen isolates
# ===================================================================

metadata_infected <- droplevels(metadata[metadata$timepoint != "NC", ])
# droplevels() is critical: subsetting a factor in R retains ALL original levels
# even for groups with zero observations. Without it, 'isolate' still carries a
# "NC" level → model.matrix adds an all-zero dummy column → DESeq2 errors.
counts_host_infected <- counts_host_filtered[, rownames(metadata_infected)]

cat(sprintf("\nAnalyzing infected samples only: %d samples\n", ncol(counts_host_infected)))
cat("Samples per isolate at infection timepoints:\n")
print(table(metadata_infected$isolate, metadata_infected$timepoint))

# Create DESeq2 dataset
# Design: isolate + timepoint
# NOTE: 'batch' (replicate number) is NESTED within isolate — rep 1 of WISS47
# and rep 1 of Xtra7 are independent experiments, not shared processing batches.
# Including batch alongside isolate in the same formula creates a rank-deficient
# model matrix (they are linearly dependent). Drop batch here; the per-timepoint
# subsets below also omit it for the same reason.
dds_host <- DESeqDataSetFromMatrix(
  countData = counts_host_infected,
  colData = metadata_infected,
  design = ~ isolate + timepoint
)

# Run DESeq2
dds_host <- DESeq(dds_host)

cat("\nDESeq2 analysis complete for host genes!\n")
```

## PCA - Host Response by Isolate

```{r pca-host-isolate}
# Variance stabilizing transformation
vst_host <- vst(dds_host, blind = FALSE)

# PCA plot colored by isolate
pcaData <- plotPCA(vst_host, intgroup = c("timepoint", "isolate"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color = isolate, shape = timepoint)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA: Host Gene Expression by Isolate") +
  theme_minimal() +
  theme(legend.position = "right")
```

## Host DEG Analysis: High vs Low Aggressive Isolates — All Timepoints

```{r host-high-vs-low-all-timepoints}
# ===================================================================
# Compare host responses to Aggressive vs Low isolates at 24h, 48h, 96h
# Produces one CSV per timepoint for downstream GO term analysis
# ===================================================================

timepoints <- c("24h", "48h", "96h")

# Store results in a list for later use
res_agg_list <- list()

for (tp in timepoints) {
  
  cat("\n============================================================\n")
  cat(sprintf("  Processing timepoint: %s\n", tp))
  cat("============================================================\n")
  
  # Subset metadata to this timepoint
  metadata_tp <- metadata_infected[metadata_infected$timepoint == tp, ]
  dds_tp <- dds_host[, rownames(metadata_tp)]
  
  # Create simplified aggressiveness grouping
  # High + Soy-specific → "Aggressive"; Low → "Low"; others → "Other"
  colData(dds_tp)$agg_simple <- ifelse(
    colData(dds_tp)$aggressiveness %in% c("High", "Soy-specific"), "Aggressive",
    ifelse(colData(dds_tp)$aggressiveness == "Low", "Low", "Other")
  )
  colData(dds_tp)$agg_simple <- factor(colData(dds_tp)$agg_simple)
  
  # Show sample counts per group
  cat("Samples per group:\n")
  print(table(colData(dds_tp)$agg_simple))
  
  # Re-run DESeq2 with simplified design
  # NOTE: Batch removed — nested within isolate at each single timepoint
  design(dds_tp) <- ~ agg_simple
  dds_tp <- DESeq(dds_tp)
  
  # Get results: Aggressive vs Low
  res <- results(dds_tp, contrast = c("agg_simple", "Aggressive", "Low"), alpha = 0.05)
  res <- res[order(res$padj), ]
  
  # Store in list
  res_agg_list[[tp]] <- res
  
  # Print summary
  n_sig   <- sum(res$padj < 0.05, na.rm = TRUE)
  n_up    <- sum(res$padj < 0.05 & res$log2FoldChange > 0, na.rm = TRUE)
  n_down  <- sum(res$padj < 0.05 & res$log2FoldChange < 0, na.rm = TRUE)
  
  cat(sprintf("\n--- Host Response: Aggressive vs Low (%s) ---\n", tp))
  cat(sprintf("Significant DEGs (padj < 0.05): %d\n", n_sig))
  cat(sprintf("  Upregulated in Aggressive:    %d\n", n_up))
  cat(sprintf("  Downregulated in Aggressive:  %d\n", n_down))
  
  # Volcano plot
  print(
    EnhancedVolcano(res,
                    lab = rownames(res),
                    x = 'log2FoldChange',
                    y = 'padj',
                    title = paste0('Host: Aggressive vs Low (', tp, ')'),
                    subtitle = 'Soybean genes',
                    pCutoff = 0.05,
                    FCcutoff = 1.0,
                    pointSize = 2.0,
                    labSize = 3.0,
                    maxoverlapsConnectors = 20)
  )
}
```

## Export Host DEG Results for GO Analysis

```{r export-host-high-vs-low-all-timepoints}
# ===================================================================
# Export one CSV per timepoint (significant DEGs, padj < 0.05)
# Ready for GO term enrichment analysis
# ===================================================================

for (tp in timepoints) {
  
  res <- res_agg_list[[tp]]
  
  # Keep significant genes only (remove NAs, filter padj < 0.05)
  sig <- res[!is.na(res$padj) & res$padj < 0.05, ]
  sig_df <- as.data.frame(sig)
  
  # Build filename
  filename <- paste0("Gm_host_aggressive_vs_low_DEGs_", tp, ".csv")
  
  write.csv(sig_df, filename)
  
  cat(sprintf("Exported %d DEGs → %s\n", nrow(sig_df), filename))
}

cat("\nAll 3 timepoint CSVs exported for GO analysis!\n")
```

## Host DEG Analysis: Individual Isolate Comparisons

```{r host-SSPotter-specific}
##example

# Compare sunflower response to BN172 vs generalist isolates at 48h
metadata_48h <- metadata_infected[metadata_infected$timepoint == "48h",]
dds_48h <- dds_host[, rownames(metadata_48h)]

# BN172 vs Low aggressive isolates
colData(dds_48h)$iso_group <- ifelse(
  colData(dds_48h)$isolate == "SSPotter", "SSPotter",
  ifelse(colData(dds_48h)$aggressiveness == "Low", "Low", "Other")
)
colData(dds_48h)$iso_group <- factor(colData(dds_48h)$iso_group)

# NOTE: Batch removed - nested within isolate at 48h
design(dds_48h) <- ~ iso_group
dds_48h <- DESeq(dds_48h)

res_SSPotter <- results(dds_48h, contrast = c("iso_group", "SSPotter", "Low"), alpha = 0.05)
res_SSPotter <- res_SSPotter[order(res_SSPotter$padj),]

cat("\n=== Soybean Response: SSPotter (specialist) vs Low Aggressive (48h) ===\n")
cat(sprintf("Significant DEGs: %d\n", sum(res_SSPotter$padj < 0.05, na.rm = TRUE)))
cat(sprintf("Upregulated with SSPotter: %d\n", 
            sum(res_SSPotter$padj < 0.05 & res_SSPotter$log2FoldChange > 0, na.rm = TRUE)))

# Volcano plot
EnhancedVolcano(res_SSPotter,
                lab = rownames(res_SSPotter),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Soybean: SSPotter (specialist) vs Low Aggressive',
                pCutoff = 0.05,
                FCcutoff = 1.0)
```

# PATHOGEN GENE ANALYSIS

## Prepare DESeq2 Dataset - Pathogen Genes

```{r deseq2-pathogen-setup}
# Round counts
counts_pathogen <- round(counts_pathogen)

# Note: metadata_infected was already created in host analysis section above
# It excludes NC samples (negative controls without pathogen)
counts_pathogen_infected <- counts_pathogen[, rownames(metadata_infected)]

# Filter low-expressed genes
keep_pathogen <- rowSums(counts_pathogen_infected >= 5) >= 3
counts_pathogen_filtered <- counts_pathogen_infected[keep_pathogen, ]

cat(sprintf("Pathogen genes after filtering: %d (from %d)\n", 
            nrow(counts_pathogen_filtered), nrow(counts_pathogen)))

# Create DESeq2 dataset
# NOTE: batch is nested within isolate (same reason as host model above),
# so the design uses isolate + timepoint only.
dds_pathogen <- DESeqDataSetFromMatrix(
  countData = counts_pathogen_filtered,
  colData = metadata_infected,
  design = ~ isolate + timepoint
)

# Run DESeq2
dds_pathogen <- DESeq(dds_pathogen)

cat("\nDESeq2 analysis complete for pathogen genes!\n")
```

## PCA - Pathogen Gene Expression by Isolate

```{r pca-pathogen-isolate}
vst_pathogen <- vst(dds_pathogen, blind = FALSE)

pcaData_path <- plotPCA(vst_pathogen, intgroup = c("timepoint", "isolate"), returnData = TRUE)
percentVar_path <- round(100 * attr(pcaData_path, "percentVar"))

ggplot(pcaData_path, aes(PC1, PC2, color = isolate, shape = timepoint)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab(paste0("PC1: ", percentVar_path[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_path[2], "% variance")) +
  ggtitle("PCA: Pathogen Gene Expression by Isolate") +
  theme_minimal() +
  theme(legend.position = "right")
```

## Pathogen DEG: High vs Low Aggressive Isolates — All Timepoints

```{r pathogen-high-vs-low-all-timepoints}
# ===================================================================
# Compare pathogen gene expression: Aggressive vs Low at 24h, 48h, 96h
# Produces one CSV per timepoint for downstream GO term analysis
# ===================================================================

timepoints <- c("24h", "48h", "96h")

# Store results in a list for later use
res_path_agg_list <- list()

for (tp in timepoints) {
  
  cat("\n============================================================\n")
  cat(sprintf("  Processing pathogen genes at timepoint: %s\n", tp))
  cat("============================================================\n")
  
  # Subset metadata to this timepoint
  metadata_path_tp <- metadata_infected[metadata_infected$timepoint == tp, ]
  dds_path_tp <- dds_pathogen[, rownames(metadata_path_tp)]
  
  # Create simplified aggressiveness grouping
  # High + Soy-specific → "Aggressive"; Low → "Low"; others → "Other"
  colData(dds_path_tp)$agg_simple <- ifelse(
    colData(dds_path_tp)$aggressiveness %in% c("High", "Soy-specific"), "Aggressive",
    ifelse(colData(dds_path_tp)$aggressiveness == "Low", "Low", "Other")
  )
  colData(dds_path_tp)$agg_simple <- factor(colData(dds_path_tp)$agg_simple)
  
  # Show sample counts per group
  cat("Samples per group:\n")
  print(table(colData(dds_path_tp)$agg_simple))
  
  # Re-run DESeq2 with simplified design
  # NOTE: Batch removed — nested within isolate at each single timepoint
  design(dds_path_tp) <- ~ agg_simple
  dds_path_tp <- DESeq(dds_path_tp)
  
  # Get results: Aggressive vs Low
  res <- results(dds_path_tp, contrast = c("agg_simple", "Aggressive", "Low"), alpha = 0.05)
  res <- res[order(res$padj), ]
  
  # Store in list
  res_path_agg_list[[tp]] <- res
  
  # Print summary
  n_sig   <- sum(res$padj < 0.05, na.rm = TRUE)
  n_up    <- sum(res$padj < 0.05 & res$log2FoldChange > 0, na.rm = TRUE)
  n_down  <- sum(res$padj < 0.05 & res$log2FoldChange < 0, na.rm = TRUE)
  
  cat(sprintf("\n--- Pathogen Genes: Aggressive vs Low (%s) ---\n", tp))
  cat(sprintf("Significant DEGs (padj < 0.05): %d\n", n_sig))
  cat(sprintf("  Upregulated in Aggressive:    %d\n", n_up))
  cat(sprintf("  Downregulated in Aggressive:  %d\n", n_down))
  
  # Volcano plot
  print(
    EnhancedVolcano(res,
                    lab = rownames(res),
                    x = 'log2FoldChange',
                    y = 'padj',
                    title = paste0('Pathogen: Aggressive vs Low (', tp, ')'),
                    subtitle = 'S. sclerotiorum genes',
                    pCutoff = 0.05,
                    FCcutoff = 1.0,
                    pointSize = 2.0,
                    labSize = 3.0)
  )
}
```

## Export Pathogen DEG Results for GO Analysis

```{r export-pathogen-high-vs-low-all-timepoints}
# ===================================================================
# Export one CSV per timepoint (significant DEGs, padj < 0.05)
# Ready for GO term enrichment analysis
# ===================================================================

for (tp in timepoints) {
  
  res <- res_path_agg_list[[tp]]
  
  # Keep significant genes only (remove NAs, filter padj < 0.05)
  sig <- res[!is.na(res$padj) & res$padj < 0.05, ]
  sig_df <- as.data.frame(sig)
  
  # Build filename
  filename <- paste0("Gm_pathogen_aggressive_vs_low_DEGs_", tp, ".csv")
  
  write.csv(sig_df, filename)
  
  cat(sprintf("Exported %d DEGs → %s\n", nrow(sig_df), filename))
}

cat("\nAll 3 pathogen timepoint CSVs exported for GO analysis!\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
