---
title: "Sunflower-Sclerotinia DESeq2 Analysis (Salmon)"
author: "Analysis by wan00965"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 7)
```

# Setup

## Load Libraries

```{r load-libraries}
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(EnhancedVolcano)

cat("All libraries loaded successfully!\n")
```

## Set Working Directory

```{r set-directory}
# Update this path to where your count files are located
setwd("/Users/andrew20445/Desktop/To_do_list/RNAseq_data/New_salmon_round3/new_run_count")

# Verify we're in the right place
list.files(pattern = "gene_counts.*sunflower.*SALMON")
```

# Load Data

```{r load-data}
# Load count matrices
counts_all <- read.table("gene_counts_sunflower_SALMON.txt", 
                         header = TRUE, row.names = 1, check.names = FALSE)
counts_host <- read.table("gene_counts_sunflower_SALMON_HOST_only.txt", 
                          header = TRUE, row.names = 1, check.names = FALSE)
counts_pathogen <- read.table("gene_counts_sunflower_SALMON_PATHOGEN_only.txt", 
                              header = TRUE, row.names = 1, check.names = FALSE)

cat(sprintf("Total genes: %d\n", nrow(counts_all)))
cat(sprintf("Host genes (Ha_): %d\n", nrow(counts_host)))
cat(sprintf("Pathogen genes (Ss_): %d\n", nrow(counts_pathogen)))
cat(sprintf("Samples: %d\n", ncol(counts_all)))
```

## Create Metadata

```{r create-metadata}
# Extract sample information from column names
# Sample name pattern: timepoint_replicate_species_Ssisolate
# Examples: "0_1_Ha_NC", "24_1_Ha_HA89", "48_2_Ha_RHA274", "96_3_Ha_XRQ"

sample_names <- colnames(counts_all)

metadata <- data.frame(
  sample = sample_names,
  row.names = sample_names,
  stringsAsFactors = FALSE
)

# Extract components from sample names
metadata$replicate <- sapply(strsplit(sample_names, "_"), `[`, 1)
metadata$batch <- sapply(strsplit(sample_names, "_"), `[`, 2)
metadata$species <- sapply(strsplit(sample_names, "_"), `[`, 3)
metadata$Ssisolate <- sapply(strsplit(sample_names, "_"), `[`, 4)

# Extract timepoint - FIXED VERSION
# For NC samples: last element is "NC"
# For time-course samples: first element is the timepoint number
metadata$timepoint <- sapply(sample_names, function(x) {
  parts <- strsplit(x, "_")[[1]]
  # Check if last part is "NC" (non-challenged)
  if(parts[length(parts)] == "NC") {
    return("NC")
  } else {
    # Otherwise use the first number with "h" suffix
    return(paste0(parts[1], "h"))
  }
})

# Convert to factors with correct levels
metadata$timepoint <- factor(metadata$timepoint, levels = c("NC", "24h", "48h", "96h"))
metadata$batch <- factor(metadata$batch)

print(head(metadata))
cat("\nSamples per timepoint:\n")
print(table(metadata$timepoint, useNA = "always"))
```

# Quality Control

## Library Sizes

```{r qc-library-sizes}
lib_sizes <- colSums(counts_host)

lib_df <- data.frame(
  sample = names(lib_sizes),
  size = lib_sizes / 1e6,
  timepoint = metadata$timepoint
)

ggplot(lib_df, aes(x = sample, y = size, fill = timepoint)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Sizes - Host Genes",
       x = "Sample", y = "Library size (millions)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))

cat(sprintf("\nLibrary sizes: %.2f - %.2f million reads\n", 
            min(lib_sizes)/1e6, max(lib_sizes)/1e6))
```

## Pathogen Load Progression

```{r pathogen-load}
pathogen_reads <- colSums(counts_pathogen)
host_reads <- colSums(counts_host)
pathogen_pct <- 100 * pathogen_reads / (pathogen_reads + host_reads)

pathogen_load_df <- data.frame(
  sample = colnames(counts_all),
  timepoint = metadata$timepoint,
  pathogen_percent = pathogen_pct
)

ggplot(pathogen_load_df, aes(x = timepoint, y = pathogen_percent)) +
  geom_boxplot(aes(fill = timepoint), outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pathogen Load Progression in Sunflower",
       x = "Timepoint", y = "Pathogen reads (%)") +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

# Summary statistics
pathogen_summary <- pathogen_load_df %>% 
  group_by(timepoint) %>% 
  summarise(
    mean = mean(pathogen_percent),
    sd = sd(pathogen_percent),
    min = min(pathogen_percent),
    max = max(pathogen_percent)
  )
print(pathogen_summary)
```

# DESeq2 Analysis - Host Genes

## Filter and Create DESeq2 Object

```{r deseq2-host-setup}
# IMPORTANT: Salmon produces non-integer counts, so we need to round them
# DESeq2 requires integer counts
counts_host <- round(counts_host)

# Filter low count genes (keep genes with >=10 reads in >=3 samples)
keep_host <- rowSums(counts_host >= 10) >= 3
counts_host_filtered <- counts_host[keep_host, ]

cat(sprintf("Host genes after filtering: %d (%.1f%% retained)\n",
            nrow(counts_host_filtered),
            100 * nrow(counts_host_filtered) / nrow(counts_host)))

# Create DESeq2 dataset
# NOTE: Using timepoint-only design to avoid potential batch nesting issues.
dds_host <- DESeqDataSetFromMatrix(
  countData = counts_host_filtered,
  colData = metadata,
  design = ~ timepoint
)

cat("\nRunning DESeq2 analysis for host genes...\n")
dds_host <- DESeq(dds_host)
cat("DESeq2 analysis complete!\n")
```

## PCA Plot

```{r pca-host}
vst_host <- vst(dds_host, blind = FALSE)

pcaData <- plotPCA(vst_host, intgroup = c("timepoint", "batch"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color = timepoint, shape = batch)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  labs(title = "PCA - Host Genes (Ha_)") +
  theme_bw(base_size = 14)
```

## Sample Distance Heatmap

```{r sample-distances-host}
sampleDists_host <- dist(t(assay(vst_host)))
sampleDistMatrix_host <- as.matrix(sampleDists_host)
rownames(sampleDistMatrix_host) <- paste(vst_host$timepoint, vst_host$batch, sep = "_")
colnames(sampleDistMatrix_host) <- NULL

pheatmap(sampleDistMatrix_host,
         clustering_distance_rows = sampleDists_host,
         clustering_distance_cols = sampleDists_host,
         col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
         main = "Sample Distance Matrix - Host Genes")
```

## Differential Expression Results

```{r de-host-comparisons}
# Define contrasts
contrasts_list <- list(
  "24h_vs_NC" = c("timepoint", "24h", "NC"),
  "48h_vs_NC" = c("timepoint", "48h", "NC"),
  "96h_vs_NC" = c("timepoint", "96h", "NC"),
  "48h_vs_24h" = c("timepoint", "48h", "24h"),
  "96h_vs_48h" = c("timepoint", "96h", "48h")
)

# Store results
results_host_list <- list()

# Get results for each comparison
for (name in names(contrasts_list)) {
  res <- results(dds_host, contrast = contrasts_list[[name]], alpha = 0.05)
  res <- res[order(res$padj), ]
  results_host_list[[name]] <- res
  
  # Print summary
  cat(sprintf("\n%s:\n", name))
  cat(sprintf("  Significant genes (padj < 0.05): %d\n", 
              sum(res$padj < 0.05, na.rm = TRUE)))
  cat(sprintf("  Upregulated: %d\n", 
              sum(res$padj < 0.05 & res$log2FoldChange > 0, na.rm = TRUE)))
  cat(sprintf("  Downregulated: %d\n", 
              sum(res$padj < 0.05 & res$log2FoldChange < 0, na.rm = TRUE)))
}
```

## Volcano Plots - Host Genes

```{r volcano-plots-host, fig.height=8}
for (name in names(results_host_list)) {
  res <- results_host_list[[name]]
  
  p <- EnhancedVolcano(res,
                       lab = rownames(res),
                       x = 'log2FoldChange',
                       y = 'padj',
                       title = paste('Host Genes:', name),
                       pCutoff = 0.05,
                       FCcutoff = 1.0,
                       pointSize = 2.0,
                       labSize = 3.0,
                       max.overlaps = 15)
  print(p)
}
```

## Top Significant Genes Heatmap

```{r heatmap-host-top}
# Get top 50 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:50]
mat <- assay(vst_host)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation
annotation_col <- data.frame(
  Timepoint = vst_host$timepoint,
  Ss_isolate = vst_host$Ssisolate,
  row.names = colnames(mat)
)
annotation_colors <- list(
  Timepoint = c("NC" = "#2E8B57",      # Sea green
                "24h" = "#FFD700",     # Gold
                "48h" = "#FF6347",     # Tomato
                "96h" = "#8B0000"),    # Dark red
  
  Ss_isolate = c("NC" = "#ffff47",         #yellow
            "WISS47" = "#1f77b4",          # Blue
            "JS659" = "#ff7f0e",          # Orange  
            "BN172" = "#2ca02c",          # Green
            "SSPotter" = "#8c564b",          # Brown
            "MNSS6" = "#9467bd",          # Purple
            "Xtra7" = "#d62728")          # Red
)

sample_order <- order(annotation_col$Ss_isolate, 
                     match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h")))

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]


pheatmap(mat,
         annotation_col = annotation_col,
         annotation_colors = annotation_colors,  
         show_rownames = TRUE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 50 Significant Genes",
         fontsize_row = 6)
```

```{r host_heatmap}
# Get top 100 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:100]
mat <- assay(vst_host)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_host$timepoint, levels = c("NC", "24h", "48h", "96h")),
  Ss_isolate = factor(vst_host$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors in the SAME ORDER as factor levels
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      
                "24h" = "#FFD700",     
                "48h" = "#FF6347",     
                "96h" = "#8B0000"),    
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order (excluding NC since filtered out)
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# Order samples by your custom isolate order, then by timepoint within each isolate
sample_order <- order(match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h"), 
                     match(annotation_col$Ss_isolate, isolate_order)))

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,  # Keep FALSE if you want to maintain your custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 100 Significant Genes on sunflower",
         fontsize_row = 6)

```

```{r host_heatmap_add}


# Get top 100 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:100]
mat <- assay(vst_host)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_host$timepoint, levels = c("NC", "24h", "48h", "96h")),
  Ss_isolate = factor(vst_host$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors in the SAME ORDER as factor levels
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      
                "24h" = "#FFD700",     
                "48h" = "#FF6347",     
                "96h" = "#8B0000"),    
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# Order samples by your custom isolate order, then by timepoint within each isolate
sample_order <- order(match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h"), 
                     match(annotation_col$Ss_isolate, isolate_order)))

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,  # Keep FALSE if you want to maintain your custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 100 Significant Genes on soybean",
         fontsize_row = 6)

```



# DESeq2 Analysis - Pathogen Genes

## Filter and Create DESeq2 Object

```{r deseq2-pathogen-setup}
# IMPORTANT: Round Salmon counts to integers
counts_pathogen <- round(counts_pathogen)

# Filter low count genes
keep_pathogen <- rowSums(counts_pathogen >= 10) >= 3
counts_pathogen_filtered <- counts_pathogen[keep_pathogen, ]

cat(sprintf("Pathogen genes after filtering: %d (%.1f%% retained)\n",
            nrow(counts_pathogen_filtered),
            100 * nrow(counts_pathogen_filtered) / nrow(counts_pathogen)))

# Create DESeq2 dataset
# NOTE: Using timepoint-only design to avoid potential batch nesting issues.
dds_pathogen <- DESeqDataSetFromMatrix(
  countData = counts_pathogen_filtered,
  colData = metadata,
  design = ~ timepoint
)

cat("\nRunning DESeq2 analysis for pathogen genes...\n")
dds_pathogen <- DESeq(dds_pathogen)
cat("DESeq2 analysis complete!\n")
```

## PCA Plot - Pathogen

```{r pca-pathogen}
vst_pathogen <- vst(dds_pathogen, blind = FALSE)

pcaData <- plotPCA(vst_pathogen, intgroup = c("timepoint", "batch"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color = timepoint, shape = batch)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  labs(title = "PCA - Pathogen Genes (Ss_)") +
  theme_bw(base_size = 14)
```

```{r pca-pathogen_2}
# Create VST
vst_pathogen <- vst(dds_pathogen, blind = FALSE)
# Filter out NC samples from the data
non_nc_samples <- vst_pathogen$timepoint != "NC"
vst_pathogen_filtered <- vst_pathogen[, non_nc_samples]

# Remove unused factor levels (like "NC") after subsetting
colData(vst_pathogen_filtered) <- droplevels(colData(vst_pathogen_filtered))

# Create PCA data with isolate and timepoint
pcaData <- plotPCA(vst_pathogen_filtered, intgroup = c("Ssisolate", "timepoint"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Convert isolate to factor with desired order
pcaData$isolate <- factor(pcaData$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter"))

# Define colors in the same order as factor levels
isolate_colors <- c("WISS47" = "#1f77b4",     # Blue
                   "JS659" = "#1f77b4",      # Blue (same as WISS47)
                   "MNSS6" = "#d62728",      # Red  
                   "Xtra7" = "#d62728",      # Red (same as MNSS6)
                   "BN172" = "#e682fa",      # Purple/Pink
                   "SSPotter" = "#8c564b")   # Brown

# Create the plot
ggplot(pcaData, aes(PC1, PC2, color = isolate, shape = timepoint)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = isolate_colors) +
  scale_shape_manual(values = c("24h" = 16,    # Circle
                               "48h" = 17,    # Triangle
                               "96h" = 15)) + # Square
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  labs(title = "PCA: Pathogen Gene Expression by Isolate",
       color = "Isolate",  # Legend title
       shape = "Timepoint") +
  theme_bw(base_size = 14)
```

## Differential Expression Results - Pathogen

```{r de-pathogen-comparisons}
# Store results
results_pathogen_list <- list()

# Get results for each comparison
for (name in names(contrasts_list)) {
  res <- results(dds_pathogen, contrast = contrasts_list[[name]], alpha = 0.05)
  res <- res[order(res$padj), ]
  results_pathogen_list[[name]] <- res
  
  # Print summary
  cat(sprintf("\n%s:\n", name))
  cat(sprintf("  Significant genes (padj < 0.05): %d\n", 
              sum(res$padj < 0.05, na.rm = TRUE)))
  cat(sprintf("  Upregulated: %d\n", 
              sum(res$padj < 0.05 & res$log2FoldChange > 0, na.rm = TRUE)))
  cat(sprintf("  Downregulated: %d\n", 
              sum(res$padj < 0.05 & res$log2FoldChange < 0, na.rm = TRUE)))
}
```

## Volcano Plots - Pathogen Genes

```{r volcano-plots-pathogen, fig.height=8}
for (name in names(results_pathogen_list)) {
  res <- results_pathogen_list[[name]]
  
  p <- EnhancedVolcano(res,
                       lab = rownames(res),
                       x = 'log2FoldChange',
                       y = 'padj',
                       title = paste('Pathogen Genes:', name),
                       pCutoff = 0.05,
                       FCcutoff = 1.0,
                       pointSize = 2.0,
                       labSize = 3.0,
                       max.overlaps = 15)
  print(p)
}
```

```{r pathogen_heatmap}
# Get top 50 genes from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[1:50]
mat <- assay(vst_pathogen)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation
annotation_col <- data.frame(
  Timepoint = vst_pathogen$timepoint,
  Ss_isolate = vst_pathogen$Ssisolate,
  row.names = colnames(mat)
)
annotation_colors <- list(
  Timepoint = c("NC" = "#2E8B57",      # Sea green
                "24h" = "#FFD700",     # Gold
                "48h" = "#FF6347",     # Tomato
                "96h" = "#8B0000"),    # Dark red
  
  Ss_isolate = c("NC" = "#ffff47",         #yellow
            "WISS47" = "#1f77b4",          # Blue
            "JS659" = "#ff7f0e",          # Orange  
            "BN172" = "#2ca02c",          # Green
            "SSPotter" = "#8c564b",          # Brown
            "MNSS6" = "#9467bd",          # Purple
            "Xtra7" = "#d62728")          # Red
)

sample_order <- order(annotation_col$Ss_isolate, 
                     match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h")))

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]


pheatmap(mat,
         annotation_col = annotation_col,
         annotation_colors = annotation_colors,  
         show_rownames = TRUE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen: Top 50 Significant Genes on sunflower",
         fontsize_row = 6)

```

```{r host_heatmap_add1}

# Get top 100 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:100]
mat <- assay(vst_host)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_host$timepoint, levels = c("NC", "24h", "48h", "96h")),
  Ss_isolate = factor(vst_host$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors in the SAME ORDER as factor levels
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      
                "24h" = "#FFD700",     
                "48h" = "#FF6347",     
                "96h" = "#8B0000"),    
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,  # Keep FALSE if you want to maintain your custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 100 Significant Genes on Sunflower",
         fontsize_row = 6)


```



```{r pathogen_heatmap_noNC_add1}
# Filter out NC samples from the data
non_nc_samples <- vst_pathogen$timepoint != "NC"
vst_pathogen_filtered <- vst_pathogen[, non_nc_samples]

# Remove unused factor levels (like "NC") after subsetting
colData(vst_pathogen_filtered) <- droplevels(colData(vst_pathogen_filtered))

# Get top 100 genes from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[1:100]
mat <- assay(vst_pathogen_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_pathogen_filtered$timepoint, levels = c("24h", "48h", "96h")),
  Ss_isolate = factor(vst_pathogen_filtered$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Updated annotation colors (removed NC from both)
annotation_colors <- list(
  Timepoint = c("24h" = "#FFD700",     # Gold
                "48h" = "#FF6347",     # Tomato
                "96h" = "#8B0000"),    # Dark red
  
  Ss_isolate = c(
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen: Top 100 Significant Genes on Sunflower",
         fontsize_row = 6)


```


```{r host_heatmap_add2}

# Get top 100 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:100]
mat <- assay(vst_host)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_host$timepoint, levels = c("NC", "24h", "48h", "96h")),
  Ss_isolate = factor(vst_host$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors in the SAME ORDER as factor levels
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      
                "24h" = "#FFD700",     
                "48h" = "#FF6347",     
                "96h" = "#8B0000"),    
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE if you want to maintain your custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 100 Significant Genes on Sunflower",
         fontsize_row = 6)


```



```{r pathogen_heatmap_noNC_add2}
# Filter out NC samples from the data
non_nc_samples <- vst_pathogen$timepoint != "NC"
vst_pathogen_filtered <- vst_pathogen[, non_nc_samples]

# Remove unused factor levels (like "NC") after subsetting
colData(vst_pathogen_filtered) <- droplevels(colData(vst_pathogen_filtered))

# Get top 100 genes from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[1:100]
mat <- assay(vst_pathogen_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_pathogen_filtered$timepoint, levels = c("24h", "48h", "96h")),
  Ss_isolate = factor(vst_pathogen_filtered$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Updated annotation colors (removed NC from both)
annotation_colors <- list(
  Timepoint = c("24h" = "#FFD700",     # Gold
                "48h" = "#FF6347",     # Tomato
                "96h" = "#8B0000"),    # Dark red
  
  Ss_isolate = c(
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen: Top 100 Significant Genes on Sunflower",
         fontsize_row = 6)


```

```{r host_heatmap_24hpi}
# Filter to keep only NC and 24hr timepoints
samples_nc_24hr <- vst_host$timepoint %in% c("NC", "24h")
vst_host_filtered <- vst_host[, samples_nc_24hr]

# Get top 100 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:100]
mat <- assay(vst_host_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with only NC and 24hr timepoints
annotation_col <- data.frame(
  Timepoint = factor(vst_host_filtered$timepoint, levels = c("NC", "24h")),
  Ss_isolate = factor(vst_host_filtered$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors for only NC and 24hr
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      # Light cyan
                "24h" = "#FFD700"),    # Gold
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# Order samples by isolate first, then by timepoint within each isolate (NC, then 24h)
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("NC", "24h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples - comparing NC vs 24hr
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE to maintain custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 100 Significant Genes - NC vs 24hr Comparison",
         fontsize_row = 6,
         fontsize_col = 8)





```

```{r host_heatmap_48hpi}
# Filter to keep only NC and 48hr timepoints
samples_nc_48hr <- vst_host$timepoint %in% c("NC", "48h")
vst_host_filtered <- vst_host[, samples_nc_48hr]

# Get top 100 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:100]
mat <- assay(vst_host_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with only NC and 48hr timepoints
annotation_col <- data.frame(
  Timepoint = factor(vst_host_filtered$timepoint, levels = c("NC", "48h")),
  Ss_isolate = factor(vst_host_filtered$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors for only NC and 48hr
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      # Light cyan
                "48h" = "#FF6347"),    # Gold
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# Order samples by isolate first, then by timepoint within each isolate (NC, then 48h)
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("NC", "48h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples - comparing NC vs 48hr
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE to maintain custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 100 Significant Genes - NC vs 48hr Comparison",
         fontsize_row = 6,
         fontsize_col = 8)





```


```{r host_heatmap_96hpi}
# Filter to keep only NC and 24hr timepoints
samples_nc_96hr <- vst_host$timepoint %in% c("NC", "96h")
vst_host_filtered <- vst_host[, samples_nc_96hr]



# Get top 100 genes from first comparison
top_genes <- rownames(results_host_list[[1]])[1:100]
mat <- assay(vst_host_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with only NC and 96hr timepoints
annotation_col <- data.frame(
  Timepoint = factor(vst_host_filtered$timepoint, levels = c("NC", "96h")),
  Ss_isolate = factor(vst_host_filtered$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors for only NC and 96hr
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      # Light cyan
                "96h" = "#8B0000"),    # Gold
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# Order samples by isolate first, then by timepoint within each isolate (NC, then 96h)
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("NC", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples - comparing NC vs 96hr
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE to maintain custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host: Top 100 Significant Genes - NC vs 96hr Comparison",
         fontsize_row = 6,
         fontsize_col = 8)





```


```{r pathogen_heatmap_24hpi}
# Filter out NC samples AND keep only 24hr timepoint
samples_24hr <- vst_pathogen$timepoint == "24h"
vst_pathogen_24hr <- vst_pathogen[, samples_24hr]

# Get top 100 genes from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[1:100]
mat_24hr <- assay(vst_pathogen_24hr)[top_genes, ]
mat_24hr <- t(scale(t(mat_24hr)))  # Z-score normalization

# Annotation with only 24hr samples
annotation_col_24hr <- data.frame(
  Ss_isolate = factor(vst_pathogen_24hr$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat_24hr)
)

# Colors for isolates only (no timepoint needed since all are 24hr)
annotation_colors_24hr <- list(
  Ss_isolate = c(
    "WISS47" = "#1f77b4",     # Blue
    "JS659" = "#1f77b4",      # Blue (same as WISS47)
    "MNSS6" = "#d62728",      # Red  
    "Xtra7" = "#d62728",      # Red (same as MNSS6)
    "BN172" = "#e682fa",      # Purple/Pink
    "SSPotter" = "#8c564b")   # Brown
)

# Order samples by isolate for better comparison
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")
sample_order_24hr <- order(match(annotation_col_24hr$Ss_isolate, isolate_order))

mat_24hr_ordered <- mat_24hr[, sample_order_24hr]
annotation_col_24hr_ordered <- annotation_col_24hr[sample_order_24hr, , drop = FALSE]

# Create heatmap for 24hr timepoint only
pheatmap(mat_24hr_ordered,
         annotation_col = annotation_col_24hr_ordered,
         annotation_colors = annotation_colors_24hr,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE to maintain isolate grouping
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen: Top 100 Significant Genes at 24hr Timepoint",
         fontsize_row = 6,
         fontsize_col = 8)  # Might want to increase column font size since fewer samples


```

```{r pathogen_heatmap_48hpi}
# Filter out NC samples AND keep only 48hr timepoint
samples_48hr <- vst_pathogen$timepoint == "48h"
vst_pathogen_48hr <- vst_pathogen[, samples_48hr]

# Get top 100 genes from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[1:100]
mat_48hr <- assay(vst_pathogen_48hr)[top_genes, ]
mat_48hr <- t(scale(t(mat_48hr)))  # Z-score normalization

# Annotation with only 48hr samples
annotation_col_48hr <- data.frame(
  Ss_isolate = factor(vst_pathogen_48hr$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat_48hr)
)

# Colors for isolates only (no timepoint needed since all are 24hr)
annotation_colors_48hr <- list(
  Ss_isolate = c(
    "WISS47" = "#1f77b4",     # Blue
    "JS659" = "#1f77b4",      # Blue (same as WISS47)
    "MNSS6" = "#d62728",      # Red  
    "Xtra7" = "#d62728",      # Red (same as MNSS6)
    "BN172" = "#e682fa",      # Purple/Pink
    "SSPotter" = "#8c564b")   # Brown
)

# Order samples by isolate for better comparison
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")
sample_order_48hr <- order(match(annotation_col_48hr$Ss_isolate, isolate_order))

mat_48hr_ordered <- mat_48hr[, sample_order_48hr]
annotation_col_48hr_ordered <- annotation_col_48hr[sample_order_48hr, , drop = FALSE]

# Create heatmap for 24hr timepoint only
pheatmap(mat_48hr_ordered,
         annotation_col = annotation_col_48hr_ordered,
         annotation_colors = annotation_colors_48hr,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE to maintain isolate grouping
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen: Top 100 Significant Genes at 48hr Timepoint",
         fontsize_row = 6,
         fontsize_col = 8)  # Might want to increase column font size since fewer samples


```

```{r pathogen_heatmap_96hpi}

# Filter out NC samples from the data
non_nc_samples <- vst_pathogen$timepoint != "NC"
vst_pathogen_filtered <- vst_pathogen[, non_nc_samples]

# Remove unused factor levels (like "NC") after subsetting
colData(vst_pathogen_filtered) <- droplevels(colData(vst_pathogen_filtered))

# Filter out NC samples AND keep only 96hr timepoint
samples_96hr <- vst_pathogen$timepoint == "96h"
vst_pathogen_96hr <- vst_pathogen[, samples_96hr]

# Remove unused factor levels for 96hr subset too
colData(vst_pathogen_96hr) <- droplevels(colData(vst_pathogen_96hr))

# Get top 100 genes from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[1:100]
mat_96hr <- assay(vst_pathogen_96hr)[top_genes, ]
mat_96hr <- t(scale(t(mat_96hr)))  # Z-score normalization

# Annotation with only 96hr samples
annotation_col_96hr <- data.frame(
  Ss_isolate = factor(vst_pathogen_96hr$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat_96hr)
)

# Colors for isolates only (no timepoint needed since all are 96hr)
annotation_colors_96hr <- list(
  Ss_isolate = c(
    "WISS47" = "#1f77b4",     # Blue
    "JS659" = "#1f77b4",      # Blue (same as WISS47)
    "MNSS6" = "#d62728",      # Red  
    "Xtra7" = "#d62728",      # Red (same as MNSS6)
    "BN172" = "#e682fa",      # Purple/Pink
    "SSPotter" = "#8c564b")   # Brown
)

# Order samples by isolate for better comparison
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")
sample_order_96hr <- order(match(annotation_col_96hr$Ss_isolate, isolate_order))

mat_96hr_ordered <- mat_96hr[, sample_order_96hr]
annotation_col_96hr_ordered <- annotation_col_96hr[sample_order_96hr, , drop = FALSE]

# Create heatmap for 96hr timepoint only
pheatmap(mat_96hr_ordered,
         annotation_col = annotation_col_96hr_ordered,
         annotation_colors = annotation_colors_96hr,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE to maintain isolate grouping
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen: Top 100 Significant Genes at 96hr Timepoint",
         fontsize_row = 6,
         fontsize_col = 8)  # Might want to increase column font size since fewer samples


```


```{r PCA_pathogen_re_2}

# Create VST
vst_pathogen <- vst(dds_pathogen, blind = FALSE)

# Filter out NC samples from the data
non_nc_samples <- vst_pathogen$timepoint != "NC"
vst_pathogen_filtered <- vst_pathogen[, non_nc_samples]

# Create PCA data with isolate and timepoint
pcaData <- plotPCA(vst_pathogen_filtered, intgroup = c("Ssisolate", "timepoint"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Convert isolate to factor with desired order
pcaData$isolate <- factor(pcaData$Ssisolate, 
                         levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter"))

# Create aggressiveness grouping based on your classification
pcaData$aggressiveness <- case_when(
  pcaData$Ssisolate %in% c("WISS47", "JS659") ~ "Low Aggressive",
  pcaData$Ssisolate %in% c("MNSS6", "Xtra7") ~ "High Aggressive", 
  pcaData$Ssisolate == "SSPotter" ~ "Soybean Specific",
  pcaData$Ssisolate == "BN172" ~ "Sunflower Specific"
)

# Convert to factor with desired order
pcaData$aggressiveness <- factor(pcaData$aggressiveness, 
                                levels = c("Low Aggressive", "High Aggressive", 
                                          "Soybean Specific", "Sunflower Specific"))

# Define colors for isolates (consistent across both plots)
isolate_colors <- c(
  "WISS47" = "#4169E1",    # Blue
  "JS659" = "#1f77b4",     # Steel Blue  
  "MNSS6" = "#DC143C",     # Red
  "Xtra7" = "#FF6B6B",     # Coral Red
  "BN172" = "#DA70D6",     # Purple
  "SSPotter" = "#8B4513"   # Brown
)

# Define colors for timepoints
timepoint_colors <- c("24h" = "#FFD700", "48h" = "#FF6347", "96h" = "#8B0000")

# Define colors for aggressiveness groups
aggressiveness_colors <- c(
  "Low Aggressive" = "#4169E1",      # Blue
  "High Aggressive" = "#DC143C",     # Red  
  "Soybean Specific" = "#8B4513",    # Brown
  "Sunflower Specific" = "#DA70D6"   # Purple
)

# PLOT 1: Grouped by Timepoint
p1 <- ggplot(pcaData, aes(PC1, PC2, color = isolate, shape = timepoint)) +
  
  # Add ellipses for timepoint grouping
  stat_ellipse(aes(fill = timepoint, color = timepoint), 
               alpha = 0.2, 
               type = "norm", 
               level = 0.68,
               size = 1.2) +
  
  # Data points
  geom_point(aes(color = isolate), size = 4, alpha = 0.8, stroke = 1.2) +
  
  # Color and shape scales
  scale_color_manual(values = isolate_colors) +
  scale_fill_manual(values = timepoint_colors) +
  scale_shape_manual(values = c("24h" = 16, "48h" = 17, "96h" = 15)) +
  
  # Labels
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  labs(title = "PCA: Pathogen Gene Expression Grouped by Timepoint") +
  
  # Theme
  theme_void() +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    axis.text = element_text(size = 12, face = "bold", color = "black"),
    axis.title = element_text(size = 12, face = "bold", color = "black"),
    axis.line = element_line(color = "black", size = 0.5),
    axis.ticks = element_line(color = "black", size = 0.5),
    plot.title = element_text(size = 14, face = "bold", color = "black", hjust = 0.5),
    legend.position = "bottom",
    legend.background = element_rect(fill = "white", alpha = 0.9, color = "grey80"),
    legend.title = element_text(size = 11, face = "bold", color = "black"),
    legend.text = element_text(size = 10, face = "bold", color = "black"),
    legend.box = "horizontal",
    legend.box.spacing = unit(1, "cm"),
    panel.grid.major = element_line(color = "grey90", alpha = 0.5, size = 0.5),
    panel.border = element_rect(color = "black", fill = NA, size = 0.8),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) +
  
  # Guides
  guides(
    color = guide_legend(title = "Isolate", override.aes = list(size = 4), nrow = 1),
    fill = guide_legend(title = "Timepoint Group", override.aes = list(alpha = 0.4), nrow = 1),
    shape = guide_legend(title = "Timepoint", override.aes = list(size = 4), nrow = 1)
  )

# PLOT 2: Grouped by Aggressiveness
p2 <- ggplot(pcaData, aes(PC1, PC2, color = isolate, shape = timepoint)) +
  
  # Add ellipses for aggressiveness grouping
  stat_ellipse(aes(fill = aggressiveness, color = aggressiveness), 
               alpha = 0.2, 
               type = "norm", 
               level = 0.68,
               size = 1.2) +
  
  # Data points
  geom_point(aes(color = isolate), size = 4, alpha = 0.8, stroke = 1.2) +
  
  # Color and shape scales
  scale_color_manual(values = isolate_colors) +
  scale_fill_manual(values = aggressiveness_colors) +
  scale_shape_manual(values = c("24h" = 16, "48h" = 17, "96h" = 15)) +
  
  # Labels
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  labs(title = "PCA: Pathogen Gene Expression Grouped by Aggressiveness") +
  
  # Theme
  theme_void() +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    axis.text = element_text(size = 12, face = "bold", color = "black"),
    axis.title = element_text(size = 12, face = "bold", color = "black"),
    axis.line = element_line(color = "black", size = 0.5),
    axis.ticks = element_line(color = "black", size = 0.5),
    plot.title = element_text(size = 14, face = "bold", color = "black", hjust = 0.5),
    legend.position = "bottom",
    legend.background = element_rect(fill = "white", alpha = 0.9, color = "grey80"),
    legend.title = element_text(size = 11, face = "bold", color = "black"),
    legend.text = element_text(size = 10, face = "bold", color = "black"),
    legend.box = "horizontal",
    legend.box.spacing = unit(1, "cm"),
    panel.grid.major = element_line(color = "grey90", alpha = 0.5, size = 0.5),
    panel.border = element_rect(color = "black", fill = NA, size = 0.8),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) +
  
  # Guides
  guides(
    color = guide_legend(title = "Isolate", override.aes = list(size = 4), nrow = 1),
    fill = guide_legend(title = "Aggressiveness Group", override.aes = list(alpha = 0.4), nrow = 1),
    shape = guide_legend(title = "Timepoint", override.aes = list(size = 4), nrow = 1)
  )

# Print both plots
print(p1)
print(p2)

# Optional: Save both plots
# ggsave("pca_timepoint_grouped.png", plot = p1, 
#        width = 12, height = 8, dpi = 300, bg = "transparent")
# ggsave("pca_aggressiveness_grouped.png", plot = p2, 
#        width = 12, height = 8, dpi = 300, bg = "transparent")

# Optional: Create combined plot using patchwork
# library(patchwork)
# combined_plot <- p1 / p2
# print(combined_plot)

```

# Summary

```{r summary}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("SUNFLOWER-SCLEROTINIA DUAL RNA-SEQ ANALYSIS SUMMARY\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("INPUT DATA:\n")
cat(sprintf("Total genes: %d\n", nrow(counts_all)))
cat(sprintf("Host genes (Ha_): %d\n", nrow(counts_host)))
cat(sprintf("Pathogen genes (Ss_): %d\n", nrow(counts_pathogen)))
cat(sprintf("Samples analyzed: %d\n", ncol(counts_all)))

cat("\nHOST GENE DIFFERENTIAL EXPRESSION:\n")
for (name in names(results_host_list)) {
  res <- results_host_list[[name]]
  cat(sprintf("\n%s: %d significant genes\n", name, 
              sum(res$padj < 0.05, na.rm = TRUE)))
}

cat("\nPATHOGEN GENE DIFFERENTIAL EXPRESSION:\n")
for (name in names(results_pathogen_list)) {
  res <- results_pathogen_list[[name]]
  cat(sprintf("\n%s: %d significant genes\n", name, 
              sum(res$padj < 0.05, na.rm = TRUE)))
}
```



# Export Results

```{r export-results, eval=FALSE}
# Create output directory
dir.create("sunflower_results", showWarnings = FALSE)

# Save normalized counts
write.csv(counts(dds_host, normalized = TRUE),
          "sunflower_results/normalized_counts_HOST.csv")
write.csv(counts(dds_pathogen, normalized = TRUE),
          "sunflower_results/normalized_counts_PATHOGEN.csv")

# Save DE results
for (name in names(results_host_list)) {
  write.csv(as.data.frame(results_host_list[[name]]),
            paste0("sunflower_results/HOST_", name, "_all.csv"))
}

for (name in names(results_pathogen_list)) {
  write.csv(as.data.frame(results_pathogen_list[[name]]),
            paste0("sunflower_results/PATHOGEN_", name, "_all.csv"))
}

cat("Results exported to sunflower_results/\n")
```


```{r all host sig DEGs}
# Filter to only genes that exist in the vst_host assay
top_genes <- rownames(results_host_list[[1]])[
  which(results_host_list[[1]]$padj < 0.05 & 
        abs(results_host_list[[1]]$log2FoldChange) > 1)
]

# Keep only genes present in the assay
top_genes <- intersect(top_genes, rownames(assay(vst_host_filtered)))

mat <- assay(vst_host)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_host$timepoint, levels = c("NC", "24h", "48h", "96h")),
  Ss_isolate = factor(vst_host$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors in the SAME ORDER as factor levels
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      
                "24h" = "#FFD700",     
                "48h" = "#FF6347",     
                "96h" = "#8B0000"),    
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,  # Keep FALSE if you want to maintain your custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host Significant Genes on Sunflower",
         fontsize_row = 6)





```

```{r all pathogen sig DEGs}
# Filter out NC samples from the data
non_nc_samples <- vst_pathogen$timepoint != "NC"
vst_pathogen_filtered <- vst_pathogen[, non_nc_samples]

# Remove specific samples by name (removed duplicates)
samples_to_remove <- c("96_1_Gm_BN172","96_2_Gm_BN172","96_3_Gm_BN172", "96_1_Gm_SSPotter", "96_2_Gm_SSPotter", "96_3_Gm_SSPotter")
samples_to_keep <- !colnames(vst_pathogen_filtered) %in% samples_to_remove
vst_pathogen_filtered <- vst_pathogen_filtered[, samples_to_keep]

# Get ALL significant DEGs from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[
  which(results_pathogen_list[[1]]$padj < 0.05 & 
        abs(results_pathogen_list[[1]]$log2FoldChange) > 1)
]

# Keep only genes present in the assay
top_genes <- intersect(top_genes, rownames(assay(vst_pathogen_filtered)))

mat <- assay(vst_pathogen_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization
mat <- assay(vst_pathogen_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_pathogen_filtered$timepoint, levels = c("24h", "48h", "96h")),
  Ss_isolate = factor(vst_pathogen_filtered$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Updated annotation colors (removed NC from both)
annotation_colors <- list(
  Timepoint = c("24h" = "#FFD700",     # Gold
                "48h" = "#FF6347",     # Tomato
                "96h" = "#8B0000"),    # Dark red
  
  Ss_isolate = c(
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen Significant Genes on Sunflower",
         fontsize_row = 6)

```


```{r all host sig DEGs_no_col_cluster}
# Filter to only genes that exist in the vst_host assay
top_genes <- rownames(results_host_list[[1]])[
  which(results_host_list[[1]]$padj < 0.05 & 
        abs(results_host_list[[1]]$log2FoldChange) > 1)
]

# Keep only genes present in the assay
top_genes <- intersect(top_genes, rownames(assay(vst_host_filtered)))

mat <- assay(vst_host)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_host$timepoint, levels = c("NC", "24h", "48h", "96h")),
  Ss_isolate = factor(vst_host$Ssisolate, levels = c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Annotation colors in the SAME ORDER as factor levels
annotation_colors <- list(
  Timepoint = c("NC" = "#91fffb",      
                "24h" = "#FFD700",     
                "48h" = "#FF6347",     
                "96h" = "#8B0000"),    
  
  Ss_isolate = c("NC" = "#ffffad",         # Light yellow
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("NC", "WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("NC", "24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,  
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,  # Keep FALSE if you want to maintain your custom order
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Host Significant Genes on Sunflower",
         fontsize_row = 6)





```

```{r all pathogen sig DEGs_no_col_cluster}
# Filter out NC samples from the data
non_nc_samples <- vst_pathogen$timepoint != "NC"
vst_pathogen_filtered <- vst_pathogen[, non_nc_samples]

# Remove specific samples by name (removed duplicates)
samples_to_remove <- c("96_1_Gm_BN172","96_2_Gm_BN172","96_3_Gm_BN172", "96_1_Gm_SSPotter", "96_2_Gm_SSPotter", "96_3_Gm_SSPotter")
samples_to_keep <- !colnames(vst_pathogen_filtered) %in% samples_to_remove
vst_pathogen_filtered <- vst_pathogen_filtered[, samples_to_keep]

# Get ALL significant DEGs from first comparison
top_genes <- rownames(results_pathogen_list[[1]])[
  which(results_pathogen_list[[1]]$padj < 0.05 & 
        abs(results_pathogen_list[[1]]$log2FoldChange) > 1)
]

# Keep only genes present in the assay
top_genes <- intersect(top_genes, rownames(assay(vst_pathogen_filtered)))

mat <- assay(vst_pathogen_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization
mat <- assay(vst_pathogen_filtered)[top_genes, ]
mat <- t(scale(t(mat)))  # Z-score normalization

# Annotation with factors in desired order
annotation_col <- data.frame(
  Timepoint = factor(vst_pathogen_filtered$timepoint, levels = c("24h", "48h", "96h")),
  Ss_isolate = factor(vst_pathogen_filtered$Ssisolate, levels = c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")),
  row.names = colnames(mat)
)

# Updated annotation colors (removed NC from both)
annotation_colors <- list(
  Timepoint = c("24h" = "#FFD700",     # Gold
                "48h" = "#FF6347",     # Tomato
                "96h" = "#8B0000"),    # Dark red
  
  Ss_isolate = c(
                 "WISS47" = "#1f77b4",     # Blue
                 "JS659" = "#1f77b4",      # Blue (same as WISS47)
                 "MNSS6" = "#d62728",      # Red  
                 "Xtra7" = "#d62728",      # Red (same as MNSS6)
                 "BN172" = "#e682fa",      # Purple/Pink
                 "SSPotter" = "#8c564b")   # Brown
)

# Define your preferred isolate order
isolate_order <- c("WISS47", "JS659", "MNSS6", "Xtra7", "BN172", "SSPotter")

# CORRECTED: Order samples by isolate first, then by timepoint within each isolate
sample_order <- order(
  match(annotation_col$Ss_isolate, isolate_order),
  match(annotation_col$Timepoint, c("24h", "48h", "96h"))
)

mat_ordered <- mat[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, , drop = FALSE]

# Create heatmap with ordered samples
pheatmap(mat_ordered,
         annotation_col = annotation_col_ordered,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         main = "Pathogen Significant Genes on Sunflower",
         fontsize_row = 6)

```



# Session Info

```{r session-info}
sessionInfo()
```
